### 2022-02-22 main_text.tex
initial commit

\highlight{placeholder abstract - only for general idea not to be proofread}
\begin{figure}
    \includegraphics[width=\columnwidth]{picto visual abstract.pdf}
\end{figure}
Estuaries are a challenging for plankton.
Gradients are steep and currents are strong.
Retention mechanisms crucial process for survival.
However, very little experimental and theoretical evidence has been gathered.
We will conduct a model study.
First we will show the importance of retention mechanism for plankton in the Elbe.
Secondly we will study different retention mechanism which are suspected to be applied by organism.
We compare the outcome of the different experiment and judge their effectiveness.

\section*{Introduction}

\subsection*{Why do we care about retention in estuaries?}
\begin{itemize}
    \item Estuaries are important
    \begin{itemize}
        \item high productivity relative to surface area
        \item nutrient source for marine ecosystem
        \item hatching grounds for not only for estuarine but marine species
    \end{itemize}
    %and productive ecosystem - both as fertilizer for marine ecosystems as well as for the global climate system.
    \item Ecosystem strongly "controlled" by primary producer.
    \begin{itemize}
        \item (Do I need to argue/cite something for this?)
    \end{itemize}
    \item Primary producers almost exclusively planktonic. [cite]
          % Suspected reasons: trenching, high velocities on lose soil, high turbidity
    \item strong hydrodynamic currents %and strong chemical gradiants.
    \item "retainment" crucial for survival 
    %for organisms as they would die if washed out to downstream/higher salinity areas.
    \item "swimming" upstream is impossible or too energy expensive for planktonic organisms
    \item However, retainment mechanisms are mostly unknown
    \item Q: How do they survive?
    % as they are hard to study experimentally.
\end{itemize}

\subsection*{Previous studies - aka - what is known}

There are three kind of retention studies: observations, theoretical studies, or modelling studies

\paragraph{observations}
\begin{itemize}
    \item only larger organisms which are not strictly planktonic (fish-larve, mesozooplanktion) [cite]
    \begin{itemize}
        \item found diel migration  [cite]
        \item found tide driven migration
        \item found "sticking" to "ground" [cite]
        \item 
    \end{itemize}
    \item summary: retention indications found for "planktonic" - tho not directly transferable to phytoplankton
\end{itemize}

\paragraph{theoretical}
\begin{itemize}
    \item interesting but study simplistic conditions 
    \item two interesting ones:
    \begin{itemize}
        \item investigation of reproduction threshold under exchange cite
        \item study on "reseeding" due to local upstream currents
    \end{itemize}
    summury: both "outgrowing" and "reseeding" are expected to be possible
\end{itemize}

\paragraph{(particle tracking) models}
\begin{itemize}
    \item only one particle tracking model in estuaries which studies zooplankton
    \item others studies open water or non planktonic organisms i.e. fish
    \item the existing one has mayor drawbacks - e.g. no reproduction
    \item found that sinking and diel migration slows outwashing
    \item non were able to actually study retention in realistic conditions due to technical constraints
\end{itemize}

\paragraph{Summary}
\begin{itemize}
    \item "surrounding" studies have been made
    \item none were able to tackle retention mechanisms themselves
    \item indication for potential mechanisms have been found
    \item we will investigate retention and reseeding mechanism under different plankton migration behavior
\end{itemize}

\section*{Methods}

\subsection*{general model description}
\paragraph{working principle}
\begin{itemize}
    \item particle tracking
    \begin{itemize}
        \item fast
        \item easy to interpret
        \item flexible e.g. allows "individual" processes
        \item realistic representation of motion
    \end{itemize}
\end{itemize}
\begin{figure}
    \includegraphics[width=\columnwidth]{Artboard 2.png}
    \caption{Particle Trackle visualisation - particle tracks on grid}
\end{figure}

\paragraph{data}
\begin{itemize}
    \item input is SCHISM output
    \item unstructured 3d grid of full estuary (geesthacht to helgoland with side-channels)
    \item (used) fields:
    \begin{itemize}
        \item temperature
        \item salinity
        \item water velocity 
    \end{itemize}
    \item full 2012 with 1h temporal- and ~5m spacial resolution
    \item year in reference to others
\end{itemize}
\begin{figure}
    \includegraphics[width=\columnwidth]{Artboard 1.png}
    \caption{model domain visualization - full range grid including statistical polygons for later reference and example of a scalar field (e.g. temperature)}
    \label{fig:statistial polygons}
\end{figure}

\paragraph{bio-logical mechanisms represented in the model}
\begin{itemize}
    \item release
    \item splitting
    \item culling
    \item behavior
    \begin{itemize}
        \item sinking
        \item diel migration
        \item tidal migration
    \end{itemize}
    \item sedimentation/resuspension
\end{itemize}

\begin{figure}
    \centering
    \begin{minipage}{.4\columnwidth}
        \centering
        \includegraphics[width=.3\columnwidth]{picto resuspension.pdf}
    \end{minipage}
    \begin{minipage}{.4\columnwidth}
    \centering
        \includegraphics[width=.3\columnwidth]{picto splitting.pdf}
    \end{minipage}
    \caption{examples of pictograms for bio processes}
\end{figure}

\paragraph{Data Analysis}
\begin{itemize}
    \item Based on particle counts in statistical polygons \ref{fig:statistial polygons}
    \item statistical polygons roughly 100km in length
    \item calculation of particle "centroid" and whether particles are in a section of the estuary.
    \item successful retainment is judged by whether or not particles remain alive somewhere in the estuary after one year.
\end{itemize}
\begin{figure}
    \includegraphics[width=\columnwidth]{fixed_particle_culling.pdf}
    \caption{Example 2d plot of particle counts, centroid and "retainment" threshold}
\end{figure}

\subsection*{Experimental configurations}

Two major setups with several minor configuration setups in relation to reference configurations
\begin{figure}
    \includegraphics[width=\columnwidth]{plot experimental configuration.pdf}
    \caption{2d grid with boxes for the different cases/dimension. 
             Same structure as the later sensitivity analysis plot used \ref{fig:retention_results_SA}}
    \label{fig:experimental_configuration}
\end{figure}

\paragraph{reference run - dead\&passive}
\begin{itemize}
    \item no reproduction 
    \item no behavior just drifting
\end{itemize}

\paragraph{downstream retention}
\begin{itemize}
    \item studies retention mechanisms under different behavior/growth
    \item sensitivity analysis with the following dimensions
    \begin{itemize}
        \item growth rate 
        \item sinking
        \item diel migration
        \item tidal migration
    \end{itemize} 
    \item release location description
\end{itemize}

\paragraph{upstream migration}
\paragraph{downstream retention}

(Note: This is the "release downstream and check if we find some upstream" experiment)
\begin{itemize}
    \item studies reseeding mechanisms of upstream waters under different behavior/growth
    \item same SA dimensions as above
    \item release location description
\end{itemize}

\section*{Results}

\subsection*{downstream retention}

\begin{figure}
    \includegraphics[width=\columnwidth]{plot results sa.pdf}
    \caption{2d grid with boxes for the different cases/dimension. 
             each box represents one configuration and whether they were able to retain by color code.
             I.e. a bunch of red and green boxes with axis labels}
    \label{fig:retention_results_SA}
\end{figure}

\begin{figure}
    \includegraphics[width=\columnwidth]{plot result threshold change.pdf}
    \caption{bar plot of the $\frac{\partial \mu_{min}}{\partial v_{verticle}}$ relations}
    \label{fig:retention_results_dmu}
\end{figure}

\begin{itemize}
    \item discuss which configuration maintained a population and which not
    \item describe growth threshold under different behavior scenarios
    \item describe $\frac{\partial \mu_{min}}{\partial v_{verticle}}$ for sinking, diel-, tidal migration
\end{itemize}

\paragraph{reproduction dimension}
\begin{itemize}
    \item discuss which configuration managed to colonize upstream areas
    \item discuss threshold growth rates and the impact of behavior patters similar to above
\end{itemize}

\paragraph{Discuss Results}
contextualize the results 
\begin{itemize}
    \item doubling times of reproduction of threshold reproduction rates (for context)
    \item threshold reproduction rates in relation to "realistic" reproduction rates summer/winter - aka are these mechanisms sufficient to allow survival
    \item discuss effectiveness of migration behavior - most effective mechanisms should a priori be realized in nature 
    \item (once all results are done - check if we can make predictions that are experimentally testable. e.g. average sinking velocities)
    \item reseeding (with the preliminary results) is not possible - no upstream migration above the harbor aka non-tidal-dominated-areas
\end{itemize}

\section*{Discussion}
\begin{itemize}
    \item discuss limitation of the model
    \begin{itemize}
        \item simplified bottom/sediment interaction
        \item no "small object" representation e.g. vegetation or "rocks"
        \item simplistic bio representation. does not allow for precise quantification but only indicates bounds for values
        \item other minor issues - temporal resolution, particle out-of-bounds cases etc
    \end{itemize}
    \item discuss choice of particle tracking model compared to concentration based model
\end{itemize}

% \section*{What}
% 	\begin{itemize}
% 	\item Study retention mechanism of planktion
% 	\item Theoretical approach with an mixed Eulerian Lagrangien model
% 	\item Demonstrate that "passive" organisms aka particles will ultimatly drift downsteam until and be removed from the estuary
% 	\item Study different retention patterns
% 	\begin{itemize}
% 		\item [placeholder]
% 	\end{itemize}
% 	\item Elbe
% 	\item "alive" particles
% 	\end{itemize}
% 
%  
% 
% \section*{Why}
% \begin{itemize}
% 
% \item particle without retention mechanism - aka "full plankton mode" would drift
%   downstream.
%   change in relative stream position induces changes in salinity
%   hence moves out of ideal salinity zone which results in unfavourable conditions
%   can result of an strong case of death or being outcompeted.
% \item the existance of specialized plankton organism [cite] that only live the tidal section of the river
%   also implies the existance of retention mechanisms since in their case a "reseeding" process from up- or downstream is not possible
% 
% \item the avoid extinction retention mechanism need to be present
%   [cite chesterfield bay study][cite zooplankton study]
% 
%  \item first theories have been tested but so far exclusively with "dead" particles
%   no biophychemical coupled systems
%   "dead" system are of limited explanatory power.
%   concepts like "re-seeding" can not be tested
%   reproduction and dying acts as an amplification in favourable conditions
% 
% \end{itemize}
% 
% \section*{How}
% \begin{itemize}
% 
% 	\item	SCHISM hydrodynamic model coupled with OT 
% 	\begin{itemize}
% 		\item offline coupled hyrdro to lagrangien particle tracker
% 		\item Euler by SCHISM
% 		\item Lagrangien by OceanTracker
% 		\item high speed (24h a 1e6 particles @ 2 minutes)(faster then any model before!)
% 		\item Including biochemical mechanisms
% 	\end{itemize}
% 
% 	\item Ideas to discuss the outcome Results: 
% 	\begin{itemize}
% 		\item Compare retention performance with distribution patterns
% 		\begin{itemize}
% 			\item This will most probably not possible as there is to little 
% 				  observation data to compare spacial distributions
% 			\item 
% 		\end{itemize}
%   		\item Compare "alive" to "dead"
% 	\end{itemize} 
% 
% \end{itemize}

### 2022-02-22 ms.tex
initial commit

\documentclass[a4paper, 10pt, twocolumn]{article}

\usepackage{graphicx}                % optional für Grafiken
\usepackage[font=footnotesize]{caption}
\usepackage[section]{placeins}
%\usepackage{tabularx}               % optional für Tabellen
%\usepackage{multirow}               % optional für Tabellen
\usepackage{url}                	 % optional für Internet Links
\usepackage{hyperref}

%\usepackage[small,bf]{caption3}     % Bildunterschriften

%\usepackage{parskip}

\usepackage{titlesec}
\usepackage{amsmath}                
\usepackage{amssymb}

% temporal packages
\usepackage{lipsum}
\SetLipsumDefault{1}

%\usepackage{enumitem}
%\setlist{nosep} % or \setlist{noitemsep} to leave space around whole list

\usepackage{xcolor}
\newcommand{\highlight}[1]{\colorbox{yellow}{\parbox{\columnwidth}{#1}}}

%\pagestyle{empty}                   % weder Kopf- noch Fußzeile auf 1. Seite
\graphicspath{{figures/}}

\begin{document}

\date{\small\today}                                        % kein Datum auf 1. Seite
\title{
	\vspace{-20mm}\textbf{	\large 
	Plankton Retention mechanism studies at Elbe with IBM}
	}
\author{
	Laurin Steidle\(^1\), Ross Vennel, Jana Hinners \\
	\emph{
		\small
		\(^1\)laurin.steidle@uni-hamburg.de} \\
	\emph{
		\small Universitaet Hamburg, Grosse Elbstrasse 133, Germany}
		}

%--------------------------------------Abstract-------------------------------------------------
\twocolumn[
	\begin{@twocolumnfalse}
		\maketitle
  	\end{@twocolumnfalse}
]

\input{main_text.tex}

\bibliography{references.bib}
\bibliographystyle{unsrt}

\end{document}

### 2022-11-24 main_text.tex
first "full text" draft

@@
-\highlight{placeholder abstract - only for general idea not to be proofread}
+\texttt{placeholder abstract - only for general idea not to be proofread}
+\smallskip
+
 \begin{figure}
@@
 
-\subsection*{Why do we care about retention in estuaries?}
-\begin{itemize}
-    \item Estuaries are important
-    \begin{itemize}
-        \item high productivity relative to surface area
-        \item nutrient source for marine ecosystem
-        \item hatching grounds for not only for estuarine but marine species
-    \end{itemize}
-    %and productive ecosystem - both as fertilizer for marine ecosystems as well as for the global climate system.
-    \item Ecosystem strongly "controlled" by primary producer.
-    \begin{itemize}
-        \item (Do I need to argue/cite something for this?)
-    \end{itemize}
-    \item Primary producers almost exclusively planktonic. [cite]
-          % Suspected reasons: trenching, high velocities on lose soil, high turbidity
-    \item strong hydrodynamic currents %and strong chemical gradiants.
-    \item "retainment" crucial for survival 
-    %for organisms as they would die if washed out to downstream/higher salinity areas.
-    \item "swimming" upstream is impossible or too energy expensive for planktonic organisms
-    \item However, retainment mechanisms are mostly unknown
-    \item Q: How do they survive?
-    % as they are hard to study experimentally.
-\end{itemize}
+\texttt{All texts in this old fashioned typewriter fonds are comments and thoughts on the texts and not part of the actual paper.}
+\smallskip
+
+\subsection*{Estuaries and Phytoplankton}
+
+\texttt{In this section I try to convey: why we care about estuaries and why we focus on phytoplankton and illustrate why question "How does Phytoplankton retain itself in an estuary?" is interesting.}
+
+\textbf{Estuaries are important} both for the climate system as well as important human habitat.
+From a climate perspective they offer a high productivity relative to surface area which results in large amounts of direct carbon cycling as well as their important role as a source of nutrients and hatching grounds for marine ecosystem.
+From a human habitat role they offer easy access to marine ecosystem while still providing fresh water.
+Hence, "key human hatching grounds"
+
+Complex ecosystem: Both marine and freshwater influence, complex topology and hydrodynmics, strong antropogenics make it hard to study and many open questions remain to this day.
+In this study we will take a look at phytoplankton and its mechanism to retain itself in the estuary.
+
+The aquatic part of the ecosystem is strongly controlled by \textbf{phytoplankton} because in its role as a primary producer it acts as prey for most higher trophic levels (cite)
+However phytoplankton as all planktonic organism drift in the current.
+Due to the current in the estuary we expect plankton to be drifting downstream over time and ultimately be washed out into marine high salinity waters.
+Hence, we wonder how phytoplankton manages to survive as a populating in such an environment.
+Assuming the population is not exclusively maintained by upstream plankton that is washed into the estuary.
+Then there must be some sort of retention mechanism that enables phytoplankton to not be washed out.
+While this is expected to be an crucial mechanism for survival it is currently not well understood due to the difficulty in either measuring or modeling such processes.
 
@@
 
-There are three kind of retention studies: observations, theoretical studies, or modelling studies
+\texttt{In this section I try to outline what is known and use it to illustrate the research gap}
+\smallskip
 
-\paragraph{observations}
-\begin{itemize}
-    \item only larger organisms which are not strictly planktonic (fish-larve, mesozooplanktion) [cite]
-    \begin{itemize}
-        \item found diel migration  [cite]
-        \item found tide driven migration
-        \item found "sticking" to "ground" [cite]
-        \item 
-    \end{itemize}
-    \item summary: retention indications found for "planktonic" - tho not directly transferable to phytoplankton
-\end{itemize}
+We categorize the already performed studies on the subject of phytoplankton retention in three groups: observations, theoretical- and realistic-modelling studies.
 
-\paragraph{theoretical}
-\begin{itemize}
-    \item interesting but study simplistic conditions 
-    \item two interesting ones:
-    \begin{itemize}
-        \item investigation of reproduction threshold under exchange cite
-        \item study on "reseeding" due to local upstream currents
-    \end{itemize}
-    summury: both "outgrowing" and "reseeding" are expected to be possible
-\end{itemize}
+\textbf{Observational} studies have found three major mechanism for retention - diel migration, sinking and stickiness.
+Diel migration has been found - moving verticaly up and down with the sun - but only in larger organisms which are not strictly planktonic (fish-larve, mesozooplanktion) [cite]
+Also bentic plankton has been observed that is strongly negativly buoyand aka sinking and sticking to the ground
 
-\paragraph{(particle tracking) models}
-\begin{itemize}
-    \item only one particle tracking model in estuaries which studies zooplankton
-    \item others studies open water or non planktonic organisms i.e. fish
-    \item the existing one has mayor drawbacks - e.g. no reproduction
-    \item found that sinking and diel migration slows outwashing
-    \item non were able to actually study retention in realistic conditions due to technical constraints
-\end{itemize}
+Here we want to highlight two \textbf{theoretical} studies studying the mechanisms of "loss compensation" and "reseeding".
+(cite) theorized a set up in which they  a population is able to maintain itself if the losses due to outwashing can be compensated by growth (cite).
+In a setup where a lake system was modelled they were able to quantify reproductino thersholds that are needed to maintain the population.
+However, this setup also implies that the growing part of the population is somehow retaining their position.
+If the regrowing population is also continuously drifting downstream they will also ultimately die out.
 
-\paragraph{Summary}
-\begin{itemize}
-    \item "surrounding" studies have been made
-    \item none were able to tackle retention mechanisms themselves
-    \item indication for potential mechanisms have been found
-    \item we will investigate retention and reseeding mechanism under different plankton migration behavior
-\end{itemize}
+(cite) studied the concept of "reseeding".
+Here a system is theorized in which a population is able to repopulate an previsouly uncolonized area upstream due to local upstream eddies.
+While they showed that upstream reseeding is conceptually possible they did so in an contrived system.
+
+\smallskip
+\texttt{More detail?}
+\smallskip
+
+Hence, it has been showed both "outgrowing the losses" and "reseeding" are possible retention mechanism but only under certain conditions.
+
+In (cite) they use a \textbf{particle tracking model} also known as Lagragian model to study Zooplankton movement in the San Francisco Bay.
+They were able to show  that sinking and diel migration slows the outwashing process and might be beneficial retention tactic
+However, they did so by ignoring key processes like reproduction and death in an rectangular grid model which further introduces some numerical problems.
+
+To \textbf{summarize} - several concepts for potential retention mechanisms have been found and shown to work in certain conditions.
+However, all approaches used major simplification which didn't allow them to study the full picture.
+We developed a new and more realistic approach to study retention with an particle tracking model combining many of the above outlined features into one model.
 
@@
 \subsection*{general model description}
-\paragraph{working principle}
-\begin{itemize}
-    \item particle tracking
-    \begin{itemize}
-        \item fast
-        \item easy to interpret
-        \item flexible e.g. allows "individual" processes
-        \item realistic representation of motion
-    \end{itemize}
-\end{itemize}
+
+In our study we take a particle tracking approach with a \textbf{lagrangian model} called Oceantracker (cite).
+Particle tracking offers several advantages for our question:
+While particle tracking on unstructured grids is relatively computationally expensive it allows us to reuse already validated hydrodynamic models.
+Which is overall still much faster then using concentration based models.
+Because we are tracking particles we can observer their individual tracks which makes it easier to interpret (plankton is particle)
+It also allows us to include "individual based processes" (not really used currently)
+But most importantly it allows us for a better representation of the behavior close to the edges of the estuary. (100m boxes in concentration based models)
 \begin{figure}
@@
 
-\paragraph{data}
-\begin{itemize}
-    \item input is SCHISM output
-    \item unstructured 3d grid of full estuary (geesthacht to helgoland with side-channels)
-    \item (used) fields:
-    \begin{itemize}
-        \item temperature
-        \item salinity
-        \item water velocity 
-    \end{itemize}
-    \item full 2012 with 1h temporal- and ~5m spacial resolution
-    \item year in reference to others
-\end{itemize}
+\textbf{data}
+The hydrodynamics are taken from a SCHISM model from a Johannes Pein (cite) 
+It is an three dimensional unstructured  grid model representing the full estuary from the weir at geesthacht close to helgoland including side-channels and the harbor area.
+(Look for more details in pein paper that might be interesting)
+It provides a node-based mesh field containing a range of information.
+For our purpose the water velocities represented in a flow field is the most important one
+But also salinity and many other more technical information is used (meh)
+We study the year 2012 with 1h temporal- and dynamically varying spacial resolution.
+
 \begin{figure}
@@
 
-\paragraph{bio-logical mechanisms represented in the model}
-\begin{itemize}
-    \item release
-    \item splitting
-    \item culling
-    \item behavior
-    \begin{itemize}
-        \item sinking
-        \item diel migration
-        \item tidal migration
-    \end{itemize}
-    \item sedimentation/resuspension
-\end{itemize}
+\textbf{bio-logical mechanisms represented in the model}
+On top of the "blanc" particle we add a set of biological features to achieve a more realistic behavior.
+The most important one is their ability to reproduce by creating copies of themselves and their ability to die in certain conditions.
+We also add a set of movement related features.
+This includes a vertical movement behaviour.
+We included constant verticle velocities aka sinking and rising and a time dependent one namely diel-migration.
+Their interaction with the boundaries of our model is represted by a simple sedimentation and resuspension model in the water and a stranding model on land
 
-\begin{figure}
-    \centering
-    \begin{minipage}{.4\columnwidth}
-        \centering
-        \includegraphics[width=.3\columnwidth]{picto resuspension.pdf}
-    \end{minipage}
-    \begin{minipage}{.4\columnwidth}
-    \centering
-        \includegraphics[width=.3\columnwidth]{picto splitting.pdf}
-    \end{minipage}
-    \caption{examples of pictograms for bio processes}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \begin{minipage}{.4\columnwidth}
+%         \centering
+%         \includegraphics[width=.3\columnwidth]{picto resuspension.pdf}
+%     \end{minipage}
+%     \begin{minipage}{.4\columnwidth}
+%     \centering
+%         \includegraphics[width=.3\columnwidth]{picto splitting.pdf}
+%     \end{minipage}
+%     \caption{examples of pictograms for bio processes}
+% \end{figure}
 
-\paragraph{Data Analysis}
-\begin{itemize}
-    \item Based on particle counts in statistical polygons \ref{fig:statistial polygons}
-    \item statistical polygons roughly 100km in length
-    \item calculation of particle "centroid" and whether particles are in a section of the estuary.
-    \item successful retainment is judged by whether or not particles remain alive somewhere in the estuary after one year.
-\end{itemize}
-\begin{figure}
-    \includegraphics[width=\columnwidth]{fixed_particle_culling.pdf}
-    \caption{Example 2d plot of particle counts, centroid and "retainment" threshold}
-\end{figure}
+% \begin{figure}
+%     \includegraphics[width=\columnwidth]{fixed_particle_culling.pdf}
+%     \caption{Example 2d plot of particle counts, centroid and "retainment" threshold}
+% \end{figure}
 
@@
 
-Two major setups with several minor configuration setups in relation to reference configurations
-\begin{figure}
-    \includegraphics[width=\columnwidth]{plot experimental configuration.pdf}
-    \caption{2d grid with boxes for the different cases/dimension. 
-             Same structure as the later sensitivity analysis plot used \ref{fig:retention_results_SA}}
-    \label{fig:experimental_configuration}
-\end{figure}
+% \begin{figure}
+%     \includegraphics[width=\columnwidth]{plot experimental configuration.pdf}
+%     \caption{2d grid with boxes for the different cases/dimension. 
+%              Same structure as the later sensitivity analysis plot used \ref{fig:retention_results_SA}}
+%     \label{fig:experimental_configuration}
+% \end{figure}
 
-\paragraph{reference run - dead\&passive}
-\begin{itemize}
-    \item no reproduction 
-    \item no behavior just drifting
-\end{itemize}
+\textbf{downstream retention}
+The first and major experiment of this studies examines retention mechanisms under different scenarios.
+General idea is to release a population of plankton at the weir in geesthacht and examine how the population distrubutes itself over the estuary and whether it is able to maintain its population size.
+We examine the effect of several mechanism. These include:
+\textit{Reproduction} represented as a chance of duplicating a particle,
+\textit{vertical migration} where two modes have been implemented. A constant up or downward drift and diel migration where the drift speed and direction is based on the phase of the sun.
+Particles can \textit{strand and resuspend} from the shore when the water level rises again.
+They can also \textit{sink and suspend} on the bottom and resuspend based on a critical-friction-velocity concept.
+Particles are removed if they either reach high salinity waters implemented as a mortality chance above $10 PSU$,
+when they are stranded until they dry out which is represented by killing all particles which have not been rewettet in the last $7$ days,
+and when they are starved for light which we defined as removing all particles which have been in low-light areas longer then a total of $14$ days where presence in high-light areas is subtracted from the starvation budget.
+\texttt{Hallo, mein Name ist Kleist.}
+The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes and to allow for a realistic trajectories close to boundaries.
 
-\paragraph{downstream retention}
-\begin{itemize}
-    \item studies retention mechanisms under different behavior/growth
-    \item sensitivity analysis with the following dimensions
-    \begin{itemize}
-        \item growth rate 
-        \item sinking
-        \item diel migration
-        \item tidal migration
-    \end{itemize} 
-    \item release location description
-\end{itemize}
+\smallskip
+\texttt{More detail needed? - or should that go into the appendix?}
+\smallskip
 
-\paragraph{upstream migration}
-\paragraph{downstream retention}
+The observed particle batch representing out phytoplankton population is released at new years eve in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
+We evaluate their \textbf{retention success} primarily by measuring the amount of plankton in different regions in the estuary over time.
+If the population survives a full year and its population is over a certain threshold we consider them "successfully retainted".
 
-(Note: This is the "release downstream and check if we find some upstream" experiment)
-\begin{itemize}
-    \item studies reseeding mechanisms of upstream waters under different behavior/growth
-    \item same SA dimensions as above
-    \item release location description
-\end{itemize}
+To count the particles we use "\textbf{statistical polygons}".
+We use to split the estuaries in along-shore section. See  \ref{fig:statistial polygons}
+\texttt{Well, actually we don't really use the individual polygons in any of the metrics but only for debugging and visualization purposes. Are they interesting anyway?}
+
+We also log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
+This allows us to compare the successful (died old) and unsuccessful (died young) particles 
+The logged observables are measured every 12 hours and include location, status, water level, age, distance traveled
+
+We use this set-up in a range of scenarios and compare them to each other.
+We call the first and simples scenario "dust". Here we don't allow for any reproduction and give all particles a neutral buoyancy.
+This is our reference for any retention strategy. Any beneficial strategy must perform better then these passive tracers.
+
+In the second scenario - referred to as "migrating" - re-enable reproduction and also give the particles a range of different buoyancies resulting in positive, negative or in the case of the diel-mirgrating particles oscillating vertical velocities.
+We also allow reproduction with a range of different growth rates.
+
+The third scenario has been added for illustration purposes.
+Here, we use the same set-up as in the second scenario with the only exception of disabling reproduction for particles stranded on the shores of the estuaries.
+We call this scenario "waiting-for-resuspension"
+\label{txt:waiting-for-resuspsension}
+But more on that later.
+
+The tested parameters described above like reproduction and vertical movement can be found in table \ref{tab:downstream_retention_params}.
+\smallskip
+\texttt{The table doesn't exist yet.}
+\smallskip
 
@@
 
-\subsection*{downstream retention}
+\subsection*{Retentions experiments}
+
+\subsubsection*{Scenario: "dust"}: 
+\begin{figure}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{total particles counts for scenario "dust" for all tested parameters}
+    \label{fig:dust_particle_counts}
+\end{figure}
+In fig \ref{fig:dust_particle_counts} the particle counts measured over time it the estuary are shown.
+We can clearly see that the total particle count drops to zero - in most cases already after less then two months.
+From this we deduce that particles are constantly getting washed out and do not find "sheltered areas" in which they can remain for a long time.
+This also means that reproduction is a necessary condition for the population to maintain their population even if all mortality sources except high-salinity are ignored.
+
+\subsubsection*{Scenario: "migration"}: 
+\begin{figure}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{total particles counts for scenario "migration" with a monotonic migration pattern for all tested parameters}
+    \label{fig:monotonic_particle_counts}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{total particles counts for scenario "migration" with a diel migration pattern for all tested parameters}
+    \label{fig:diel_particle_counts}
+\end{figure}
+
+In fig \ref{fig:monotonic_particle_counts} and \ref{fig:diel_particle_counts} we can see that the population manages to maintain themselves in certain conditions.
+The fact that we found conditions for survival in this simplistic setup is in itself interesting for two major reasons.
+First, all previsous studies implementing a phytoplankton in an Elbe model required constant reseeding either up and/or downstream.
+Hence, we are the first study that is able to provide an estimate for the conditions necessary for "in-estuar"-survival.
+
+\texttt{ The secondly might be better suited for the discussion.}
+
+Secondly, while we assume ad libitum nutrient conditions for the phytoplankton we do not implement any sub-grid-resolution structure on the shores.
+Hence, the river banks are perfectly flat sand banks without vegetation which might aid the retention of phytoplankton.
+
+\texttt{Add comparison of monotonic upwards to downwards, monotonic to diel, and the discuss the relative growth rate thresholds}
+
+Lets now take a closer look at the subset of pytoplankton that was able to retain themselves compared to those getting washed out.
+We divide the population now into a long-living and short-living set with an threshold age of 3 months.
+We compare these two sets now by their water depth, distance traveled, and ratio of time stranded to drifting.
+You can see these measurement visualized in the boxplots \ref{fig:migration-long-vs-short}
 
 \begin{figure}
-    \includegraphics[width=\columnwidth]{plot results sa.pdf}
-    \caption{2d grid with boxes for the different cases/dimension. 
-             each box represents one configuration and whether they were able to retain by color code.
-             I.e. a bunch of red and green boxes with axis labels}
-    \label{fig:retention_results_SA}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{Boxplot of long-vs-short showing water depth}
+    \label{fig:migration-long-vs-short}
 \end{figure}
 
+We can consistently see that the long living subset is in shallower water, does move less, and is a significantly longer stranded then drifting.
+In fig. \ref{fig:migration-long-vs-short-heatmap} we show the special distribution of the long- and short-living subset.
+
 \begin{figure}
-    \includegraphics[width=\columnwidth]{plot result threshold change.pdf}
-    \caption{bar plot of the $\frac{\partial \mu_{min}}{\partial v_{verticle}}$ relations}
-    \label{fig:retention_results_dmu}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{heatmap of long-living to short-living-particles}
+    \label{fig:migration-long-vs-short-heatmap}
 \end{figure}
 
-\begin{itemize}
-    \item discuss which configuration maintained a population and which not
-    \item describe growth threshold under different behavior scenarios
-    \item describe $\frac{\partial \mu_{min}}{\partial v_{verticle}}$ for sinking, diel-, tidal migration
-\end{itemize}
+\texttt{show zoom in with water depth?}
 
-\paragraph{reproduction dimension}
-\begin{itemize}
-    \item discuss which configuration managed to colonize upstream areas
-    \item discuss threshold growth rates and the impact of behavior patters similar to above
-\end{itemize}
+\subsubsection*{Scenario: "waiting-for-resuspension"}
 
-\paragraph{Discuss Results}
-contextualize the results 
-\begin{itemize}
-    \item doubling times of reproduction of threshold reproduction rates (for context)
-    \item threshold reproduction rates in relation to "realistic" reproduction rates summer/winter - aka are these mechanisms sufficient to allow survival
-    \item discuss effectiveness of migration behavior - most effective mechanisms should a priori be realized in nature 
-    \item (once all results are done - check if we can make predictions that are experimentally testable. e.g. average sinking velocities)
-    \item reseeding (with the preliminary results) is not possible - no upstream migration above the harbor aka non-tidal-dominated-areas
-\end{itemize}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{total particles counts for scenario "waiting-for-resuspension" with a monotonic migration pattern for all tested parameters}
+    \label{fig:stranded-monotonic-particle-counts}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{total particles counts for scenario "waiting-for-resuspension" with a diel migration pattern for all tested parameters}
+    \label{fig:stranded-diel-particle-counts}
+\end{figure}
+
+In fig \ref{fig:stranded-monotonic-particle-counts} and \ref{fig:stranded-diel-particle-counts} we again see the total particle counts.
+Keep in mind that the only difference in the model configuration to the "migration" run is that phytoplankton growth has been disabled when the plankton is stranded.
+We can clearly see that phytoplankton dies out in these conditions for the full parameter space.
+From this we deduce that the shallow areas are not only crucial for improved retention but also a necessary hatching ground.
+
+To summarize the retention experiments: We saw that the population is ultimately washed out and perishes if we surpress reproduction.
+We also saw that with reproduction phytoplankton is able to maintain their population strain without up- or downstream reseeding under certain conditions!
+In the last experiment we showed that the shallow water areas are not only helpfull but crucial for the survival of the population.
+
+\textbf{upstream migration}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{example-image-a}
+    \caption[]{particle counts in upstream statistical polygons}
+    \label{fig:upstream-migration-statistical-poly-counts}
+\end{figure}
+As previsouly described - In the upstream migration experiment we divide the estuary in short ($\sim 50km$) sections.
+We release particles in all of these sections and count their occurence in other sections.
+In fig \ref{fig:upstream-migration-statistical-poly-counts} the particle counts for the two upstream sections from the release location.
+We consider upstream migration to happen when plankton moves two sections up and is able to retain itself there.
+\texttt{Do I need to explain why two polygons and why being there only briefly won't work?}
+
+\texttt{No results yet with the proper dispersion model. In the previous run upstream migration did not occur and was considered to be a negligible mechanism in the retention process.}
 
 \section*{Discussion}
+\texttt{The discussion is currently only a list of bullet points that I considered to discuss so far. It is by no means complete or currently not worth reading.}
 \begin{itemize}
@@
         \item simplified bottom/sediment interaction
-        \item no "small object" representation e.g. vegetation or "rocks"
-        \item simplistic bio representation. does not allow for precise quantification but only indicates bounds for values
+        \item no "small object" representation e.g. vegetation or "rocks" $\rightarrow$ implices underestimation of "retention"
+        \item simplistic bio representation. does not allow for precise quantification but only indicates bounds aka "even in ideal conditions they couldn't survive if ..."
         \item other minor issues - temporal resolution, particle out-of-bounds cases etc
-    \end{itemize}
         \item discuss choice of particle tracking model compared to concentration based model
     \end{itemize}
-
-% \section*{What}
-% 	\begin{itemize}
-% 	\item Study retention mechanism of planktion
-% 	\item Theoretical approach with an mixed Eulerian Lagrangien model
-% 	\item Demonstrate that "passive" organisms aka particles will ultimatly drift downsteam until and be removed from the estuary
-% 	\item Study different retention patterns
-% 	\begin{itemize}
-% 		\item [placeholder]
-% 	\end{itemize}
-% 	\item Elbe
-% 	\item "alive" particles
-% 	\end{itemize}
-% 
-%  
-% 
-% \section*{Why}
-% \begin{itemize}
-% 
-% \item particle without retention mechanism - aka "full plankton mode" would drift
-%   downstream.
-%   change in relative stream position induces changes in salinity
-%   hence moves out of ideal salinity zone which results in unfavourable conditions
-%   can result of an strong case of death or being outcompeted.
-% \item the existance of specialized plankton organism [cite] that only live the tidal section of the river
-%   also implies the existance of retention mechanisms since in their case a "reseeding" process from up- or downstream is not possible
-% 
-% \item the avoid extinction retention mechanism need to be present
-%   [cite chesterfield bay study][cite zooplankton study]
-% 
-%  \item first theories have been tested but so far exclusively with "dead" particles
-%   no biophychemical coupled systems
-%   "dead" system are of limited explanatory power.
-%   concepts like "re-seeding" can not be tested
-%   reproduction and dying acts as an amplification in favourable conditions
-% 
-% \end{itemize}
-% 
-% \section*{How}
-% \begin{itemize}
-% 
-% 	\item	SCHISM hydrodynamic model coupled with OT 
-% 	\begin{itemize}
-% 		\item offline coupled hyrdro to lagrangien particle tracker
-% 		\item Euler by SCHISM
-% 		\item Lagrangien by OceanTracker
-% 		\item high speed (24h a 1e6 particles @ 2 minutes)(faster then any model before!)
-% 		\item Including biochemical mechanisms
-% 	\end{itemize}
-% 
-% 	\item Ideas to discuss the outcome Results: 
-% 	\begin{itemize}
-% 		\item Compare retention performance with distribution patterns
-% 		\begin{itemize}
-% 			\item This will most probably not possible as there is to little 
-% 				  observation data to compare spacial distributions
-% 			\item 
-% 		\end{itemize}
-%   		\item Compare "alive" to "dead"
-% 	\end{itemize} 
-% 
-% \end{itemize}
+    \item conclusions that we draw from this:
+    \begin{itemize}
+        \item from a population-surival perspective the plankton that does not drift to shallow water are "dead man walking" as all decendents of these particles will ultimately die.
+        \item we need cross-shore sampling to better understand this. currently all specially resolved sampling strategies are along shor - aka in the center of the river
+        \item shallow water is crucial for the survival. hence the protection and restoration of shallow regions is crucial for ecosystem healthiness
+    \end{itemize}
+\end{itemize}

### 2023-06-26 main_text.tex
"pre adding pictures and actually writing it" draft

@@
-\texttt{placeholder abstract - only for general idea not to be proofread}
-\smallskip
-
-\begin{figure}
-    \includegraphics[width=\columnwidth]{picto visual abstract.pdf}
-\end{figure}
-Estuaries are a challenging for plankton.
-Gradients are steep and currents are strong.
-Retention mechanisms crucial process for survival.
-However, very little experimental and theoretical evidence has been gathered.
-We will conduct a model study.
-First we will show the importance of retention mechanism for plankton in the Elbe.
-Secondly we will study different retention mechanism which are suspected to be applied by organism.
-We compare the outcome of the different experiment and judge their effectiveness.
-
 \section*{Introduction}
-
-\texttt{All texts in this old fashioned typewriter fonds are comments and thoughts on the texts and not part of the actual paper.}
-\smallskip
 
@@
 
-\texttt{In this section I try to convey: why we care about estuaries and why we focus on phytoplankton and illustrate why question "How does Phytoplankton retain itself in an estuary?" is interesting.}
+\textbf{Estuaries are important} both for the climate system as well as for anthropogenic use.
+From a climate perspective they offer a high productivity relative to surface area which results in large amounts of direct carbon cycling as well as their important role as a source of nutrients and hatching grounds for marine ecosystem.
+(cite)
+They are also strongly anthropogenically influenced through e.g.
+harbors, diking and fishing which can be seen as a stressor on the ecosystem but also adds am anthropogenic dependency on the well being of the ecosystem.
 
-\textbf{Estuaries are important} both for the climate system as well as important human habitat.
-From a climate perspective they offer a high productivity relative to surface area which results in large amounts of direct carbon cycling as well as their important role as a source of nutrients and hatching grounds for marine ecosystem.
-From a human habitat role they offer easy access to marine ecosystem while still providing fresh water.
-Hence, "key human hatching grounds"
+However, estuaries are also complex and strongly dynamic ecosystem such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic changes to this day.
+This complexity arises from the topology and the resulting hydrodynmics at the fresh-salt water interface.
 
-Complex ecosystem: Both marine and freshwater influence, complex topology and hydrodynmics, strong antropogenics make it hard to study and many open questions remain to this day.
-In this study we will take a look at phytoplankton and its mechanism to retain itself in the estuary.
-
-The aquatic part of the ecosystem is strongly controlled by \textbf{phytoplankton} because in its role as a primary producer it acts as prey for most higher trophic levels (cite)
-However phytoplankton as all planktonic organism drift in the current.
+As for most ecosystem - their dynamics are strongly controlled by the primary producers.
+However phytoplankton, the major aquatic primary producer drifts in the current as all planktonic organism.
 Due to the current in the estuary we expect plankton to be drifting downstream over time and ultimately be washed out into marine high salinity waters.
 Hence, we wonder how phytoplankton manages to survive as a populating in such an environment.
-Assuming the population is not exclusively maintained by upstream plankton that is washed into the estuary.
-Then there must be some sort of retention mechanism that enables phytoplankton to not be washed out.
-While this is expected to be an crucial mechanism for survival it is currently not well understood due to the difficulty in either measuring or modeling such processes.
+If we assume that the population is not exclusively maintained by upstream plankton that is washed into the estuary
+ then there must be some sort of retention mechanism that enables a phytoplankton population to not be washed out.
+While this is expected to be an crucial mechanism for survival  it is currently not well understood 
+due to the difficulty in either measuring or modeling such processes.
 
-\subsection*{Previous studies - aka - what is known}
-
-\texttt{In this section I try to outline what is known and use it to illustrate the research gap}
-\smallskip
+\subsection*{what is known}
 
@@
 \textbf{Observational} studies have found three major mechanism for retention - diel migration, sinking and stickiness.
-Diel migration has been found - moving verticaly up and down with the sun - but only in larger organisms which are not strictly planktonic (fish-larve, mesozooplanktion) [cite]
-Also bentic plankton has been observed that is strongly negativly buoyand aka sinking and sticking to the ground
+Diel migration (moving verticaly up and down with the sun) has been found but so far only shown for larger estuarine organisms which are not strictly planktonic i.e. zooplankton [cite]
+Also benthic plankton has been observed to be strongly negatively buoyant, hence sinking and sticking to the ground.
 
-Here we want to highlight two \textbf{theoretical} studies studying the mechanisms of "loss compensation" and "reseeding".
+Next want to highlight two more \textbf{theoretical} studies studying the mechanisms of "loss compensation" and "reseeding".
 (cite) theorized a set up in which they  a population is able to maintain itself if the losses due to outwashing can be compensated by growth (cite).
-In a setup where a lake system was modelled they were able to quantify reproductino thersholds that are needed to maintain the population.
+In a setup where a lagoon like ecosystem was modelled they were able to quantify reproduction thresholds that are needed to maintain the population.
 However, this setup also implies that the growing part of the population is somehow retaining their position.
@@
 (cite) studied the concept of "reseeding".
-Here a system is theorized in which a population is able to repopulate an previsouly uncolonized area upstream due to local upstream eddies.
-While they showed that upstream reseeding is conceptually possible they did so in an contrived system.
+Here a system is theorized in which a population is able to repopulate an previously 
+uncolonized area upstream due to local upstream currents e.g. through larger eddies.
+While they showed that upstream reseeding is conceptually possible they did so in a special contrived system.
+Whether this is an actual realized phenomenon depends on the topology and hydrodynamic of the estuary.
 
@@
 In (cite) they use a \textbf{particle tracking model} also known as Lagragian model to study Zooplankton movement in the San Francisco Bay.
-They were able to show  that sinking and diel migration slows the outwashing process and might be beneficial retention tactic
-However, they did so by ignoring key processes like reproduction and death in an rectangular grid model which further introduces some numerical problems.
+They were able to show  that sinking and diel migration slows the outwashing process and might be beneficial retention tactic.
+However, they did so by ignoring key processes like reproduction and death in an rectangular grid model
+which further introduces some numerical artefact.
 
-To \textbf{summarize} - several concepts for potential retention mechanisms have been found and shown to work in certain conditions.
+To \textbf{summarize} - several concepts for potential retention mechanisms have been found
+ and shown to work in certain conditions.
 However, all approaches used major simplification which didn't allow them to study the full picture.
-We developed a new and more realistic approach to study retention with an particle tracking model combining many of the above outlined features into one model.
+We developed a new and more realistic approach to study retention with an particle tracking model combining
+ many of the above outlined features into one model.
+
+\texttt{I don't have a proper "thats our hypothesis" part in my introduction which is considered to be bad style I think.
+        However, the way I frame it currently I don't really have a proper hypothesis. 
+        I approached it more as a kind of "lets see if we can find some noteworthy patterns and report them"
+        Personally I think this kind of playing around is worth communicating as well but I am not sure how the community feels about that.}
 
@@
 Particle tracking offers several advantages for our question:
-While particle tracking on unstructured grids is relatively computationally expensive it allows us to reuse already validated hydrodynamic models.
-Which is overall still much faster then using concentration based models.
-Because we are tracking particles we can observer their individual tracks which makes it easier to interpret (plankton is particle)
-It also allows us to include "individual based processes" (not really used currently)
-But most importantly it allows us for a better representation of the behavior close to the edges of the estuary. (100m boxes in concentration based models)
+While particle tracking on unstructured grids has been relatively computationally expensive until recently (cite)
+ it allows us to reuse already validated hydrodynamic models.
+Reusing already computed and calibrated hydrodynamical models is overall much faster then solving the advection-diffusion again.
+
+Additionally, because we are tracking particles individually we are able to observer their individual tracks.
+This makes it much easier and intuitive to interpret the results and allows us to  include "individual based processes".
+
+\texttt{better representation of the particle behavior close to the edges of the estuary?}
+
 \begin{figure}
@@
 
-\textbf{data}
-The hydrodynamics are taken from a SCHISM model from a Johannes Pein (cite) 
-It is an three dimensional unstructured  grid model representing the full estuary from the weir at geesthacht close to helgoland including side-channels and the harbor area.
-(Look for more details in pein paper that might be interesting)
+The hydrodynamics \textbf{data} is taken from a SCHISM model from a Johannes Pein (cite).
+It is an three dimensional unstructured  grid model representing the full Elbe estuary from the weir at Geesthacht close to helgoland including side-channels and the harbor area.
+% (Look for more details in pein paper that might be interesting)
 It provides a node-based mesh field containing a range of information.
 For our purpose the water velocities represented in a flow field is the most important one
-But also salinity and many other more technical information is used (meh)
+But also salinity and many other more technical information is used (See appendix \texttt{not yet added})
 We study the year 2012 with 1h temporal- and dynamically varying spacial resolution.
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{Artboard 1.png}
+    \includegraphics[width=\columnwidth]{dfg_foto_wettbewerbg_laurin_legendless_steidle.png}
     \caption{model domain visualization - full range grid including statistical polygons for later reference and example of a scalar field (e.g. temperature)}
@@
 
-\textbf{bio-logical mechanisms represented in the model}
-On top of the "blanc" particle we add a set of biological features to achieve a more realistic behavior.
-The most important one is their ability to reproduce by creating copies of themselves and their ability to die in certain conditions.
+On top of the "blanc" particle we add a set of textbf{biological features} to achieve a more realistic behavior.
+The most important one is their ability to reproduce by creating copies of themselves 
+and their ability to die in certain conditions.
 We also add a set of movement related features. 
-This includes a vertical movement behaviour.
-We included constant verticle velocities aka sinking and rising and a time dependent one namely diel-migration.
-Their interaction with the boundaries of our model is represted by a simple sedimentation and resuspension model in the water and a stranding model on land
+This includes a vertical movement behavior. 
+We included constant vertical velocities i.e. sinking and rising and a time dependent one namely diel-migration.
+Their interaction with the boundaries of our model is represented by a simple sedimentation and resuspension model in the water and in a stranding model on land
 
@@
 The first and major experiment of this studies examines retention mechanisms under different scenarios.
-General idea is to release a population of plankton at the weir in geesthacht and examine how the population distrubutes itself over the estuary and whether it is able to maintain its population size.
-We examine the effect of several mechanism. These include:
+General idea is to release a population of plankton at the beginning of the estuary represented by the weir in Geesthacht 
+and examine how the population distrubutes itself over the estuary and whether it is able to maintain its population size.
+With this approach we study the effect of several mechanism. These include:
 \textit{Reproduction} represented as a chance of duplicating a particle,
-\textit{vertical migration} where two modes have been implemented. A constant up or downward drift and diel migration where the drift speed and direction is based on the phase of the sun.
-Particles can \textit{strand and resuspend} from the shore when the water level rises again.
-They can also \textit{sink and suspend} on the bottom and resuspend based on a critical-friction-velocity concept.
+\textit{vertical migration} where two modes have been implemented. 
+A constant up or downward drift and diel migration where the drift speed and direction is based on the phase of the sun.
+The particle \textit{stranding and resuspension} as well as \textit{suspend and resuspension to and from the bottom}
+ based on a critical-friction-velocity model.
 Particles are removed if they either reach high salinity waters implemented as a mortality chance above $10 PSU$,
-when they are stranded until they dry out which is represented by killing all particles which have not been rewettet in the last $7$ days,
-and when they are starved for light which we defined as removing all particles which have been in low-light areas longer then a total of $14$ days where presence in high-light areas is subtracted from the starvation budget.
-\texttt{Hallo, mein Name ist Kleist.}
-The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes and to allow for a realistic trajectories close to boundaries.
+when they are stranded until they dry out which is represented by killing all particles which have not been rewettet
+ in the last $7$ days, and when they are starved for light which we defined as removing all particles which have been
+ in low-light areas longer then a total of $14$ days where presence in high-light areas is subtracted from the starvation budget.
+\texttt{Sentences to long to assume - good enough for now}
+The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes 
+and to allow for a realistic trajectories close to boundaries.
 
@@
 
-The observed particle batch representing out phytoplankton population is released at new years eve in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
+The observed particle batch representing our phytoplankton population is released at new years eve in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
 We evaluate their \textbf{retention success} primarily by measuring the amount of plankton in different regions in the estuary over time.
-If the population survives a full year and its population is over a certain threshold we consider them "successfully retainted".
+If the population survives a full year and its population is over a certain threshold we consider them "successfully retained".
 
-To count the particles we use "\textbf{statistical polygons}".
-We use to split the estuaries in along-shore section. See  \ref{fig:statistial polygons}
-\texttt{Well, actually we don't really use the individual polygons in any of the metrics but only for debugging and visualization purposes. Are they interesting anyway?}
+To count the particles we use "\textbf{statistical polygons}" that split the estuaries in along-shore sections. 
+See  \ref{fig:statistial polygons}.
+We also log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
+This allows us to compare the successful (died older then three months) and unsuccessful (died younger then then three months ) plankton.
+The logged observables are measured every 12 hours starting at midnight.
 
-We also log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
-This allows us to compare the successful (died old) and unsuccessful (died young) particles 
-The logged observables are measured every 12 hours and include location, status, water level, age, distance traveled
+\medskip
 
 We use this set-up in a range of scenarios and compare them to each other.
-We call the first and simples scenario "dust". Here we don't allow for any reproduction and give all particles a neutral buoyancy.
-This is our reference for any retention strategy. Any beneficial strategy must perform better then these passive tracers.
+We call the first and simplest scenario "dust". 
+Here we don't allow for any reproduction and give all particles a neutral buoyancy.
+This is our reference for any retention strategy. 
+For a retention strategy to be considered beneficial it must perform better then these passive tracers.
 
-In the second scenario - referred to as "migrating" - re-enable reproduction and also give the particles a range of different buoyancies resulting in positive, negative or in the case of the diel-mirgrating particles oscillating vertical velocities.
-We also allow reproduction with a range of different growth rates.
+In the second scenario referred to as "migration" reproduction is re-enabled with a range of different growth rates 
+and a range of different buoyancies i.e. vertical velocities is given to the plankton.
 
@@
 Here, we use the same set-up as in the second scenario with the only exception of disabling reproduction for particles stranded on the shores of the estuaries.
-We call this scenario "waiting-for-resuspension"
-\label{txt:waiting-for-resuspsension}
-But more on that later.
+We call this scenario "ready-for-liftoff"
+\label{txt:ready-for-liftoff}
 
+\textbf{upstream migration}
+In the second and minor experiment we test the hypothesis of "reseeding" - whether organisms can recolonize previously lost upstream habitats.
+
+\textbf{parameter overview}
 The tested parameters described above like reproduction and vertical movement can be found in table \ref{tab:downstream_retention_params}.
@@
 
+\texttt{The results presented below are not yet final and not-visualized (hence the placeholder figures)
+        but true and tested in the current version of the model.
+        Hence, we assume they might chance in some details but stay true with the improved dispersion model.
+        \newline
+        I also intermingled the pure results with some interpretations. I think this is considered bad style.
+        I plan to disentangle this more once I have the actual results to discuss if the journal style requires it.}
+
 \subsection*{Retentions experiments}
 
-\subsubsection*{Scenario: "dust"}: 
 \begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
-    \caption[]{total particles counts for scenario "dust" for all tested parameters}
-    \label{fig:dust_particle_counts}
-\end{figure}
-In fig \ref{fig:dust_particle_counts} the particle counts measured over time it the estuary are shown.
-We can clearly see that the total particle count drops to zero - in most cases already after less then two months.
-From this we deduce that particles are constantly getting washed out and do not find "sheltered areas" in which they can remain for a long time.
-This also means that reproduction is a necessary condition for the population to maintain their population even if all mortality sources except high-salinity are ignored.
-
-\subsubsection*{Scenario: "migration"}: 
-\begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
+    \includegraphics[width=\columnwidth]{diel_success.png}
     \caption[]{total particles counts for scenario "migration" with a monotonic migration pattern for all tested parameters}
@@
 \end{figure}
-\begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
-    \caption[]{total particles counts for scenario "migration" with a diel migration pattern for all tested parameters}
-    \label{fig:diel_particle_counts}
-\end{figure}
+% \begin{figure}
+%     \includegraphics[width=\columnwidth]{example-image-a}
+%     \caption[]{total particles counts for scenario "migration" with a diel migration pattern for all tested parameters}
+%     \label{fig:diel_particle_counts}
+% \end{figure}
 
-In fig \ref{fig:monotonic_particle_counts} and \ref{fig:diel_particle_counts} we can see that the population manages to maintain themselves in certain conditions.
+In fig \ref{fig:dust_particle_counts} the particle counts measured over time it the estuary are shown.
+We can clearly see that the total particle count drops to zero - in most cases already after less then two months.
+From this we deduce that particles are constantly getting washed out and do not find "sheltered areas" 
+in which they can remain for a long time.
+This also means that reproduction is a necessary condition for the population to maintain their size 
+even if all mortality sources except high-salinity are ignored.
+
+In fig \ref{fig:monotonic_particle_counts} and \ref{fig:diel_particle_counts} we can see 
+that the population manages to maintain themselves in certain conditions.
 The fact that we found conditions for survival in this simplistic setup is in itself interesting for two major reasons.
@@
 Hence, we are the first study that is able to provide an estimate for the conditions necessary for "in-estuar"-survival.
-
 \texttt{ The secondly might be better suited for the discussion.}
-
 Secondly, while we assume ad libitum nutrient conditions for the phytoplankton we do not implement any sub-grid-resolution structure on the shores.
@@
 
-Lets now take a closer look at the subset of pytoplankton that was able to retain themselves compared to those getting washed out.
-We divide the population now into a long-living and short-living set with an threshold age of 3 months.
-We compare these two sets now by their water depth, distance traveled, and ratio of time stranded to drifting.
+Taking a closer look at the subset of pytoplankton that was able to retain themselves compared to those getting washed out.
 You can see these measurement visualized in the boxplots \ref{fig:migration-long-vs-short}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
+    \includegraphics[width=\columnwidth]{dbf_bar.png}
     \caption[]{Boxplot of long-vs-short showing water depth}
@@
 We can consistently see that the long living subset is in shallower water, does move less, and is a significantly longer stranded then drifting.
-In fig. \ref{fig:migration-long-vs-short-heatmap} we show the special distribution of the long- and short-living subset.
+In fig. \ref{fig:migration-long-vs-short-heatmap} we show the special distribution of the long- and short-living subset
+ also indication their predominance of long living phytoplankton close to the shores and in side channels.
 
 \begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
+    \includegraphics[width=\columnwidth]{heatmap.png}
     \caption[]{heatmap of long-living to short-living-particles}
@@
 
-\subsubsection*{Scenario: "waiting-for-resuspension"}
+\subsubsection*{Scenario: "ready-for-liftoff"}
 
 \begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
-    \caption[]{total particles counts for scenario "waiting-for-resuspension" with a monotonic migration pattern for all tested parameters}
+    \includegraphics[width=\columnwidth]{diel_success.png}
+    \caption[]{total particles counts for scenario "ready-for-liftoff" with a monotonic migration pattern for all tested parameters}
     \label{fig:stranded-monotonic-particle-counts}
 \end{figure}
-\begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
-    \caption[]{total particles counts for scenario "waiting-for-resuspension" with a diel migration pattern for all tested parameters}
-    \label{fig:stranded-diel-particle-counts}
-\end{figure}
 
-In fig \ref{fig:stranded-monotonic-particle-counts} and \ref{fig:stranded-diel-particle-counts} we again see the total particle counts.
+In fig \ref{fig:stranded-monotonic-particle-counts} we again show the total particle counts.
 Keep in mind that the only difference in the model configuration to the "migration" run is that phytoplankton growth has been disabled when the plankton is stranded.
@@
 
-To summarize the retention experiments: We saw that the population is ultimately washed out and perishes if we surpress reproduction.
-We also saw that with reproduction phytoplankton is able to maintain their population strain without up- or downstream reseeding under certain conditions!
-In the last experiment we showed that the shallow water areas are not only helpfull but crucial for the survival of the population.
+To summarize the retention experiments: We saw that the population is ultimately washed out and perishes if we suppress reproduction.
+We also saw that reproducing phytoplankton is able to maintain their population strain without up- or downstream reseeding under certain conditions!
+We further showed that the shallow water areas are not only helpful for retention but crucial for the survival of the population.
 
 \textbf{upstream migration}
+
+% \highlight{I dropped this part for the time being.}
 \begin{figure}
@@
 \end{figure}
-As previsouly described - In the upstream migration experiment we divide the estuary in short ($\sim 50km$) sections.
-We release particles in all of these sections and count their occurence in other sections.
-In fig \ref{fig:upstream-migration-statistical-poly-counts} the particle counts for the two upstream sections from the release location.
+
+In the upstream migration experiment we divide the estuary in short ($\sim 50km$) sections.
+We release particles in all of these sections and count their occurrence in other sections.
+As previously described - in fig \ref{fig:upstream-migration-statistical-poly-counts} we see the 
+the particle counts for the two upstream sections from the release location.
 We consider upstream migration to happen when plankton moves two sections up and is able to retain itself there.
@@
 
-\texttt{No results yet with the proper dispersion model. In the previous run upstream migration did not occur and was considered to be a negligible mechanism in the retention process.}
+\texttt{No results yet with the proper dispersion model. 
+        In the previous run upstream migration did not occur and was considered 
+        to be a negligible mechanism in the retention process.
+        However, this might chance with the enabling of tidal pumping mechanisms.}
 
 \section*{Discussion}
+
+\subsection*{Interpretation and contextualization of Results}
+
+\begin{itemize}
+    \item retention success threshold as a measure of estimating outwashing losses
+    \item very low outwashing rates imply unfavorable conditions for phytoplankton? or maybe the conditions are good and they grow nicely but not enough to compensate for losses in harbor?
+    \item 
+\end{itemize}
+
+\subsection*{Limitations of the model}
+
 \texttt{The discussion is currently only a list of bullet points that I considered to discuss so far. It is by no means complete or currently not worth reading.}

### 2023-06-26 ms.tex
first "full text" draft
"pre adding pictures and actually writing it" draft

@@
 
+% verbatim with line line wrapping
+% \usepackage{listings}
+% \lstset{
+%    breaklines=true,
+%    basicstyle=\ttfamily}
+
 \usepackage{titlesec}
@@
 \usepackage{amssymb}
+\usepackage{abstract}
 
@@
 
-%--------------------------------------Abstract-------------------------------------------------
+%--------------------------------------Abstract------------------------------------------------
+
 \twocolumn[
@@
 		\maketitle
+
+		\begin{abstract}
+			\begin{center}
+				\texttt{All texts in this typewriter fond are drafting comments and thoughts on the text itself and not part of the actual paper.}
+				\smallskip
+			\end{center}
+
+			\begin{center}
+				\texttt{placeholder abstract - not to be proofread}
+				\smallskip
+			\end{center}
+
+			% \begin{figure}
+			% 	\includegraphics[width=\columnwidth]{picto visual abstract.pdf}
+			% \end{figure}
+			Estuaries are a challenging environment for plankton.
+			Gradients are steep and currents are strong.
+			Retention mechanisms are crucial for their survival.
+			However, very little experimental and theoretical evidence has been gathered on the mechanism of phytoplankton retention.
+			We will conduct a model study.
+			First we will show the importance of retention mechanism for plankton in the Elbe.
+			Secondly we will study different retention mechanism and comment on their importance.
+			\vspace{1cm}
+		\end{abstract}
   	\end{@twocolumnfalse}

### 2023-07-24 main_text.tex
first draft hand out to philipp and jana

@@
 \section*{Introduction}
+%------------------------------------------------------------------------------
+%------------------------------------------------------------------------------
 
-\subsection*{Estuaries and Phytoplankton}
+% \subsection*{Estuaries and Phytoplankton}
 
 \textbf{Estuaries are important} both for the climate system as well as for anthropogenic use.
-From a climate perspective they offer a high productivity relative to surface area which results in large amounts of direct carbon cycling as well as their important role as a source of nutrients and hatching grounds for marine ecosystem.
+From a climate perspective they are highly productivity relative to their surface area 
+which results in large amounts of direct carbon cycling 
+as well as their important role as a source of nutrients and hatching grounds for marine ecosystem. (cite)
+They are also strongly anthropogenically influenced through e.g.
+harbors, diking, dredging, fishing and pollution which can be seen as a stressor on the ecosystem but it also adds an anthropogenic dependency on the ecosystem.
 (cite)
-They are also strongly anthropogenically influenced through e.g.
-harbors, diking and fishing which can be seen as a stressor on the ecosystem but also adds am anthropogenic dependency on the well being of the ecosystem.
 
-However, estuaries are also complex and strongly dynamic ecosystem such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic changes to this day.
-This complexity arises from the topology and the resulting hydrodynmics at the fresh-salt water interface.
+However, estuaries are also complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic changes. (cite)
+This complexity arises in part from the topology and the resulting hydrodynamics at the fresh-salt water interface. (cite)
 
 As for most ecosystem - their dynamics are strongly controlled by the primary producers.
-However phytoplankton, the major aquatic primary producer drifts in the current as all planktonic organism.
-Due to the current in the estuary we expect plankton to be drifting downstream over time and ultimately be washed out into marine high salinity waters.
-Hence, we wonder how phytoplankton manages to survive as a populating in such an environment.
+Phytoplankton, the 	predominant aquatic primary producer is mostly drifting passively through the water.
+Hence, its trajectory is strongly controlled by the currents and tides.
+With the estuary having a net outwards flow we expect phytoplankton to be moving
+ downstream over time and ultimately to be washed out into marine high salinity waters.
+Therefore, we are wondering how phytoplankton manages to survive as a populating in such an environment.
 If we assume that the population is not exclusively maintained by upstream plankton that is washed into the estuary
- then there must be some sort of retention mechanism that enables a phytoplankton population to not be washed out.
-While this is expected to be an crucial mechanism for survival  it is currently not well understood 
+ then there must be some sort of retention mechanism that enables a phytoplankton population to remain in the low-salinity area of the estuary.
+While we suspect that such a mechanism is crucial for survival it is currently not well understood 
 due to the difficulty in either measuring or modeling such processes.
 
-\subsection*{what is known}
+\subsection*{Current state of knowledge}
 
-We categorize the already performed studies on the subject of phytoplankton retention in three groups: observations, theoretical- and realistic-modelling studies.
+We categorize the already performed studies on the subject of phytoplankton retention in three groups: observations, theoretical- and applied-modelling.
 
-\textbf{Observational} studies have found three major mechanism for retention - diel migration, sinking and stickiness.
-Diel migration (moving verticaly up and down with the sun) has been found but so far only shown for larger estuarine organisms which are not strictly planktonic i.e. zooplankton [cite]
-Also benthic plankton has been observed to be strongly negatively buoyant, hence sinking and sticking to the ground.
+\textbf{Observational} studies have found three major mechanism for retention - sinking, diel migration, and stickiness. (cite)
+Diel migration (moving up and down with the sun) has been found but so far only shown for larger estuarine organisms which are not strictly planktonic i.e. zooplankton species [cite]
+Benthic plankton especially has been observed to be strongly negatively buoyant, hence sinking to the ground and remaining there for a long time (cite).
+Recent studies also found sticky compounds phytoplankton aggregates that might allow them to not only stick to themselves but to their surroundings aiding retention. (cite)
 
-Next want to highlight two more \textbf{theoretical} studies studying the mechanisms of "loss compensation" and "reseeding".
-(cite) theorized a set up in which they  a population is able to maintain itself if the losses due to outwashing can be compensated by growth (cite).
+Next we want to highlight two \textbf{theoretical modelling} studies studying the mechanisms of "loss compensation" and "reseeding".
+(cite) theorized a set up in which a population is able to maintain itself if the losses due to outwashing can be compensated by growth.
 In a setup where a lagoon like ecosystem was modelled they were able to quantify reproduction thresholds that are needed to maintain the population.
@@
 If the regrowing population is also continuously drifting downstream they will also ultimately die out.
-
 (cite) studied the concept of "reseeding".
 Here a system is theorized in which a population is able to repopulate an previously 
-uncolonized area upstream due to local upstream currents e.g. through larger eddies.
-While they showed that upstream reseeding is conceptually possible they did so in a special contrived system.
+lost or uncolonized area upstream due to local upstream currents e.g. through tides or dispersion.
+While they showed that upstream reseeding is conceptually possible they did so in a specifically contrived system resembling groyne fields .
 Whether this is an actual realized phenomenon depends on the topology and hydrodynamic of the estuary.
-
-\smallskip
-\texttt{More detail?}
-\smallskip
-
-Hence, it has been showed both "outgrowing the losses" and "reseeding" are possible retention mechanism but only under certain conditions.
+Nevertheless, it has been shown that both "outgrowing the losses" and "reseeding" are possible retention mechanism under certain conditions.
 
@@
 They were able to show  that sinking and diel migration slows the outwashing process and might be beneficial retention tactic.
-However, they did so by ignoring key processes like reproduction and death in an rectangular grid model
-which further introduces some numerical artefact.
+However, they did so by ignoring key processes like reproduction, mortality and any stranding or sedimentation processes in an low resolution structured grid model \texttt{specify low-res?}.
 
@@
 However, all approaches used major simplification which didn't allow them to study the full picture.
-We developed a new and more realistic approach to study retention with an particle tracking model combining
- many of the above outlined features into one model.
-
-\texttt{I don't have a proper "thats our hypothesis" part in my introduction which is considered to be bad style I think.
-        However, the way I frame it currently I don't really have a proper hypothesis. 
-        I approached it more as a kind of "lets see if we can find some noteworthy patterns and report them"
-        Personally I think this kind of playing around is worth communicating as well but I am not sure how the community feels about that.}
+We developed a new and more realistic approach to study retention using a particle tracking model.
+This model combines the features outlined above, namely: reproduction, mortality, vertical migration, stranding, sedimentation, and reseeding.
+We examine if such a system can reproduce phytoplankton retention in the Elbe estuary with these mechanisms and if so under which conditions.
 
 \section*{Methods}
+%------------------------------------------------------------------------------
+%------------------------------------------------------------------------------
 
-\subsection*{general model description}
+\subsection*{Model description}
+
+\subsubsection*{Particle tracking}
 
@@
 
-Additionally, because we are tracking particles individually we are able to observer their individual tracks.
-This makes it much easier and intuitive to interpret the results and allows us to  include "individual based processes".
-
-\texttt{better representation of the particle behavior close to the edges of the estuary?}
+Additionally, because we are simulating individually particles we are able to observer their tracks.
+This makes it much easier and intuitive to interpret the results and allows us to include individual based properties and processes.
 
@@
     \includegraphics[width=\columnwidth]{Artboard 2.png}
-    \caption{Particle Trackle visualisation - particle tracks on grid}
+    \caption{Particle trackl visualisation}
 \end{figure}
 
-The hydrodynamics \textbf{data} is taken from a SCHISM model from a Johannes Pein (cite).
-It is an three dimensional unstructured  grid model representing the full Elbe estuary from the weir at Geesthacht close to helgoland including side-channels and the harbor area.
+\subsubsection*{Data}
+We use the hydrodynamics \textbf{data} of the most recent SCHISM model (cite Johannes Pein).
+It is an three dimensional unstructured  grid model representing the full Elbe estuary from the weir at Geesthacht close to Helgoland including side-channels and the harbor.
 % (Look for more details in pein paper that might be interesting)
 It provides a node-based mesh field containing a range of information.
-For our purpose the water velocities represented in a flow field is the most important one
-But also salinity and many other more technical information is used (See appendix \texttt{not yet added})
-We study the year 2012 with 1h temporal- and dynamically varying spacial resolution.
+Most importantly for our purpose is the water velocities.
+But also salinity, water level, dispersion and many other more technical properties are used (See appendix \texttt{not yet added})
+The year represented in that dataset is 2012 with 1h temporal- and dynamically varying spacial resolution with a median distance between nodes of $\sim60m$.
 
@@
 
+\subsubsection*{Particle processes}
+
 On top of the "blanc" particle we add a set of textbf{biological features} to achieve a more realistic behavior.
-The most important one is their ability to reproduce by creating copies of themselves 
-and their ability to die in certain conditions.
-We also add a set of movement related features. 
-This includes a vertical movement behavior. 
-We included constant vertical velocities i.e. sinking and rising and a time dependent one namely diel-migration.
-Their interaction with the boundaries of our model is represented by a simple sedimentation and resuspension model in the water and in a stranding model on land
+The most important one is their ability to reproduce by creating copies of themselves.
+We also remove them under certain biologically unfavorable conditions. Specifically when they are stranded on the shore, starved for light or float in high salinity waters for two long (see appendix).
+A set of vertical movement features is added to model migration behavior. 
+These include a constant vertical velocity i.e. sinking and rising and a diel-migration pattern where they move up and down depending on the phase of the sun.
+A large area of the estuary is very shallow with average water depths of below one meter. 
+We include a simple sedimentation and resuspension model to represent stranding by the tide and also sedimentation model for the water column where resupension is depending on a critical sheer velocity (see appendix).
 
+\texttt{I once drew some pictogramms visualizing these processes but I wasn't sure if they were particualrly helpful. What do you think?}
 % \begin{figure}
@@
 
-\textbf{downstream retention}
-The first and major experiment of this studies examines retention mechanisms under different scenarios.
-General idea is to release a population of plankton at the beginning of the estuary represented by the weir in Geesthacht 
+\subsubsection*{Downstream retention}
+The first and major experiment of this studies examines retention success under different scenarios.
+The general idea is to release a population of plankton at the upstream end of the estuary represented by the weir in Geesthacht
 and examine how the population distrubutes itself over the estuary and whether it is able to maintain its population size.
-With this approach we study the effect of several mechanism. These include:
+With this approach we study the effect of several mechanism in a sensitivity analysis. These include reproduction and migration:
 \textit{Reproduction} represented as a chance of duplicating a particle,
@@
  based on a critical-friction-velocity model.
-Particles are removed if they either reach high salinity waters implemented as a mortality chance above $10 PSU$,
-when they are stranded until they dry out which is represented by killing all particles which have not been rewettet
- in the last $7$ days, and when they are starved for light which we defined as removing all particles which have been
- in low-light areas longer then a total of $14$ days where presence in high-light areas is subtracted from the starvation budget.
-\texttt{Sentences to long to assume - good enough for now}
+Particles are removed if either of the following criteria is met: 
+When they reach high salinity waters implemented as a mortality chance above $10 PSU$,
+When they are stranded outside of the water and lie dry for more then $7$ consecutive days.
+When they are starved for light which is represented as a budget of $14$ days 
+where energy is lost in in low-light and gained in high light areas.
 The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes 
@@
 
-The observed particle batch representing our phytoplankton population is released at new years eve in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
-We evaluate their \textbf{retention success} primarily by measuring the amount of plankton in different regions in the estuary over time.
-If the population survives a full year and its population is over a certain threshold we consider them "successfully retained".
+The initial particle batch that is representing our phytoplankton population is released at the beginning of the year in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
+We evaluate their \textbf{retention success} by counting the amount of plankton in different regions in the estuary over time.
+If the population is growing at the end of the year we consider them "successfully retained".
+We further establish a metric where we measure the final population size relative to the size of the population after an initial spin up time of three months.
+\texttt{explain why?}
+% At this time most of the population has already been washed out and only a small fraction of the inital population ($\sim10\%$) remains.
 
-To count the particles we use "\textbf{statistical polygons}" that split the estuaries in along-shore sections. 
-See  \ref{fig:statistial polygons}.
+To count the particles we use "\textbf{statistical polygons}" that split the estuaries in along-shore sections (see fig  \ref{fig:statistial polygons}).
 We also log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
@@
 The logged observables are measured every 12 hours starting at midnight.
-
 \medskip
@@
 Here we don't allow for any reproduction and give all particles a neutral buoyancy.
+Hence they are drifting like dust similar to a passive tracer
 This is our reference for any retention strategy. 
@@
 Here, we use the same set-up as in the second scenario with the only exception of disabling reproduction for particles stranded on the shores of the estuaries.
-We call this scenario "ready-for-liftoff"
+We call this scenario "ready-for-liftoff" (\texttt{silly placeholder name})
 \label{txt:ready-for-liftoff}
 
-\textbf{upstream migration}
+\subsubsection*{upstream migration}
+
 In the second and minor experiment we test the hypothesis of "reseeding" - whether organisms can recolonize previously lost upstream habitats.
-
-\textbf{parameter overview}
-The tested parameters described above like reproduction and vertical movement can be found in table \ref{tab:downstream_retention_params}.
-\smallskip
-\texttt{The table doesn't exist yet.}
-\smallskip
+\texttt{This is currently not included anymore. Seemed to uninteresting and far off track. What do you think?}
 
 \section*{Results}
-
-\texttt{The results presented below are not yet final and not-visualized (hence the placeholder figures)
-        but true and tested in the current version of the model.
-        Hence, we assume they might chance in some details but stay true with the improved dispersion model.
-        \newline
-        I also intermingled the pure results with some interpretations. I think this is considered bad style.
-        I plan to disentangle this more once I have the actual results to discuss if the journal style requires it.}
+%------------------------------------------------------------------------------
+%------------------------------------------------------------------------------
 
@@
 
+\subsubsection*{Scenario: dust, mono- and diel migration}
+
 \begin{figure}
     \includegraphics[width=\columnwidth]{diel_success.png}
-    \caption[]{total particles counts for scenario "migration" with a monotonic migration pattern for all tested parameters}
-    \label{fig:monotonic_particle_counts}
+    \caption[]{relative population changes indication retention success for a set of scenarios with monotonic vertical migration}
+    \label{fig:monotonic_retention_success}
 \end{figure}
-% \begin{figure}
-%     \includegraphics[width=\columnwidth]{example-image-a}
-%     \caption[]{total particles counts for scenario "migration" with a diel migration pattern for all tested parameters}
-%     \label{fig:diel_particle_counts}
-% \end{figure}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{diel.png}
+    \caption[]{relative population changes indication retention success for a set of scenarios with diel vertical migration}
+    \label{fig:diel_retention_success}
+\end{figure}
 
-In fig \ref{fig:dust_particle_counts} the particle counts measured over time it the estuary are shown.
-We can clearly see that the total particle count drops to zero - in most cases already after less then two months.
-From this we deduce that particles are constantly getting washed out and do not find "sheltered areas" 
-in which they can remain for a long time.
-This also means that reproduction is a necessary condition for the population to maintain their size 
-even if all mortality sources except high-salinity are ignored.
+The results of the retention experiments are visualized as heatmaps in fig. \ref{fig:monotonic_retention_success} and \ref{fig:diel_retention_success}.
+Fig. \ref{fig:monotonic_retention_success} shows the results for the monotonic vertical migration scenarios e.g. constant sinking or rising.
+Fig.
+\ref{fig:diel_retention_success} shows the results for the diel vertical migration scenarios.
+For the monotonic case a positive verticle velocity indicates an upwards motion.
+For the the diel case a positive vertical velocity means that the particles are moving upwards during the day and downwards during the night.
+Each pixel in the heatmap represents a scenario with a specific vertical velocity and a specific reproduction rate.
+The coloring indicates the relative population change after one year.
+A positive (green) value is consider a successful retention while a negative (red) value indicates that the population is going to die out.
 
-In fig \ref{fig:monotonic_particle_counts} and \ref{fig:diel_particle_counts} we can see 
-that the population manages to maintain themselves in certain conditions.
-The fact that we found conditions for survival in this simplistic setup is in itself interesting for two major reasons.
-First, all previsous studies implementing a phytoplankton in an Elbe model required constant reseeding either up and/or downstream.
-Hence, we are the first study that is able to provide an estimate for the conditions necessary for "in-estuar"-survival.
-\texttt{ The secondly might be better suited for the discussion.}
-Secondly, while we assume ad libitum nutrient conditions for the phytoplankton we do not implement any sub-grid-resolution structure on the shores.
-Hence, the river banks are perfectly flat sand banks without vegetation which might aid the retention of phytoplankton.
+The first thing that we wish to highlight is that the population successfully retains itself in certain scenarios.
+For the case of the monotonic migration we see a trend that the populations with higher positive velocities and higher growth rates are more successful in retaining themselves.
+For the case of the diel migration we see a trend to higher vertical velocities both upwards and downwards.
 
-\texttt{Add comparison of monotonic upwards to downwards, monotonic to diel, and the discuss the relative growth rate thresholds}
+Another thing that we would like to highlight is that populations with no reproduction are all unsuccessful in retaining themselves.
 
-Taking a closer look at the subset of pytoplankton that was able to retain themselves compared to those getting washed out.
-You can see these measurement visualized in the boxplots \ref{fig:migration-long-vs-short}
+We would also like to note that a net-zero-growth rates represent the cases in which the growth is equal to the mortality.
+Hence, the white pixels or the boundary between green and brown can be interpreted as an estimate for the total relative losses due to outwashing, dry-out when stranded, and light starvation.
+
+\subsubsection*{Scenario: ready-for-liftoff}
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{death.png}
+    \caption[]{relative population changes indication retention
+    success for a set of scenarios with diel vertical migration. \texttt{This one feels a bit silly to show. Its in here for the time being for consistency}}
+    \label{fig:stranded-monotonic-retention-success}
+\end{figure}
+
+In fig \ref{fig:stranded-monotonic-retention-success} we show the retention success for the monotonic case with reproduction disabled for non-moving particles - e.g. stranded or sedimented particles.
+
+Here we would like to highlight the difference in overall success compared to the previous scenarios.
+None of the tested parameters was able to successfully retain the population.
+
+\subsubsection*{Where do they survive?}
 
@@
 
-We can consistently see that the long living subset is in shallower water, does move less, and is a significantly longer stranded then drifting.
-In fig. \ref{fig:migration-long-vs-short-heatmap} we show the special distribution of the long- and short-living subset
- also indication their predominance of long living phytoplankton close to the shores and in side channels.
+Taking a closer look at the subset of pytoplankton that was able to retain themselves compared to those getting washed out.
+In fig.
+\ref{fig:migration-long-vs-short} we see a box plot comparing the average water depth at particle location for the long living and short living subset.
+The long living being older then three months.
+Depths is measured relative to the water surface. Hence, a values above zero indicate that the particle is stranded on the shore.
+We can clearly see that the long living subset predominantly lives in shallower water.
 
@@
 
-\texttt{show zoom in with water depth?}
-
-\subsubsection*{Scenario: "ready-for-liftoff"}
-
-\begin{figure}
-    \includegraphics[width=\columnwidth]{diel_success.png}
-    \caption[]{total particles counts for scenario "ready-for-liftoff" with a monotonic migration pattern for all tested parameters}
-    \label{fig:stranded-monotonic-particle-counts}
-\end{figure}
-
-In fig \ref{fig:stranded-monotonic-particle-counts} we again show the total particle counts.
-Keep in mind that the only difference in the model configuration to the "migration" run is that phytoplankton growth has been disabled when the plankton is stranded.
-We can clearly see that phytoplankton dies out in these conditions for the full parameter space.
-From this we deduce that the shallow areas are not only crucial for improved retention but also a necessary hatching ground.
-
-To summarize the retention experiments: We saw that the population is ultimately washed out and perishes if we suppress reproduction.
-We also saw that reproducing phytoplankton is able to maintain their population strain without up- or downstream reseeding under certain conditions!
-We further showed that the shallow water areas are not only helpful for retention but crucial for the survival of the population.
+This is further supported by fig. \ref{fig:migration-long-vs-short-heatmap}.
+Here we see a hex-bin-plot of the estuary. The coloring shows the average age of the particles in the bin.
+Bins with an average age of over three months are shown in a yellowish color.
+These yellow areas can exclusively be found along the shores in shallow areas.
+The major areas being Mühlenberger Loch (a), Wedeler Geest (b) and the Hasseldorfer Binnenelbe (c).
+We can also find minor areas at the river mouths of the Wischhafener Süderelbe (d), Ruthenstrom, and Stör (e).
+\texttt{labeling not yet added}
+\texttt{add a show zoom in with water depth?}
 
@@
 
-% \highlight{I dropped this part for the time being.}
-\begin{figure}
-    \includegraphics[width=\columnwidth]{example-image-a}
-    \caption[]{particle counts in upstream statistical polygons}
-    \label{fig:upstream-migration-statistical-poly-counts}
-\end{figure}
+\texttt{exclude this part for the time being for not being intersting enough.}
+% \begin{figure}
+%     \includegraphics[width=\columnwidth]{example-image-a}
+%     \caption[]{particle counts in upstream statistical polygons}
+%     \label{fig:upstream-migration-statistical-poly-counts}
+% \end{figure}
 
-In the upstream migration experiment we divide the estuary in short ($\sim 50km$) sections.
-We release particles in all of these sections and count their occurrence in other sections.
-As previously described - in fig \ref{fig:upstream-migration-statistical-poly-counts} we see the 
-the particle counts for the two upstream sections from the release location.
-We consider upstream migration to happen when plankton moves two sections up and is able to retain itself there.
-\texttt{Do I need to explain why two polygons and why being there only briefly won't work?}
+% In the upstream migration experiment we divide the estuary in short ($\sim 50km$) sections.
+% We release particles in all of these sections and count their occurrence in other sections.
+% As previously described - in fig \ref{fig:upstream-migration-statistical-poly-counts} we see the 
+% the particle counts for the two upstream sections from the release location.
+% We consider upstream migration to happen when plankton moves two sections up and is able to retain itself there.
+% \texttt{Do I need to explain why two polygons and why being there only briefly won't work?}
 
-\texttt{No results yet with the proper dispersion model. 
-        In the previous run upstream migration did not occur and was considered 
-        to be a negligible mechanism in the retention process.
-        However, this might chance with the enabling of tidal pumping mechanisms.}
+% \texttt{No results yet with the proper dispersion model. 
+%         In the previous run upstream migration did not occur and was considered 
+%         to be a negligible mechanism in the retention process.
+%         However, this might chance with the enabling of tidal pumping mechanisms.}
 
 \section*{Discussion}
+%------------------------------------------------------------------------------
+%------------------------------------------------------------------------------
 
@@
 
-\begin{itemize}
-    \item retention success threshold as a measure of estimating outwashing losses
-    \item very low outwashing rates imply unfavorable conditions for phytoplankton? or maybe the conditions are good and they grow nicely but not enough to compensate for losses in harbor?
-    \item 
-\end{itemize}
+\subsection*{Interpretation of the results}
+The range of growth rates that we found to be necessary for survival are small compared to those typically realized in nature where plankton populations can double their size in a day in ideal conditions.
+Our model is simplistic in the sense that we do not include any nutrient limitation which are assumed to drastically decrease the growth rates compared to ideal conditions. 
+Nevertheless, the growth rates that we found to be necessary are still much small compared to those typically realized in nature (cite).
+
+The importance of shallow near-shore areas for retention has been suggested several times by our findings. 
+We assume that this is for the following reasons.
+Plankton that lives in this area falls dry on the shore on a regularly due to the tides basis. Hence, they are not moving downstream for a large portion of the day.
+We further see that positively buoyent plankton is more successful in retaining themselves. 
+This is likely due to the fact that they are more likely be washed up high up at shore where the water reaches them less frequently.
+This effect is emphasized in flatter regions where distances between the driftline and waters that don't fall dry is large, increasing the distance that the plankton has to travel without settling to be washed out.
+We assume that this effect would be further emphasized if we were to represent stokes drifts in our model even though wind forcing on the surface water is included.
+
+Initially we expected that negatively buoyant plankton would be more successful in retaining themselves.
+Sinking particles have a reduced downstream velocity as they tend to get stuck on the ground more often.
+Additionally, the bottom water layer in the Elbe typically has higher upstream velocity comparts to the upper water column.
+However, they still seem to be less successful then positively buoyant particles.
+Examining our model we think  that this is because they tend to quickly aggregate in deep waters with little to no light where they are unable to sustain themselves and ultimately die out.
+This is hinting at the negative impact of dreading on sinking plankton as it both increases the volume of deep water and increases turbidity reducing the volume of illuminated water.
+
+It is interesting to see that in case of diel migration success does not seem to depend on weather they are moving up during the day or the night.
+We assume that this is to a large part due to them only surviving in shallow near-shore waters where light limitation does not play a role.
+We think the reason for higher success with higher vertical velocities is similar to the constant vertical velocity case.
+When the upwards diel migration aligns with a high tide the particles are more likely to be stranded far out onto the shore decreasing their risk to be washed out quickly.
+The higher the upwards velocities the higher the chance to be in the outmost part of the water body during high tide
+However, we also want to note that our light limitation model is very simplistic to keep the model complexity low.
+For plankton to be not light limited they only need to be above a depth of one meter in the high turbidity water of the Elbe unregarded of current illumination conditions and local turbidity.
+We therefore think that a more realistic light limitation model might introduce a positive bias towards those particles that move upwards during the day.
+
+\texttt{What else?}
 
@@
 
-\texttt{The discussion is currently only a list of bullet points that I considered to discuss so far. It is by no means complete or currently not worth reading.}
-\begin{itemize}
-    \item discuss limitation of the model
-    \begin{itemize}
-        \item simplified bottom/sediment interaction
-        \item no "small object" representation e.g. vegetation or "rocks" $\rightarrow$ implices underestimation of "retention"
-        \item simplistic bio representation. does not allow for precise quantification but only indicates bounds aka "even in ideal conditions they couldn't survive if ..."
-        \item other minor issues - temporal resolution, particle out-of-bounds cases etc
-        \item discuss choice of particle tracking model compared to concentration based model
-    \end{itemize}
-    \item conclusions that we draw from this:
-    \begin{itemize}
-        \item from a population-surival perspective the plankton that does not drift to shallow water are "dead man walking" as all decendents of these particles will ultimately die.
-        \item we need cross-shore sampling to better understand this. currently all specially resolved sampling strategies are along shor - aka in the center of the river
-        \item shallow water is crucial for the survival. hence the protection and restoration of shallow regions is crucial for ecosystem healthiness
-    \end{itemize}
-\end{itemize}
+One major limitation of our model is the lack of nutrient limitation or the lack of a realistic ecosystem model in general.
+This was done to keep both developing and computational time low for an initial proof of concept as coupling a lagrangian model online to an eularian model would have drastically increased the time complexity to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
+The key draw back of this is that growth rates could only be modeled as a constant rate assuming an "ad libitum" environment at all times.
+This introduces systematic errors in the population growth as areas which are beneficial for retention might in reality quickly run out of nutrients and could not sustain the growth conditions necessary to sustain the local population.
+
+Another major limitation is the lack of any sub-grid-resolution structure on the shores - especially vegetation.
+In our representation we assume perfectly flat surfaces with an median distance between nodes of approximately 60m.
+This "polished" model representation is assumed to cause a strong negative bias towards retention success as some plankton could easily stick to the vegetation or other cavities caused by the sub-grid-resolution structure and reseed the area.
+
+While our model does have a sedimentation and resuspension mechanic based on critical sheer velocities we still assume a static bathymetrie with sediments not able to move or bury particles. 
+This maskes potential losses due to particles being buried but also decreases resuspension times.
+
+\subsection*{Outlook}
+
+Our results clearly suggested the importance of shallow near-shore areas for the survival of phytoplankton in the Elbe estuary.
+After our initial experiments suggested this, we considered improving our model to resolve additional ecosystem interactions especially nutrient limitation.
+This would allow us to predict the population dynamics more realistically to the point where we would be able to quantize importance of shallow areas more accurately.
+However, the main limitation to do so was the lack of validation data for these regions. 
+Almost all chlorophyll data is gather in the center of the river.
+While we consider the general trend shown by our model to be true - shallow areas are essential for survival.
+We can not quantify the effect of changes in these areas that might be caused by future dredging, diking or restoration attempt.
+To do this, we first require more specially resolved data for these areas.
+
+We were also surprised by our finding that sinking particles have a harder timer surviving in the estuary. 
+Since several decades the yearly average chlorophyll concentration in the Elbe estuary concentrations declined (cite) while upstream concentrations do not show this effect.
+The reasons for this are not yet fully understood but one potential reason is the increase in dreading activities.
+These increases the average turbidity reducing the water volume where phytoplankton is able to grow.
+A large fraction of the upstream phytoplankton consists of diatoms (cite).
+Diatoms typically have negative buoyancies (cite) making them especially vulnerable to sinking to light limited waters.
+
+Another mechanism that might in part explain this effect is but is not yet explored in our model is the phytoplankton stickiness (cite). 
+Phytoplankton especially blooming one has been shown to be sticky due to exudates (cite Alenka Male 1993).
+In combination with increased turbidity this might increases losses due to plankton aggregates sticking to negatively buoyant suspended matter and subsequently sinking to the ground where they are starved of light.
+We are curious to which extend these mechanisms might explain the phytoplankton losses in the dreadged areas of the estuary and suggest both a field and modling study.
+A field study could provide estimates on suspended matter concentration dependant aggregation speeds of phytoplankton and their sinking velocities.
+While the model study might create estimates on chlorophyll concentration losses caused by this effect.
+
+\section*{Conclusion}
+
+We saw that reproducing phytoplankton is able to maintain their population strain without up- or downstream reseeding under certain conditions.
+We were also able to give a first estimate on the growth rates necessary for survival of the population depending on their buoyancy and their respective outwashing losses.
+We also saw that the population is ultimately washed out and perishes if we suppress reproduction.
+The same is true if we do not suppress reproduction completely but only for particles currently not floating in the water body indicating the importance of the shores for retention.
+To investigate this further we looked at the average water depth of long living particles and the areas where they are found which further indicated the importance of shallow areas for retention.
+
+We suggest further sampling to measure chlorophyll concentrations in shallow waters.
+Specifically at Mühlenberger Loch and Hasseldorfer Binnenelbe to quantify the importance of these areas on population health.

### 2023-07-24 ms.tex
first draft hand out to philipp and jana

@@
 	\vspace{-20mm}\textbf{	\large 
-	Plankton Retention mechanism studies at Elbe with IBM}
+	\texttt{Placeholder Title:}  Retention}
 	}
@@
 			\begin{center}
-				\texttt{All texts in this typewriter fond are drafting comments and thoughts on the text itself and not part of the actual paper.}
-				\smallskip
-			\end{center}
-
-			\begin{center}
-				\texttt{placeholder abstract - not to be proofread}
+				\texttt{All texts in this typewriter font are drafting comments.}
 				\smallskip
@@
 			% \end{figure}
-			Estuaries are a challenging environment for plankton.
+			\texttt{placeholder abstract: }Estuaries are a challenging environment for plankton.
 			Gradients are steep and currents are strong.
-			Retention mechanisms are crucial for their survival.
+			This make retention mechanisms are crucial for their survival.
 			However, very little experimental and theoretical evidence has been gathered on the mechanism of phytoplankton retention.

### 2023-07-24 main_text.tex
added philipps comments

@@
 
-\textbf{Estuaries are important} both for the climate system as well as for anthropogenic use.
-From a climate perspective they are highly productivity relative to their surface area 
+\texttt{Add a short paragraph as an outlook/summary of the paper.}
+
+Estuaries are important both for the climate system as well as for anthropogenic use.
+From a climate perspective they are highly productiv relative to their surface area 
 which results in large amounts of direct carbon cycling 
-as well as their important role as a source of nutrients and hatching grounds for marine ecosystem. (cite)
+along with their important role as a source of nutrients and hatching grounds for marine ecosystem. (cite)
 They are also strongly anthropogenically influenced through e.g.
@@
 
-However, estuaries are also complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic changes. (cite)
-This complexity arises in part from the topology and the resulting hydrodynamics at the fresh-salt water interface. (cite)
+Estuaries are also complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic changes. (cite)
+This complexity arises in part from the bathymetrie and the resulting hydrodynamics at the fresh-salt water interface. (cite)
 
@@
  downstream over time and ultimately to be washed out into marine high salinity waters.
-Therefore, we are wondering how phytoplankton manages to survive as a populating in such an environment.
+Therefore, we are wondering how phytoplankton manages to survive as a population in such an environment.
 If we assume that the population is not exclusively maintained by upstream plankton that is washed into the estuary
@@
 
-\subsection*{Current state of knowledge}
+% \subsection*{Current state of knowledge}
 
@@
 
-\textbf{Observational} studies have found three major mechanism for retention - sinking, diel migration, and stickiness. (cite)
-Diel migration (moving up and down with the sun) has been found but so far only shown for larger estuarine organisms which are not strictly planktonic i.e. zooplankton species [cite]
-Benthic plankton especially has been observed to be strongly negatively buoyant, hence sinking to the ground and remaining there for a long time (cite).
-Recent studies also found sticky compounds phytoplankton aggregates that might allow them to not only stick to themselves but to their surroundings aiding retention. (cite)
+Observational studies have found three major mechanism for retention - sinking, diel migration, and stickiness. (cite)
+Diel migration - moving up and down with the sun -  has been found to be favorable for retention by allowing organisms to stay close to bottoms slow moving waters.
+However, so far this has only shown for larger motile estuarine organisms i.e. zooplankton species [cite].
+Benthic phytoplankton especially has been observed to be strongly negatively buoyant, hence sinking to the ground and remaining there for a long time (cite).
+Recent studies also found sticky compounds in phytoplankton aggregates that are suspected to let them stick to themselves and their surroundings aiding retention. (cite)
 
-Next we want to highlight two \textbf{theoretical modelling} studies studying the mechanisms of "loss compensation" and "reseeding".
+Next we want to highlight several modelling study that examined whether certain retention mechanism could work in theory.
+Two theoretical modelling studies studying the mechanisms of "loss compensation" and "reseeding".
+\texttt{I don't like the term theoretical modelling study. Maybe "conceptual", "abstract" or "simple"?}
 (cite) theorized a set up in which a population is able to maintain itself if the losses due to outwashing can be compensated by growth.
@@
 Particle tracking offers several advantages for our question:
-While particle tracking on unstructured grids has been relatively computationally expensive until recently (cite)
+While particle tracking on unstructured grids has been relatively computationally expensive until recently (cite).
 it allows us to reuse already validated hydrodynamic models.
-Reusing already computed and calibrated hydrodynamical models is overall much faster then solving the advection-diffusion again.
+Reusing already computed hydrodynamical models to study tracer like objects with an Lagrangian model  is overall much faster then recalculating the advection-diffusion-equation in an Eularian model.
 
-Additionally, because we are simulating individually particles we are able to observer their tracks.
+Additionally, because we are simulating individually particles we are able to observe their tracks.
 This makes it much easier and intuitive to interpret the results and allows us to include individual based properties and processes.
@@
 It provides a node-based mesh field containing a range of information.
-Most importantly for our purpose is the water velocities.
+Most important for our purpose is the water velocity.
 But also salinity, water level, dispersion and many other more technical properties are used (See appendix \texttt{not yet added})
@@
 
-On top of the "blanc" particle we add a set of textbf{biological features} to achieve a more realistic behavior.
+On top of the "blank" particle we add a set of textbf{biological features} to achieve a more realistic behavior.
 The most important one is their ability to reproduce by creating copies of themselves.
-We also remove them under certain biologically unfavorable conditions. Specifically when they are stranded on the shore, starved for light or float in high salinity waters for two long (see appendix).
+We also remove them under certain biologically unfavorable conditions. Specifically when they are stranded on the shore, starved for light or float in high salinity waters for too long (see appendix).
 A set of vertical movement features is added to model migration behavior. 
@@
 We include a simple sedimentation and resuspension model to represent stranding by the tide and also sedimentation model for the water column where resupension is depending on a critical sheer velocity (see appendix).
-
-\texttt{I once drew some pictogramms visualizing these processes but I wasn't sure if they were particualrly helpful. What do you think?}
-% \begin{figure}
-%     \centering
-%     \begin{minipage}{.4\columnwidth}
-%         \centering
-%         \includegraphics[width=.3\columnwidth]{picto resuspension.pdf}
-%     \end{minipage}
-%     \begin{minipage}{.4\columnwidth}
-%     \centering
-%         \includegraphics[width=.3\columnwidth]{picto splitting.pdf}
-%     \end{minipage}
-%     \caption{examples of pictograms for bio processes}
-% \end{figure}
-
-% \begin{figure}
-%     \includegraphics[width=\columnwidth]{fixed_particle_culling.pdf}
-%     \caption{Example 2d plot of particle counts, centroid and "retainment" threshold}
-% \end{figure}
 
@@
 
-\subsubsection*{Downstream retention}
-The first and major experiment of this studies examines retention success under different scenarios.
+The first experiment of this study examines retention success under different scenarios.
 The general idea is to release a population of plankton at the upstream end of the estuary represented by the weir in Geesthacht
@@
 Particles are removed if either of the following criteria is met: 
-When they reach high salinity waters implemented as a mortality chance above $10 PSU$,
+When they reach high salinity waters implemented as a mortality chance in fixed intervals above $10 PSU$ (see appendix).
 When they are stranded outside of the water and lie dry for more then $7$ consecutive days.
 When they are starved for light which is represented as a budget of $14$ days 
-where energy is lost in in low-light and gained in high light areas.
+where energy is lost in a fixed rate in low-light and gained in high light areas.
 The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes 
@@
 
-\smallskip
-\texttt{More detail needed? - or should that go into the appendix?}
-\smallskip
+The initial particle batch that is representing our phytoplankton population is released at the beginning of the year in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
+We evaluate their retention success by counting the amount of plankton in different regions in the estuary over time.
+If the population is growing at the end of the year we consider them "successfully retained".
+Population changes are measured at the end of the year relative to the population size after after an initial spin up time of three months where most of the initial batch has been washed out and the population typically is at its minimum.
 
-The initial particle batch that is representing our phytoplankton population is released at the beginning of the year in a volume at the weir in Geesthacht which marks the beginning of the Estuary.
-We evaluate their \textbf{retention success} by counting the amount of plankton in different regions in the estuary over time.
-If the population is growing at the end of the year we consider them "successfully retained".
-We further establish a metric where we measure the final population size relative to the size of the population after an initial spin up time of three months.
-\texttt{explain why?}
-% At this time most of the population has already been washed out and only a small fraction of the inital population ($\sim10\%$) remains.
-
-To count the particles we use "\textbf{statistical polygons}" that split the estuaries in along-shore sections (see fig  \ref{fig:statistial polygons}).
-We also log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
+For each particle we log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
 This allows us to compare the successful (died older then three months) and unsuccessful (died younger then then three months ) plankton.
 The logged observables are measured every 12 hours starting at midnight.
+
 \medskip
@@
 \ref{fig:diel_retention_success} shows the results for the diel vertical migration scenarios.
-For the monotonic case a positive verticle velocity indicates an upwards motion.
+For the monotonic case a positive vertical velocity indicates an upwards motion.
 For the the diel case a positive vertical velocity means that the particles are moving upwards during the day and downwards during the night.
@@
 
-\subsubsection*{Scenario: ready-for-liftoff}
+\medskip
 
-\begin{figure}
-    \includegraphics[width=\columnwidth]{death.png}
-    \caption[]{relative population changes indication retention
-    success for a set of scenarios with diel vertical migration. \texttt{This one feels a bit silly to show. Its in here for the time being for consistency}}
-    \label{fig:stranded-monotonic-retention-success}
-\end{figure}
-
-In fig \ref{fig:stranded-monotonic-retention-success} we show the retention success for the monotonic case with reproduction disabled for non-moving particles - e.g. stranded or sedimented particles.
-
-Here we would like to highlight the difference in overall success compared to the previous scenarios.
-None of the tested parameters was able to successfully retain the population.
+In the scenario "ready-for-liftoff" we examined the retention success for the monotonic case with reproduction disabled for non-moving particles - e.g. stranded or sedimented particles.
+For this scenario none of the tested parameters was able to successfully retain the population.
 
@@
 
-Taking a closer look at the subset of pytoplankton that was able to retain themselves compared to those getting washed out.
+We are now taking a closer look at the subset of pytoplankton that was able to retain themselves.
 In fig.
@@
 The long living being older then three months.
-Depths is measured relative to the water surface. Hence, a values above zero indicate that the particle is stranded on the shore.
+Depth is measured relative to the water surface. Hence, a value above zero indicate that the particle is stranded on the shore.
 We can clearly see that the long living subset predominantly lives in shallower water.
@@
 
-\textbf{upstream migration}
+% \textbf{upstream migration}
 
-\texttt{exclude this part for the time being for not being intersting enough.}
 % \begin{figure}
@@
 Our model is simplistic in the sense that we do not include any nutrient limitation which are assumed to drastically decrease the growth rates compared to ideal conditions. 
-Nevertheless, the growth rates that we found to be necessary are still much small compared to those typically realized in nature (cite).
+Nevertheless, the growth rates that we found to be necessary are still much smaller compared to those typically realized in nature (cite).
 
@@
 Examining our model we think  that this is because they tend to quickly aggregate in deep waters with little to no light where they are unable to sustain themselves and ultimately die out.
-This is hinting at the negative impact of dreading on sinking plankton as it both increases the volume of deep water and increases turbidity reducing the volume of illuminated water.
+This is hinting at the negative impact of dredging on sinking plankton as it both increases the volume of deep water and increases turbidity reducing the volume of illuminated water.
 
-It is interesting to see that in case of diel migration success does not seem to depend on weather they are moving up during the day or the night.
+It is interesting to see that in case of diel migration success does not seem to depend on whether they are moving up during the day or the night.
 We assume that this is to a large part due to them only surviving in shallow near-shore waters where light limitation does not play a role.
@@
 
-\texttt{What else?}
-
 \subsection*{Limitations of the model}
@@
 One major limitation of our model is the lack of nutrient limitation or the lack of a realistic ecosystem model in general.
-This was done to keep both developing and computational time low for an initial proof of concept as coupling a lagrangian model online to an eularian model would have drastically increased the time complexity to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
+This could have been done by coupling our Lagrangian model online to an Eularian model that would have advected changes in concentrations fields by biotic activity throughout the model domain
+However, this would have drastically increased both developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
 The key draw back of this is that growth rates could only be modeled as a constant rate assuming an "ad libitum" environment at all times.
@@
 Phytoplankton especially blooming one has been shown to be sticky due to exudates (cite Alenka Male 1993).
-In combination with increased turbidity this might increases losses due to plankton aggregates sticking to negatively buoyant suspended matter and subsequently sinking to the ground where they are starved of light.
+In combination with increased turbidity this losses due to plankton aggregates sticking to negatively buoyant suspended matter and subsequently sinking to the ground where they are starved of light.
 We are curious to which extend these mechanisms might explain the phytoplankton losses in the dreadged areas of the estuary and suggest both a field and modling study.
@@
 
-We suggest further sampling to measure chlorophyll concentrations in shallow waters.
-Specifically at Mühlenberger Loch and Hasseldorfer Binnenelbe to quantify the importance of these areas on population health.
+We suggest further sampling to measure chlorophyll concentrations in shallow waters - specifically at Mühlenberger Loch and Hasseldorfer Binnenelbe - to quantify the importance of these areas on the general population health.

### 2023-08-01 main_text.tex
draft sent to ross

@@
 
-\texttt{Add a short paragraph as an outlook/summary of the paper.}
+\texttt{Add a short paragraph as an outlook of the paper? I was told thats what the cool kids do but I feel like thats already in the abstract.}
 
 Estuaries are important both for the climate system as well as for anthropogenic use.
-From a climate perspective they are highly productiv relative to their surface area 
+From a climate perspective they are highly productive relative to their surface area 
 which results in large amounts of direct carbon cycling 
@@
 
-In (cite) they use a \textbf{particle tracking model} also known as Lagragian model to study Zooplankton movement in the San Francisco Bay.
+In (cite) they use a particle tracking model also known as Lagragian model to study Zooplankton movement in the San Francisco Bay.
 They were able to show  that sinking and diel migration slows the outwashing process and might be beneficial retention tactic.
-However, they did so by ignoring key processes like reproduction, mortality and any stranding or sedimentation processes in an low resolution structured grid model \texttt{specify low-res?}.
+However, they did so by ignoring key processes like reproduction, mortality and any stranding or sedimentation processes in an low resolution structured grid model.
 
-To \textbf{summarize} - several concepts for potential retention mechanisms have been found
+We are examining the Elbe estuary as a case study.
+Like most alluvial estuaries it is relatively flat.
+Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centauries.
+Most notably diking to restrain it to a narrow channel and dreading to improve access to the Hamburg harbor.
+Unlike other major european harbors, the Hamburg harbor is located far inwards roughly 100km away from the coast.
+This causes the river to experience a sudden jump in bathymetry from 5m at border of the city to almost 20m in the harbor (see fig. \ref{fig:bathymetry}). 
+This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton biomass causing an increase in oxygen depletion and high ammonium reminaralization after the bathymetric jump.
+(cite)
+While major aspects of the along-channel biochemical dynamics have been studied little is known regarding their vertical and shore-to-shore dynamics.
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{harbor.png}
+    \caption{Bathymetry of the Elbe model around Hamburg.
+Note the bathymetric jumps from ~5m on the right, upstream side to a short 10m step in the upper harbor area to the ~20m in the lower harbor area all the way to the north sea.
+Also note that there exists no channel that does not pass through the 10km long exclusively 20m deep area.
+}
+    \label{fig:bathymetry}
+\end{figure}
+
+To summarize - several concepts for potential retention mechanisms have been found
  and shown to work in certain conditions.
@@
 
-In our study we take a particle tracking approach with a \textbf{lagrangian model} called Oceantracker (cite).
+In our study we take a particle tracking approach with a lagrangian model called Oceantracker (cite).
 Particle tracking offers several advantages for our question:
@@
 it allows us to reuse already validated hydrodynamic models.
-Reusing already computed hydrodynamical models to study tracer like objects with an Lagrangian model  is overall much faster then recalculating the advection-diffusion-equation in an Eularian model.
+Reusing already computed hydrodynamical models to study tracer like objects with an Lagrangian model  is overall much faster than recalculating the advection-diffusion-equation in an Eularian model.
 
@@
 
-\begin{figure}
-    \includegraphics[width=\columnwidth]{Artboard 2.png}
-    \caption{Particle trackl visualisation}
-\end{figure}
-
-\subsubsection*{Data}
-We use the hydrodynamics \textbf{data} of the most recent SCHISM model (cite Johannes Pein).
-It is an three dimensional unstructured  grid model representing the full Elbe estuary from the weir at Geesthacht close to Helgoland including side-channels and the harbor.
+\subsubsection*{Hydrological model}
+We use the hydrodynamics data of the most recent SCHISM model (cite Johannes Pein).
+It is an three dimensional unstructured  grid model representing the full Elbe estuary from the weir at Geesthacht close to Helgoland including side-channels and the harbor (see fig.
+\ref{fig:model_domain}).
 % (Look for more details in pein paper that might be interesting)
-It provides a node-based mesh field containing a range of information.
+The model provides a node-based mesh field containing a range of information.
 Most important for our purpose is the water velocity.
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{dfg_foto_wettbewerbg_laurin_legendless_steidle.png}
-    \caption{model domain visualization - full range grid including statistical polygons for later reference and example of a scalar field (e.g. temperature)}
-    \label{fig:statistial polygons}
+    \includegraphics[width=\columnwidth]{dfg_foto_wettbewerbg_laurin_legendless_steidle_small.png}
+    \caption{Map of the full model domain with Geesthacht being the upstream boarder on the right and the Northsea being the downstream edge on the top left.
+The gray outline indicates the edge of the model domain.
+Blue and green dots show an example snapshot of a phytoplankton distribution in the model.
+Blue being floating, green being stranded particles.}
+    \label{fig:model_domain}
 \end{figure}
 
-\subsubsection*{Particle processes}
+\subsubsection*{Particle based processes}
 
-On top of the "blank" particle we add a set of textbf{biological features} to achieve a more realistic behavior.
+On top of the "blank" particle we add a set of biological features to achieve a more realistic behavior.
 The most important one is their ability to reproduce by creating copies of themselves.
-We also remove them under certain biologically unfavorable conditions. Specifically when they are stranded on the shore, starved for light or float in high salinity waters for too long (see appendix).
+This is represented as a constant chance of duplicating a particle.
+We also remove them under certain biologically unfavorable conditions. Specifically when they are stranded on the shore, starved for light or float in high salinity waters for too long.
+More precisely, they are removed if either of the following criteria is met: 
+When they reach high salinity waters implemented as a mortality chance $1\%$ every minute.
+of  in fixed intervals above $10$PSU.
+When they are stranded outside of the water and lie dry for more then $7$ consecutive days.
+When they are starved for light which is represented as a budget of $14$ days 
+where energy is lost in a fixed rate in low-light and gained in high light areas.
+
 A set of vertical movement features is added to model migration behavior. 
-These include a constant vertical velocity i.e. sinking and rising and a diel-migration pattern where they move up and down depending on the phase of the sun.
+This is implemented in two modes. 
+One being a constant vertical velocity i.e. sinking and rising
+The second being diel-migration where they move up and down depending on the phase of the sun.
 A large area of the estuary is very shallow with average water depths of below one meter. 
@@
 
+Particles are stranded and temporarily stuck in place when they cross a model boundary or when their current cell falls dry. 
+Particle are resuspended from the bottom based on a critical sheer velocity model.
+If they are stranded they resuspend as soon as there location is flooded again.
+
+The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes and to allow for more realistic trajectories.
+
 \subsection*{Experimental configurations}
-
-% \begin{figure}
-%     \includegraphics[width=\columnwidth]{plot experimental configuration.pdf}
-%     \caption{2d grid with boxes for the different cases/dimension. 
-%              Same structure as the later sensitivity analysis plot used \ref{fig:retention_results_SA}}
-%     \label{fig:experimental_configuration}
-% \end{figure}
 
@@
 and examine how the population distrubutes itself over the estuary and whether it is able to maintain its population size.
-With this approach we study the effect of several mechanism in a sensitivity analysis. These include reproduction and migration:
-\textit{Reproduction} represented as a chance of duplicating a particle,
-\textit{vertical migration} where two modes have been implemented. 
-A constant up or downward drift and diel migration where the drift speed and direction is based on the phase of the sun.
-The particle \textit{stranding and resuspension} as well as \textit{suspend and resuspension to and from the bottom}
- based on a critical-friction-velocity model.
-Particles are removed if either of the following criteria is met: 
-When they reach high salinity waters implemented as a mortality chance in fixed intervals above $10 PSU$ (see appendix).
-When they are stranded outside of the water and lie dry for more then $7$ consecutive days.
-When they are starved for light which is represented as a budget of $14$ days 
-where energy is lost in a fixed rate in low-light and gained in high light areas.
-The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes 
-and to allow for a realistic trajectories close to boundaries.
+% With this approach we study the effect of two retention mechanism in a sensitivity analysis. These being reproduction and migration.
 
@@
 If the population is growing at the end of the year we consider them "successfully retained".
-Population changes are measured at the end of the year relative to the population size after after an initial spin up time of three months where most of the initial batch has been washed out and the population typically is at its minimum.
+Population changes are measured at the end of the year relative to the population size after after an initial spin up time of three months where most of the initial batch has been washed out and the population is typically at its minimum.
 
 For each particle we log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
-This allows us to compare the successful (died older then three months) and unsuccessful (died younger then then three months ) plankton.
+This allows us to compare the successful (died older then three months) and unsuccessful (died younger then three months ) plankton.
 The logged observables are measured every 12 hours starting at midnight.
@@
 
-We use this set-up in a range of scenarios and compare them to each other.
+We use this set-up in a range of scenarios.
 We call the first and simplest scenario "dust". 
-Here we don't allow for any reproduction and give all particles a neutral buoyancy.
+Here we don not allow for any reproduction and give all particles a neutral buoyancy.
 Hence they are drifting like dust similar to a passive tracer
 This is our reference for any retention strategy. 
-For a retention strategy to be considered beneficial it must perform better then these passive tracers.
+For a retention strategy to be considered beneficial it must perform better than these passive tracers.
 
-In the second scenario referred to as "migration" reproduction is re-enabled with a range of different growth rates 
-and a range of different buoyancies i.e. vertical velocities is given to the plankton.
+In the second scenario referred to as "migration" reproduction is re-enabled.
+In this scenario we study the effect of different growth rates and buoyancies i.e. vertical velocities on the phytoplankton population in a sensitivity analysis.
 
@@
 Here, we use the same set-up as in the second scenario with the only exception of disabling reproduction for particles stranded on the shores of the estuaries.
-We call this scenario "ready-for-liftoff" (\texttt{silly placeholder name})
-\label{txt:ready-for-liftoff}
 
-\subsubsection*{upstream migration}
+% \subsubsection*{upstream migration}
 
-In the second and minor experiment we test the hypothesis of "reseeding" - whether organisms can recolonize previously lost upstream habitats.
-\texttt{This is currently not included anymore. Seemed to uninteresting and far off track. What do you think?}
+% In the second and minor experiment we test the hypothesis of "reseeding" - whether organisms can recolonize previously lost upstream habitats.
+% \texttt{This is currently not included anymore. Seemed to uninteresting and far off track. What do you think?}
 
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{diel_success.png}
-    \caption[]{relative population changes indication retention success for a set of scenarios with monotonic vertical migration}
+    \includegraphics[width=\columnwidth]{retention_success_sa_mono.png}
+    \caption[]{Relative population changes for the monotonic migration scenarios.
+Positive vertical velocities indicate an upwards drift direction.
+Positive population changes indicate a retention success (shown in green) while negative population changes predict a long term loss of the population (shown in brown).}
     \label{fig:monotonic_retention_success}
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{diel.png}
-    \caption[]{relative population changes indication retention success for a set of scenarios with diel vertical migration}
+    \includegraphics[width=\columnwidth]{retention_success_sa_diel.png}
+    \caption[]{Relative population changes for the diel migration scenarios.
+Positive vertical velocities indicate an upwards drift direction.
+Positive population changes indicate a retention success (shown in green) while negative population changes predict a long term loss of the population (shown in brown).}
     \label{fig:diel_retention_success}
@@
 The coloring indicates the relative population change after one year.
-A positive (green) value is consider a successful retention while a negative (red) value indicates that the population is going to die out.
+A positive (green) value is consider a successful retention while a negative (brown) value indicates that the population is going to die out.
 
@@
 
-We would also like to note that a net-zero-growth rates represent the cases in which the growth is equal to the mortality.
+We would also like to note that a net-zero-growth rates represent the cases in which the growth is equal to the net losses.
 Hence, the white pixels or the boundary between green and brown can be interpreted as an estimate for the total relative losses due to outwashing, dry-out when stranded, and light starvation.
@@
 
-In the scenario "ready-for-liftoff" we examined the retention success for the monotonic case with reproduction disabled for non-moving particles - e.g. stranded or sedimented particles.
-For this scenario none of the tested parameters was able to successfully retain the population.
+In the scenario where we disabled reproduction for stranded particles we found that the population was unable to retain itself in all tested parameterizations.
 
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{dbf_bar.png}
-    \caption[]{Boxplot of long-vs-short showing water depth}
+    \includegraphics[width=\columnwidth]{retention_boxplot.png}
+    \caption[]{Box and violing plot showing the vertical distribution of particles.
+Short-living are those younger then 3 months and long-living all those older then that.
+Depth is measure relative to the current water surface with positive numbers being above the water surface i.e.
+stranded on the shore.}
     \label{fig:migration-long-vs-short}
@@
     \includegraphics[width=\columnwidth]{heatmap.png}
-    \caption[]{heatmap of long-living to short-living-particles}
+    \caption[]{Hexbin heatmap showing average particle age per bin showing the brackish area of the Elbe downstream of the harbor.
+Yellowish colors indicate an average age of over three months.
+\texttt{add labels and maybe use a different colormap?}}
     \label{fig:migration-long-vs-short-heatmap}
@@
 This is likely due to the fact that they are more likely be washed up high up at shore where the water reaches them less frequently.
-This effect is emphasized in flatter regions where distances between the driftline and waters that don't fall dry is large, increasing the distance that the plankton has to travel without settling to be washed out.
+This effect is emphasized in flatter regions.
+Here the distance between the wash margin and deeper waters that do not fall dry is larger. 
 We assume that this effect would be further emphasized if we were to represent stokes drifts in our model even though wind forcing on the surface water is included.
@@
 Additionally, the bottom water layer in the Elbe typically has higher upstream velocity comparts to the upper water column.
-However, they still seem to be less successful then positively buoyant particles.
+However, they still seem to be less successful than positively buoyant particles.
 Examining our model we think  that this is because they tend to quickly aggregate in deep waters with little to no light where they are unable to sustain themselves and ultimately die out.
@@
 
-It is interesting to see that in case of diel migration success does not seem to depend on whether they are moving up during the day or the night.
+We find it interesting to see that in the case of diel migration success does not seem to depend on whether they are moving up during the day or the night.
 We assume that this is to a large part due to them only surviving in shallow near-shore waters where light limitation does not play a role.
@@
 One major limitation of our model is the lack of nutrient limitation or the lack of a realistic ecosystem model in general.
-This could have been done by coupling our Lagrangian model online to an Eularian model that would have advected changes in concentrations fields by biotic activity throughout the model domain
+This could have been done by coupling our Lagrangian model online to an Eularian model that would have advected changes in concentrations fields by biotic activity throughout the model domain.
 However, this would have drastically increased both developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.

### 2023-08-01 ms.tex
draft sent to ross

--- 2022-02-22 ms.tex
@@
 	\vspace{-20mm}\textbf{	\large 
-	Plankton Retention mechanism studies at Elbe with IBM}
+	Phytoplankton Retention Mechanisms in Estuaries: A Case Study of the Elbe Estuary
+	}
 	}
@@
 		\maketitle
+
+		\begin{abstract}
+			\begin{center}
+				\texttt{All texts in this typewriter font are drafting comments.}
+				\smallskip
+			\end{center}
+
+			Phytoplankton are essential for estuarine ecosystems.
+			However, little is known how estuarine plankton retain themselves in an environment that continuously washes them into unfavorable marine waters.
+			We present a new individual-based Lagrangian model of the Elbe estuary examining several retention mechanisms - most importantly vertical migration and tidal stranding.
+			We further examine their ability to outgrow losses and provide first estimates on outwashing losses.
+			We found that shallow waters are crucial for the long term for the survival of the phytoplankton population.
+			\texttt{I feel like it sounds a bit dry. Maybe sparkle in a bit more excitement?}
+
+			\vspace{1cm}
+		\end{abstract}
   	\end{@twocolumnfalse}

### 2023-08-21 ms.tex
draft sent to jana

@@
-\documentclass[a4paper, 10pt, twocolumn]{article}
+%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
+%% ---------------------------------
+%% This template should be used for copernicus.cls
+%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
+%% For further assistance please contact Copernicus Publications at: production@copernicus.org
+%% https://publications.copernicus.org/for_authors/manuscript_preparation.html
 
-\usepackage{graphicx}                % optional für Grafiken
-\usepackage[font=footnotesize]{caption}
-\usepackage[section]{placeins}
-%\usepackage{tabularx}               % optional für Tabellen
-%\usepackage{multirow}               % optional für Tabellen
-\usepackage{url}                	 % optional für Internet Links
-\usepackage{hyperref}
+%% Please use the following documentclass and journal abbreviations for preprints and final revised papers.
 
-%\usepackage[small,bf]{caption3}     % Bildunterschriften
+%% 2-column papers and preprints
+\documentclass[npg, manuscript]{copernicus}
 
-%\usepackage{parskip}
-
-% verbatim with line line wrapping
-% \usepackage{listings}
-% \lstset{
-%    breaklines=true,
-%    basicstyle=\ttfamily}
-
-\usepackage{titlesec}
-\usepackage{amsmath}                
-\usepackage{amssymb}
-\usepackage{abstract}
-
-% temporal packages
-\usepackage{lipsum}
-\SetLipsumDefault{1}
-
-%\usepackage{enumitem}
-%\setlist{nosep} % or \setlist{noitemsep} to leave space around whole list
-
-\usepackage{xcolor}
-\newcommand{\highlight}[1]{\colorbox{yellow}{\parbox{\columnwidth}{#1}}}
-
-%\pagestyle{empty}                   % weder Kopf- noch Fußzeile auf 1. Seite
-\graphicspath{{figures/}}
+%% \usepackage commands included in the copernicus.cls:
+%\usepackage[german, english]{babel}
+%\usepackage{tabularx}
+%\usepackage{cancel}
+%\usepackage{multirow}
+%\usepackage{supertabular}
+%\usepackage{algorithmic}
+%\usepackage{algorithm}
+%\usepackage{amsthm}
+%\usepackage{float}
+%\usepackage{subfig}
+%\usepackage{rotating}
 
@@
 
-\date{\small\today}                                        % kein Datum auf 1. Seite
-\title{
-	\vspace{-20mm}\textbf{	\large 
-	Phytoplankton Retention Mechanisms in Estuaries: A Case Study of the Elbe Estuary
-	}
-	}
-\author{
-	Laurin Steidle\(^1\), Ross Vennel, Jana Hinners \\
-	\emph{
-		\small
-		\(^1\)laurin.steidle@uni-hamburg.de} \\
-	\emph{
-		\small Universitaet Hamburg, Grosse Elbstrasse 133, Germany}
-		}
+\title{Phytoplankton Retention Mechanisms in Estuaries: A Case Study of the Elbe Estuary}
 
-%--------------------------------------Abstract------------------------------------------------
+% \Author[affil]{given_name}{surname}
 
-\twocolumn[
-	\begin{@twocolumnfalse}
+\Author[1]{Laurin}{Steidle}
+\Author[2]{Ross}{Vennell}
+\Author[3]{Jana}{Hinners}
+
+\affil[1]{Universität Hamburg, Olbersweg 24, 22767 Hamburg, Germany}
+\affil[2]{Cawthron Institute, 98 Halifax Street East,
+Nelson 7010, New Zealand}
+\affil[3]{Helmholtz-Zentrum Hereon, Max-Planck-Straße 1, 21502 Geesthacht, Germany}
+
+%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
+
+%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Smith}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".
+
+%% If authors contributed equally, please mark the respective author names with an asterisk, e.g.
+"\Author[2,*]{Anton}{Smith}" and "\Author[3,*]{Bradley}{Miller}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".
+
+\correspondence{Laurin Steidle (laurin.steidle@uni-hamburg.de)}
+
+\runningtitle{TEXT}
+
+\runningauthor{TEXT}
+
+\received{}
+\pubdiscuss{} %% only important for two-stage journals
+\revised{}
+\accepted{}
+\published{}
+
+%% These dates will be inserted by Copernicus Publications during the typesetting process.
+
+\firstpage{1}
+
 \maketitle
@@
 \begin{abstract}
-			\begin{center}
-				\texttt{All texts in this typewriter font are drafting comments.}
-				\smallskip
-			\end{center}
-
-			Phytoplankton are essential for estuarine ecosystems.
+    Phytoplankton is essential for the proliferation of estuarine ecosystems.
     However, little is known how estuarine plankton retain themselves in an environment that continuously washes them into unfavorable marine waters.
@@
     We further examine their ability to outgrow losses and provide first estimates on outwashing losses.
-			We found that shallow waters are crucial for the long term for the survival of the phytoplankton population.
-			\texttt{I feel like it sounds a bit dry. Maybe sparkle in a bit more excitement?}
+    We found that shallow waters are crucial for the long term survival of the phytoplankton population.
+\end{abstract}
 
-			\vspace{1cm}
-		\end{abstract}
-  	\end{@twocolumnfalse}
-]
+% \copyrightstatement{TEXT} %% This section is optional and can be used for copyright transfers.
 
-\input{main_text.tex}
+\introduction  %% \introduction[modified heading if necessary]
+\texttt{Add a short paragraph as an outlook of the paper? I was told thats what the cool kids do, but I have a hard time getting it work.}
 
-\bibliography{references.bib}
-\bibliographystyle{unsrt}
+Estuaries are important both for the climate system and for anthropogenic use.
+From a climate perspective they are highly productive relative to their surface area,
+which results in large amounts of direct carbon cycling 
+along with their important role as a source of nutrients and hatching grounds for marine ecosystem 
+\citep{Cloern2014,Arevalo2023}
+% (Phytoplankton primary production in the world’s estuarine-coastal ecosystems,Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment). 
+They are also strongly anthropogenically influenced through e.g.
+harbors, diking, dredging, fishing and pollution which can be seen as a stressor on the ecosystem, but also adds an anthropogenic dependency on the ecosystem.
+\citep{Jennerjahn2013, Brown2022, Wilson2002}
+% (Pressures, stresses, shocks and trends in estuarine ecosystems: An introduction and synthesis,
+% Anthropogenic-estuarine interactions cause disproportionate greenhouse gas production: A review of the evidence base,
+% Productivity, Fisheries and Aquaculture in Temperate Estuaries)
+
+Estuaries are also complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic changes.
+This complexity arises in part from the bathymetrie and the resulting hydrodynamics at the model boundaries \citep{Michael2016, Fringer2019}. 
+% (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
+% The future of coastal and estuarine modeling: Findings from a workshop)
+
+As for most ecosystems - their dynamics are strongly controlled by the primary producers \citep{Chen2023}
+% (Advances in phytoplankton population ecology in the Pearl river estuary).
+Phytoplankton, the 	predominant aquatic primary producer, is mostly drifting passively through the water.
+Hence, its trajectory is strongly controlled by the currents and tides.
+With the estuary having a net outwards flow, we would expect phytoplankton to be moving
+ downstream over time and, ultimately, to be washed out into marine, high salinity waters.
+Therefore, we are wondering how phytoplankton manage to survive as a population in such an environment.
+If we assume that the population is not exclusively maintained by upstream plankton that is washed into the estuary,
+ then there must be some sort of retention mechanism that enables a phytoplankton population to remain in the low-salinity area of the estuary.
+While we suspect that such a mechanism is crucial for survival, it is currently not well understood 
+due to the difficulty in either measuring or modeling such processes.
+
+% \subsection{Current state of knowledge}
+
+We categorize the already performed studies on the subject of phytoplankton retention in observational and modelling studies.
+
+Observational studies have found three major mechanism for retention - sinking, diel migration, and stickiness.
+Diel migration - moving up and down with the sun - has been found to be favorable for retention.
+However, this only has been demonstrated for larger motile estuarine organisms i.e. zooplankton species \citep{Hall2015,Kimmerer2002, Crawford1991}
+\citep{Anderson1985} was able to describe an estuarine embayment environnement in which dinoflagellates outcompeted other non-motile phytoplankton by  diel vertical migration.
+They were also able to show that outwashing loses were lower than their growth rate - increasing their abundance.
+However, this also implies that the growing part of the population is somehow retaining their position.
+If the regrowing population is also continuously drifting downstream they will will not able to sustain their population in that area and ultimately die out due to unfavorable salinity conditions \citep{Admiraal1976, vonAlvensleben2016, Jiang2020}.
+% (Vertical spatio-temporal patterns of phytoplankton due to migration behaviors in two shallow, microtidal estuaries: Influence on phytoplankton function and structure,
+% Persistence of Tidally-oriented Vertical Migration by Zooplankton in a Temperate Estuary,
+% Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate, Flocculation kinetics and mechanisms of microalgae- and clay-containing suspensions in different microalgal growth phases).
+Diatoms or benthic microalgea in particular have been observed to be strongly negatively buoyant, hence sinking to the ground and remaining there for a long time \citep{Passow1991}
+% (Species-specific sedimentation and sinking velocities of diatoms).
+Studies also found sticky compounds in phytoplankton aggregates that are suspected to let them stick to suspended particles making them sink to the ground or sticking to their surroundings aiding retention \citep{Kiørboe1993,VanderLee2000}.
+% (Phytoplankton aggregate formation: observations of patterns and mechanisms of cell sticking and the significance of exopolymeric material,
+% Parameters affecting mud floc size on a seasonal time scale: The impact of a phytoplankton bloom in the Dollard estuary)
+
+Next we want to highlight several modelling studies that examined whether certain retention mechanism could work in theory.
+% Two theoretical modelling studies examine the mechanisms of "loss compensation" and "reseeding".
+% \texttt{I don't like the term theoretical modelling study. Maybe "conceptual", "abstract" or "simple"?}
+% (Selective retention of two dinoflagellates in a well-mixed estuarine embayment: the importance of diel vertical migration and surface avoidance)
+% \citep{Anderson1985} describes a set-up in which a population is able to maintain itself if the losses due to outwashing can be compensated by growth.
+% In a environment where a lagoon like ecosystem was examined whether  they were able to quantify reproduction thresholds that are needed to maintain the population.
+% However, this setup also implies that the growing part of the population is somehow retaining their position.
+% If the regrowing population is also continuously drifting downstream they will also ultimately die out due to unfavorable salinity conditions \citep{Admiraal1976, vonAlvensleben2016, Jiang2020}.
+% (Salinity Tolerance of Benthic Estuarine Diatoms as Tested with a Rapid Polarographic Measurement of Photosynthesis,
+% Salinity tolerance of four freshwater microalgal species and the effects of salinity and nutrient limitation on biochemical profiles,
+% Drivers of the spatial phytoplankton gradient in estuarine–coastal systems: generic implications of a case study in a Dutch tidal bay) 
+% \citep{}
+% studied the concept of "reseeding".
+% Here a system is theorized in which a population is able to repopulate a previously 
+% lost or uncolonized area upstream due to local upstream currents e.g. through tides or dispersion.
+% While they showed that upstream reseeding is conceptually possible they did so in a specifically contrived system resembling groyne fields.
+% Whether this is an actually realized phenomenon depends on the topology and hydrodynamic of the estuary.
+% Nevertheless, it has been shown that both "outgrowing the losses" and "reseeding" are possible retention mechanism under certain conditions.
+In \citep{Simons2006,Kimmerer2014} they use a particle tracking model also known as Lagragian model to study Zooplankton retention.
+They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention tactic.
+However, they did so by ignoring key processes like reproduction, mortality and any stranding or sedimentation processes in a low resolution structured grid model.
+
+We are examining the Elbe estuary as a case study.
+Like most alluvial estuaries it is relatively flat.
+Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centuries.
+Most notably diking to restrain it to a narrow channel and dreading to improve access to the Hamburg harbor.
+Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
+This causes the river to experience a sudden jump in bathymetry from approx. 5 \unit{m} at border of the city to approx. 20 \unit{m} in the harbor (see fig. \ref{fig:bathymetry}). 
+This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton biomass causing an increase in oxygen depletion and high ammonium reminaralization after the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
+While major aspects of the along-channel biochemical dynamics have been studied little is known regarding their vertical and shore-to-shore dynamics.
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/harbor.png}
+    \caption{Bathymetry of the Elbe model around Hamburg.
+Note the bathymetric jumps from ~5 \unit{m} on the right, upstream side to a short 10 \unit{m} step in the upper harbor area to the ~20 \unit{m} in the lower harbor area all the way to the North Sea.
+Also note that there exists no channel that does not pass through the 10 \unit{km} of exclusively 20 \unit{m} deep channel.}
+    \label{fig:bathymetry}
+\end{figure}
+
+To summarize - several concepts for potential retention mechanisms have been found
+ and shown to work in certain conditions.
+However, all approaches used major simplification which didn't allow them to study the full picture.
+We developed a new and more realistic approach to study retention using a particle tracking model.
+This model combines the features outlined above, namely: reproduction, mortality, vertical migration, stranding, sedimentation, and reseeding.
+We examine if such a system can reproduce phytoplankton retention in the Elbe estuary with these mechanisms and if so under which conditions.
+
+\section{Methods}
+\subsection{Model description}
+
+\subsubsection{Particle tracking}
+
+In our study we take a particle tracking approach with a Lagrangian model called OceanTracker \citep{Vennell2021}.
+Particle tracking offers several advantages for our question:
+While particle tracking on unstructured grids has been relatively computationally expensive until recently \citep{Vennell2021}
+it allows us to reuse already validated hydrodynamic models to model tracer like objects.
+This approach is overall much faster than recalculating the advection-diffusion-equation in an Eularian model.
+
+Additionally, because we are simulating individually particles we are able to observe their tracks.
+This makes it much easier and intuitive to interpret the results and allows us to include individual based properties and processes.
+
+\subsubsection{Hydrological model}
+We use the hydrodynamics data of the most recent SCHISM model \citep{Pein2021}.
+It is a three-dimensional unstructured  grid model representing the full Elbe estuary from the weir at Geesthacht to the North Sea including several side-channels and the harbor area (see fig.
+\ref{fig:model_domain}).
+% (Look for more details in pein paper that might be interesting)
+The model provides a node-based mesh field containing a range of information.
+Most important for our purpose is the water velocity but also salinity, water level and dispersion.
+The year represented in that dataset is 2012 with 1h temporal- and dynamically varying spacial resolution with a median distance between nodes of approx. 60 \unit{m}.
+
+\begin{figure}
+    \includegraphics[width=12cm]{figures/dfg_foto_wettbewerbg_laurin_legendless_steidle_small.png}
+    \caption{Map of the full model domain with Geesthacht being the upstream boarder on the right and the North-sea being the downstream edge on the top left.
+The gray outline indicates the edge of the model domain.
+Blue and green dots show an example snapshot of a phytoplankton distribution in the model.
+Blue being floating, green being stranded particles.}
+    \label{fig:model_domain}
+\end{figure}
+
+\subsubsection{Particle based processes}
+
+On top of the "blank" particle we add a set of biological features to achieve a more realistic planktonic behavior.
+The most important one is their ability to reproduce by creating copies of themselves.
+This is represented as a constant chance of duplicating a particle.
+We also remove them under certain biologically unfavorable conditions. Specifically when they are stranded on the shore or starved for light or float in high salinity waters for too long.
+More precisely, they are removed if either of the following criteria is met: 
+When they reach high salinity waters above $10$PSU implemented as a mortality chance $1\%$ every minute.
+When they are stranded outside the water and lie dry for more than $7$ consecutive days.
+When they are starved for light which is represented as a budget of $14$ days 
+where energy is lost in a fixed rate in low-light and gained in high light areas.
+\texttt{I got these numbers from you - do you have a reference or reasoning for them?}
+
+A set of vertical movement features is added to represent migration behavior. 
+This is implemented in two modes. 
+One is a constant vertical velocity i.e. sinking and rising
+The second is diel vertical migration where they move up and down depending on the phase of the sun.
+A large area of the estuary is very shallow with average water depths of below one meter. 
+We include a simple sedimentation and resuspension model to represent stranding by the tide and also sedimentation model for the water column where resuspension is depending on a critical sheer velocity (see appendix).
+
+Particles are stranded and temporarily stuck in place when they cross a model boundary or when their current location falls dry. 
+Particle are resuspended from the bottom based on a critical sheer velocity model.
+If they are stranded they resuspend as soon as there location is flooded again.
+
+The model also implements dynamic dispersion that is crucial to represent tidal-pumping processes and to allow for more realistic trajectories.
+
+\subsection{Experimental configurations}
+
+The first experiment of this study examines retention success under different scenarios.
+The general idea is to release a population of plankton at beginning of the estuary represented by the weir in Geesthacht.
+We then examine how the population distributes itself over the estuary and whether it is able to maintain its population size.
+% With this approach we study the effect of two retention mechanism in a sensitivity analysis. These being reproduction and migration.
+The initial particle batch that is representing our phytoplankton population is released at the beginning of the year.
+We evaluate their retention success by counting the amount of plankton in different regions in the estuary over time.
+If the population is growing at the end of the year we consider them "successfully retained".
+Population changes are measured at the end of the year relative to the population size after an initial spin-up time of three months when most of the initial batch has been washed out and the population is typically at its minimum.
+
+For each particle we log their distance traveled, age, water depth, and status - whether they are drifting, stranded on the shore or bottom.
+This allows us for example to compare successful (died older than three months) and unsuccessful (died younger than three months) plankton individuals.
+The logged observables are measured every 12 hours starting at midnight.
+
+\medskip
+
+We use this set-up in a range of scenarios.
+We call the first and simplest scenario "dust". 
+Here we do not allow for any reproduction and give all particles a neutral buoyancy.
+Hence, they are drifting like dust similar to a passive tracer.
+This is our reference for any retention strategy. 
+For a retention strategy to be considered beneficial it must perform better than these passive tracers.
+
+In the second scenario referred to as "migration" reproduction is re-enabled.
+In this scenario we study the effect of different growth rates and buoyancies i.e. vertical velocities on the phytoplankton population in a sensitivity analysis.
+
+The third scenario has been added for illustration purposes.
+Here, we use the same set-up as in the second scenario with the only exception of disabling reproduction for particles stranded on the shores of the estuaries.
+
+% \subsubsection{upstream migration}
+
+% In the second and minor experiment we test the hypothesis of "reseeding" - whether organisms can recolonize previously lost upstream habitats.
+% \texttt{This is currently not included anymore. Seemed to uninteresting and far off track. What do you think?}
+
+\subsection{Results}
+
+\subsection{Retentions success in different scenarios}
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/retention_success_sa_mono.png}
+    \caption[]{Relative population changes for the monotonic migration scenarios.
+Positive vertical velocities indicate an upwards drift.
+Positive population changes indicate a retention success (shown in green) while negative population changes predict a long term loss of the population (shown in brown).}
+    \label{fig:monotonic_retention_success}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/retention_success_sa_diel.png}
+    \caption[]{Relative population changes for the diel migration scenarios.
+Positive vertical velocities indicate an upwards drift.
+Positive population changes indicate a retention success (shown in green) while negative population changes predict a long term loss of the population (shown in brown).}
+    \label{fig:diel_retention_success}
+\end{figure}
+
+The results of the retention experiments are visualized as heatmaps in fig. \ref{fig:monotonic_retention_success} and \ref{fig:diel_retention_success}.
+Fig. \ref{fig:monotonic_retention_success} shows the results for the monotonic vertical migration scenarios e.g. constant sinking or rising.
+Fig.
+\ref{fig:diel_retention_success} shows the results for the diel vertical migration scenarios.
+For the monotonic case a positive vertical velocity indicates an upwards motion.
+For the diel migration case a positive vertical velocity means that the particles are moving upwards during the day and downwards during the night.
+Each pixel in the heatmap represents a scenario with a specific combination of vertical velocity and growth rate.
+The coloring indicates the relative population change after one year.
+A positive (green) value is considered a successful retention while a negative (brown) value indicates that the population is dying out.
+
+The first thing that we wish to highlight is that the population successfully retains itself in certain scenarios.
+For the case of the monotonic migration we see a trend that the populations with higher positive velocities and higher growth rates are more successful in retaining themselves.
+For the case of the diel migration we see a trend to higher vertical velocities both upwards and downwards.
+
+Another thing that we would like to point out is that populations with no reproduction are all unsuccessful in retaining themselves.
+
+We would also like to note that a white pixels or the boundary between green and brown are scenarios with a net-zero growth rate.
+Because in this case the losses must be equal to the growth we can use the growth rate as an estimate for the total relative losses due to outwashing, dry-out when stranded, and light starvation.
+
+\medskip
+
+In the scenario where we disabled reproduction for stranded particles we found that the population was unable to retain itself in all tested parameterizations.
+
+\subsection{Where do they survive?}
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/retention_boxplot.png}
+    \caption[]{Box and violin plot showing the vertical distribution of particles.
+Short-living are those younger than 3 months and long-living all those older than that.
+Depth is measure relative to the current water surface with positive numbers being above the water surface i.e.
+stranded on shore.}
+    \label{fig:migration-long-vs-short}
+\end{figure}
+
+We are now taking a closer look at the subset of phytoplankton that was able to retain themselves.
+In fig.
+\ref{fig:migration-long-vs-short} we see a box plot comparing the average water depth at particle location for the long living and short living subset.
+The long living being older than three months.
+Depth is measured relative to the water surface. Hence, a value above zero indicate that the particle is stranded on the shore.
+We can clearly see that the long living subset predominantly lives close to the shores in shallower water.
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/age_hexbin.png}
+    \caption[]{Hexbin heatmap of the Elbe with Hamburg in the bottom right showing average particle age per bin.  Yellowish colors indicate an average age of over three months.}
+    \label{fig:migration-long-vs-short-heatmap}
+\end{figure}
+
+This is further supported by fig. \ref{fig:migration-long-vs-short-heatmap}.
+Here we see a hex-bin-plot of the estuary. The coloring shows the average age of the particles in the bin.
+Bins with an average age of over three months are shown in a yellowish color.
+These yellow areas can exclusively be found along the shores in shallow areas.
+The major areas being Mühlenberger Loch (a) and the Haseldorfer Binnenelbe (b).
+We can also find minor areas at Wedeler Marsch (c), at the mouth of Wischhafener Süderelbe (d), Ruthenstrom, and Stör (e), and at Nordkedding (g) and Neufelder Marsch (h).
+
+\subsubsection{Discussion}
+
+\subsection{Interpretation and contextualization of Results}
+
+The range of growth rates that we found to be necessary for survival are small compared to those typically realized in nature where plankton populations can double their size in a day in ideal conditions.
+Our model is simplistic in the sense that we do not include any nutrient limitation which are assumed to drastically decrease the growth rates compared to ideal conditions. 
+Nevertheless, the growth rates that we found to be necessary are still much smaller compared to those typically realized in nature \citep{Koch2004}.
+
+The importance of shallow near-shore areas for retention has been suggested several times by our findings. 
+We assume that this is for the following reasons.
+Plankton that lives in this area falls dry on the shore on a regular basis due to the tides. Hence, they are not moving downstream for a large portion of the day.
+We further see that positively buoyant plankton is more successful in retaining themselves. 
+This is likely due to the fact that they are more likely to be washed up high up at shore where the water reaches them less frequently.
+This effect is emphasized in flatter regions.
+In flatter regions the distance between the wash margin and constantly flooded areas is larger. 
+We assume that this effect would be further emphasized if we were to represent Stokes drift in our model even though wind forcing on the surface water is included.
+
+Initially we expected that negatively buoyant plankton would be more successful in retaining themselves.
+Sinking particles have a reduced downstream velocity as they tend to get stuck on the ground more often.
+Additionally, the bottom water layer in the Elbe typically has higher upstream velocity compared to the upper water column.
+However, they still seem to be less successful than positively buoyant particles.
+Examining our model we think  that this is because they tend to quickly aggregate in deep waters with little to no light where they are unable to sustain themselves and ultimately die out.
+This is hinting at the negative impact of dredging on sinking plankton as it both increases depth and turbidity increasing the volume of dark water relative to the amount of illuminated waters.
+
+Interestingly, diel migration success does not seem to depend on whether they are moving up during day or night.
+We assume that this is to a large part due to them only surviving in shallow near-shore waters anyway where light limitation does not play a significant role.
+We think the reason for higher success with higher vertical velocities is similar to the constant vertical velocity case.
+When the upwards diel migration aligns with a high tide the particles are more likely to be stranded far out onto the shore decreasing their risk to be washed out quickly.
+The higher the upwards velocities the higher the chance to be in the utmost part of the water body during high tide.
+However, we also want to note that our light limitation model is very simplistic to keep the model complexity low.
+For plankton not to be light limited they only need to be above a depth of one meter regardless of current illumination conditions and local turbidity.
+Therefore, a more realistic light limitation model might introduce a positive bias towards those particles that move upwards during the day.
+
+\subsection{Limitations of the model}
+
+One major limitation of our model is the lack of nutrient limitation or the lack of a realistic ecosystem model in general.
+This could have been done by coupling our Lagrangian model online to an Eularian model that would have advected changes in concentrations fields by biotic activity throughout the model domain.
+However, this would have drastically increased both developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
+The key draw back of this is that growth rates could only be modeled as a constant rate assuming an "ad libitum" environment at all times.
+This introduces systematic errors in the population growth.
+In reality, areas which are beneficial for retention might run out of nutrients quickly and could not sustain the growth conditions necessary to sustain the local population.
+
+Another major limitation is the lack of any sub-grid-resolution structure on the shores - especially vegetation.
+In our representation we assume perfectly flat surfaces with a median distance between nodes of approximately 60 \unit{m}.
+This "polished" model representation is assumed to cause a strong negative bias towards retention success as some plankton could easily stick to the vegetation or other cavities caused by the sub-grid-resolution structure and reseed the area.
+
+While our model does have a sedimentation and resuspension mechanic based on critical sheer velocities we still assume a static bathymetrie with sediments not able to move or bury particles. 
+This masks potential losses due to particles being buried but also decreases resuspension times.
+
+\subsection{Outlook}
+
+Our results clearly suggested the importance of shallow near-shore areas for the survival of phytoplankton in the Elbe estuary.
+After our initial experiments suggested this, we considered improving our model to resolve additional ecosystem interactions especially nutrient limitation.
+This would allow us to predict the population dynamics more realistically to the point where we would be able to quantize the importance of shallow areas more accurately.
+However, the main limitation to do so was the lack of validation data for these regions. 
+Almost all chlorophyll data is gathered in the center of the river.
+While we consider the general trend shown by our model to be true - shallow areas are essential for survival.
+We can not quantify the effect of changes in these areas that might be caused by future dredging, diking or restoration attempt.
+To do this, we first require more specially resolved data for these areas.
+
+We were also surprised by our finding that sinking particles have a harder timer surviving in the estuary. 
+Since several decades the yearly average chlorophyll concentration in the Elbe estuary concentrations declined (data available at www.fgg-elbe.de/elbe-datenportal.html) while upstream concentrations do not show this effect.
+The reasons for this are not yet fully understood, but one potential reason is the increase in dreading activities.
+These increases the average turbidity therefore reducing the water volume where phytoplankton is able to grow.
+A large fraction of the upstream phytoplankton consists of diatoms \citep{Muylaert1999}.
+% (Spring phytoplankton assemblages in and around the maximum turbidity zone of the estuaries of the Elbe (Germany), the Schelde (Belgium/The Netherlands) and the Gironde (France)).
+Diatoms typically have negative buoyancies \citep{Passow1991}
+% (Species-specific sedimentation and sinking velocities of diatoms) 
+making them especially vulnerable to sinking to light limited waters.
+
+Another mechanism that might in part explain this effect but which is not yet explored in our model is the phytoplankton stickiness. 
+Phytoplankton, especially blooming one, has been shown to be sticky due to exudates \citep{Kiørboe1993,VanderLee2000,Dutz2005}.
+% (Phytoplankton aggregate formation: observations of patterns and mechanisms of cell sticking and the significance of exopolymeric material,
+% Parameters affecting mud floc size on a seasonal time scale: The impact of a phytoplankton bloom in the Dollard estuary,
+%  The Netherlands,Inhibition of copepod grazing by diatom exudates: a factor in the development of mucus aggregates?).
+In combination with dredging induced turbidity this induces losses due to plankton aggregates sticking to negatively buoyant suspended matter and subsequently sinking to the ground where they are starved of light.
+We are curious to which extent these mechanisms might explain the phytoplankton losses in the dredged areas of the estuary and suggest both a field and modeling study.
+A field study could provide estimates on aggregation speeds of phytoplankton and their sinking velocities depending on suspended matter concentration.
+A model study could create estimates on chlorophyll concentration losses caused by this effect.
+
+\conclusions  %% \conclusions[modified heading if necessary]
+
+We saw that reproducing phytoplankton is able to maintain their population strain without up- or downstream reseeding under certain conditions.
+We were also able to give a first estimate on the growth rates necessary for survival of the population depending on their buoyancy and their respective outwashing losses.
+Furthermore, the population is ultimately washed out and perishes if we suppress reproduction.
+The same is true if we do not suppress reproduction completely but only for particles currently not floating in the water body indicating the importance of the shores for retention.
+To investigate this further we looked at the average water depth of long living particles and the areas where they are found which further indicated the importance of shallow areas for retention.
+
+We suggest further sampling to measure chlorophyll concentrations in shallow waters - specifically at Mühlenberger Loch and Haseldorfer Binnenelbe - to quantify the importance of these areas on the general population health.
+
+%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
+%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
+
+% \codeavailability{TEXT} %% use this section when having only software code available
+
+% \dataavailability{TEXT} %% use this section when having only data sets available
+
+\codedataavailability{Input data can be requested from Johannes Pein (johannes.pein@hereon.de).
+OceanTracker's source code is available on github (https://github.com/oceantracker/oceantracker).
+Model outputs and their configuration file are registered at [add doi once pre-published]} %% use this section when having data sets and software code available
+
+% \sampleavailability{TEXT} %% use this section when having geoscientific samples available
+
+% \videosupplement{TEXT} %% use this section when having video supplements available
+
+\appendix
+
+\section{}    %% Appendix A
+
+\subsection{Detailed model description}     %% Appendix A1, A2, etc.
+
+For details on the hydrodynamical data used in this study we refer to \citep{Pein2021}.
+
+\medskip
+
+Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 10 \unit{s}.
+Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates.
+Dispersion was modeled using a random walk approach using a random number generator with a normal distribution.
+Horizontally the standard distribution of the random walk was set to 0.1 \unit{m/s}.
+The vertical dispersion used a dynamic standard deviation based on the eddy diffusivity of the SCHISM model represent tidal pumping.
+To avoid particle accumulation on the top and bottom of the water column we used a correction based on the eddy diffusivity gradient similar to \citep{Yamazaki2014} with the vertical displacement $\partial z$ of a particle $i$ is given by
+\begin{equation}
+    \partial z_{i} = K_{v}^{'}(z_{i}(n))\partial t + N(0, 2K_{v}(z_{i}))
+\end{equation}
+where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
+Buoyencies are implemented as constant vertical displacement governed by the sinking or rising velocity.
+Particles can strand on land or on the bottom of the water body.
+When stranded on the bottom they are stationary until resuspended based on a critical friction velocity of 0.009 \unit{m/s}.
+When stranded on land they are stationary until the cell is rewettet.
+A cell is considered wet if all three nodes of the cells' triangle are considered wet by the SCHISM wet flag.
+Particles are not allowed to disperse into dry cells.
+
+Particles are release at the beginning of the simulation at midnight of the 1.1.2012 at the upstream boundary of the model domain and are randomly distributed throughout the water column.
+The particle buffer size i.e. the number of particles tracked simultaneously is set to 10000 with half of that release initially.
+
+% Biological properties that govern reproduction and mortality are evaluated every minute.
+% For reproduction each particle has a constant chance to split based on the growth rate. 
+% If a particle splits the new particle is placed at the same position as the parent particle inheriting its parents properties.
+% Mortality is governed by salinity, light availability and moisture.
+% When a particle is in waters above 10 \unit{psu} it has a constant chance of to die 1\%.
+% Particles are release with a full light budget of  14 days
+
+\noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).
+
+%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
+
+%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
+%% They will be correctly named automatically.
+
+%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
+%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:
+
+% \appendixfigures  %% needs to be added in front of appendix figures
+
+\appendixtables   %% needs to be added in front of appendix tables
+
+%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.
+
+\authorcontribution{LS, RV, and JH contributed to conception of the study.
+LS designed the studies details and organized the hydrodynamic data.
+RV provided the source code for the original oceantracker.
+RS,LS improved on the original physical model and LS developed the biological model.
+LS performed the model simulations, post-processing, and visualization.
+LS wrote the draft of the manuscript.
+All authors contributed to manuscript revision, read, and approved the submitted version.}
+
+\competinginterests{All authors declare no competing interests.} %% this section is mandatory even if you declare that no competing interests are present
+
+% \disclaimer{TEXT} %% optional section
+
+\begin{acknowledgements}
+We thank Johannes Pein for providing essential hydrodynamic model data and his assistance in making sense of the data.
+Further, we thank Philipp Porada for providing helpful comments on the manuscript.
+This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) within the Research Training Group 2530: “Biota-mediated effects on Carbon cycling in Estuaries” (project number 407270017), contribution to Universität Hamburg and Leibniz-Institut für Gewässerökologie und Binnenfischerei (IGB) im Forschungsverbund Berlin e.V.
+\end{acknowledgements}
+
+%% REFERENCES
+
+%% The reference list is compiled as follows:
+
+% \begin{thebibliography}{}
+
+%     \bibitem{Michael2016} Michael, L and Eli, S and Stephen, G and Macwilliams, Michael L and Ateljevich, Eli S and Monismith, Stephen G and Enright, Chris.
+{San Francisco Estuary and Watershed Science An Overview of Multi-Dimensional Models of the Sacramento – San Joaquin Delta}.
+2016.
+
+%     \bibitem{Arevalo2023} Arevalo, Elorri and Cabral, Henrique N.
+and Villeneuve, Bertrand and Poss{\'{e}}m{\'{e}}, Carl and Lepage, Mario.
+{Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment}.
+Fish and Fisheries, 2023.
+
+%     \bibitem{Kimmerer2014} Kimmerer, Wim J.
+and Gross, Edward S.
+and MacWilliams, Michael L..
+{Tidal migration and retention of estuarine zooplankton investigated using a particle-tracking model}.
+Limnology and Oceanography, 2014.
+
+%     \bibitem{Kimmerer2002} Kimmerer, W. J. and Burau, J. R. and Bennett, W. A.. {Persistence of tidally-oriented vertical migration by zooplankton in a temperate estuary}. Estuaries, 2002.
+
+%     \bibitem{Vennell2021} Vennell, Ross and Scheel, Max and Weppe, Simon and Knight, Ben and Smeaton, Malcolm.
+{Fast lagrangian particle tracking in unstructured ocean model grids}.
+Ocean Dynamics, 2021.
+
+%     \bibitem{Kerner2007} Kerner, Martin. {Effects of deepening the Elbe Estuary on sediment regime and water quality}. Estuarine, Coastal and Shelf Science, 2007.
+
+%     \bibitem{Brown2022} Brown, Alison M.
+and Bass, Adrian M.
+and Pickard, Amy E..
+{Anthropogenic-estuarine interactions cause disproportionate greenhouse gas production: A review of the evidence base}.
+Marine Pollution Bulletin, 2022.
+
+%     \bibitem{Sanders2018} Sanders, Tina and Sch{\"{o}}l, Andreas and D{\"{a}}hnke, Kirstin.
+{Hot Spots of Nitrification in the Elbe Estuary and Their Impact on Nitrate Regeneration}.
+Estuaries and Coasts, 2018.
+
+%     \bibitem{Schroeder1997} Schroeder, Friedhelm.
+{Water quality in the Elbe estuary: Significance of different processes for the oxygen deficit at Hamburg}.
+Environmental Modeling and Assessment, 1997.
+
+%     \bibitem{Holzwarth} Holzwarth, Ingrid. {Implications of direct anthropogenic pressures on dissolved oxygen dynamics in a well-mixed estuary}.
+
+%     \bibitem{Pein2021} Pein, Johannes and Eisele, Annika and Sanders, Tina and Daewel, Ute and Stanev, Emil V.
+and van Beusekom, Justus E.
+E.
+and Staneva, Joanna and Schrum, Corinna.
+{Seasonal Stratification and Biogeochemical Turnover in the Freshwater Reach of a Partially Mixed Dredged Estuary}.
+Frontiers in Marine Science, 2021.
+
+%     \bibitem{Sukhodolov2004} Sukhodolov, Alexander and Nicklisch, Andreas and Engelhardt, Christof and Kru, Angela.
+{A study of phytoplankton spatial distributions , flow structure and characteristics of mixing in a river reach with groynes}.
+2004.
+
+%     \bibitem{Wilson2002} Wilson, J. G.. {Productivity, fisheries and aquaculture in temperate estuaries}. Estuarine, Coastal and Shelf Science, 2002.
+
+%     \bibitem{Passow1991} Passow, U.. {Species-specific sedimentation and sinking velocities of diatoms}. Marine Biology, 1991.
+
+%     \bibitem{Kiørboe1993} Ki{\o}rboe, Thomas and Hansen, J{\o}rgen L.s..
+{Phytoplankton aggregate formation: Observations of patterns and mechanisms of cell sticking and the significance of exopolymeric material}.
+Journal of Plankton Research, 1993.
+
+%     \bibitem{Holzwarth2019} Holzwarth, Ingrid and Weilbeer, Holger and Wirtz, Kai W. {The effect of bathymetric modification on water age in the Elbe Estuary}. 2019.
+
+%     \bibitem{Holzwarth2018} Holzwarth, Ingrid and Wirtz, Kai. {Anthropogenic impacts on estuarine oxygen dynamics: A model based evaluation}. Estuarine, Coastal and Shelf Science, 2018.
+
+%     \bibitem{Cloern2014} Cloern, J. E. and Foster, S. Q. and Kleckner, A. E.. {Phytoplankton primary production in the world's estuarine-coastal ecosystems}. Biogeosciences, 2014.
+
+%     \bibitem{Jennerjahn2013} Jennerjahn, Tim C.
+and Mitchell, Steve B..
+{Pressures, stresses, shocks and trends in estuarine ecosystems - An introduction and synthesis}.
+Estuarine, Coastal and Shelf Science, 2013.
+
+%     \bibitem{Zhang2016} Zhang, Yinglong J. and Ye, Fei and Stanev, Emil V. and Grashorn, Sebastian. {Seamless cross-scale modeling with SCHISM}. Ocean Modelling, 2016.
+
+%     \bibitem{Hall2015} Hall, Nathan S and Whipple, Anthony C and Paerl, Hans W.
+{Vertical spatio-temporal patterns of phytoplankton due to migration behaviors in two shallow, microtidal estuaries: Influence on phytoplankton function and structure}.
+Estuarine, Coastal and Shelf Science, 2015.
+
+%     \bibitem{VanderLee2000} van der Lee, W.T.B.. {Parameters affecting mud floc size on a seasonal time scale: The impact of a phytoplankton bloom in the Dollard estuary, The Netherlands}. 2000.
+
+%     \bibitem{Admiraal1976} Admiraal, W.. {Salinity tolerance of benthic estuarine diatoms as tested with a rapid polarographic measurement of photosynthesis}. Marine Biology, 1976.
+
+%     \bibitem{Simons2006} Simons, Rachel D.
+and Monismith, Stephen G.
+and Johnson, Ladd E.
+and Winkler, Gesche and Saucier, Fran{\c{c}}ois J..
+{Zooplankton retention in the estuarine transition zone of the St.
+Lawrence Estuary}.
+Limnology and Oceanography, 2006.
+
+%     \bibitem{Crawford1991} Crawford, DW and Purdie, DA. {Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate}. Marine Ecology Progress Series, 1991.
+
+%     \bibitem{Koch2004} Koch, Richard W.
+and Guelda, Debbie L.
+and Bukaveckas, Paul A..
+{Phytoplankton growth in the Ohio, Cumberland and Tennessee Rivers, USA: Inter-site differences in light and nutrient limitation}.
+Aquatic Ecology, 2004.
+
+%     \bibitem{VonAlvensleben2016} von Alvensleben, Nicolas and Magnusson, Marie and Heimann, Kirsten.
+{Salinity tolerance of four freshwater microalgal species and the effects of salinity and nutrient limitation on biochemical profiles}.
+Journal of Applied Phycology, 2016.
+
+%     \bibitem{Anderson1985} Anderson, DM and Stolzenbach, KD.
+{Selective retention of two dinoflagellates in a well-mixed estuarine embayment: the importance of diel vertical migration and surface avoidance}.
+Marine Ecology Progress Series, 1985.
+
+%     \bibitem{Dutz2005} Dutz, J.
+and {Klein Breteler}, W.C.M.
+and Kramer, G..
+{Inhibition of copepod feeding by exudates and transparent exopolymer particles (TEP) derived from a Phaeocystis globosa dominated phytoplankton community}.
+Harmful Algae, 2005.
+
+% \end{thebibliography}
+
+%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
+%% authors experienced with BibTeX only have to include the following two lines:
+%%
+\bibliographystyle{copernicus}
+\bibliography{references2.bib}
+%%
+%% URLs and DOIs can be entered in your BibTeX file as:
+%%
+%% URL = {http://www.xyz.org/~jones/idx_g.htm}
+%% DOI = {10.5194/xyz}
+
+%% LITERATURE CITATIONS
+%%
+%% command                        & example result
+%% \citet{jones90}|               & Jones et al. (1990)
+%% \citep{jones90}|               & (Jones et al., 1990)
+%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
+%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
+%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
+%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
+%% \citeauthor{jones90}|          & Jones et al.
+%% \citeyear{jones90}|            & 1990
+
+%% FIGURES
+
+%% When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage
+%% between bibliography and first table and/or figure as well as between each table and/or figure.
+
+% The figure files should be labelled correctly with Arabic numerals (e.g. fig01.jpg, fig02.png).
+
+%% ONE-COLUMN FIGURES
+
+%%f
+%\begin{figure}[t]
+%\includegraphics[width=8.3cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure}
+%
+%%% TWO-COLUMN FIGURES
+%
+%%f
+%\begin{figure*}[t]
+%\includegraphics[width=12cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure*}
+%
+%
+%%% TABLES
+%%%
+%%% The different columns must be seperated with a & command and should
+%%% end with \\ to identify the column brake.
+%
+%%% ONE-COLUMN TABLE
+%
+%%t
+%\begin{table}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table}
+%
+%%% TWO-COLUMN TABLE
+%
+%%t
+%\begin{table*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table*}
+%
+%%% LANDSCAPE TABLE
+%
+%%t
+%\begin{sidewaystable*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{sidewaystable*}
+%
+%
+%%% MATHEMATICAL EXPRESSIONS
+%
+%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
+%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
+%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
+%%%
+%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
+%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
+%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
+%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
+%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
+%%% Vectors are identified in bold italic font using \vec{x}
+%%% Matrices are identified in bold roman font
+%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
+%%% The character * should not be applied as mutliplication sign
+%
+%
+%%% EQUATIONS
+%
+%%% Single-row equation
+%
+%\begin{equation}
+%
+%\end{equation}
+%
+%%% PHYSICAL UNITS
+%%%
+%%% Please use \unit{} and apply the exponential notation
 

### 2023-09-08 ms.tex
draft sent to ross (previous one did not get feedback)

--- 2023-07-24 ms.tex
@@
-\documentclass[a4paper, 10pt, twocolumn]{article}
+%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
+%% ---------------------------------
+%% This template should be used for copernicus.cls
+%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
+%% For further assistance please contact Copernicus Publications at: production@copernicus.org
+%% https://publications.copernicus.org/for_authors/manuscript_preparation.html
 
-\usepackage{graphicx}                % optional für Grafiken
-\usepackage[font=footnotesize]{caption}
-\usepackage[section]{placeins}
-%\usepackage{tabularx}               % optional für Tabellen
-%\usepackage{multirow}               % optional für Tabellen
-\usepackage{url}                	 % optional für Internet Links
+%% Please use the following documentclass and journal abbreviations for preprints and final revised papers.
+
+%% 2-column papers and preprints
+\documentclass[npg, manuscript]{copernicus}
+
+%% \usepackage commands included in the copernicus.cls:
+%\usepackage[german, english]{babel}
+%\usepackage{tabularx}
+%\usepackage{cancel}
+%\usepackage{multirow}
+%\usepackage{supertabular}
+%\usepackage{algorithmic}
+%\usepackage{algorithm}
+%\usepackage{amsthm}
+%\usepackage{float}
+%\usepackage{subfig}
+%\usepackage{rotating}
 \usepackage{hyperref}
-
-%\usepackage[small,bf]{caption3}     % Bildunterschriften
-
-%\usepackage{parskip}
-
-% verbatim with line line wrapping
-% \usepackage{listings}
-% \lstset{
-%    breaklines=true,
-%    basicstyle=\ttfamily}
-
-\usepackage{titlesec}
-\usepackage{amsmath}                
-\usepackage{amssymb}
-\usepackage{abstract}
-
-% temporal packages
-\usepackage{lipsum}
-\SetLipsumDefault{1}
-
-%\usepackage{enumitem}
-%\setlist{nosep} % or \setlist{noitemsep} to leave space around whole list
-
-\usepackage{xcolor}
-\newcommand{\highlight}[1]{\colorbox{yellow}{\parbox{\columnwidth}{#1}}}
-
-%\pagestyle{empty}                   % weder Kopf- noch Fußzeile auf 1. Seite
-\graphicspath{{figures/}}
 
@@
 
-\date{\small\today}                                        % kein Datum auf 1. Seite
-\title{
-	\vspace{-20mm}\textbf{	\large 
-	\texttt{Placeholder Title:}  Retention}
-	}
-\author{
-	Laurin Steidle\(^1\), Ross Vennel, Jana Hinners \\
-	\emph{
-		\small
-		\(^1\)laurin.steidle@uni-hamburg.de} \\
-	\emph{
-		\small Universitaet Hamburg, Grosse Elbstrasse 133, Germany}
-		}
+\title{Phytoplankton Retention Mechanisms in Estuaries: A Case Study of the Elbe Estuary}
 
-%--------------------------------------Abstract------------------------------------------------
+% \Author[affil]{given_name}{surname}
 
-\twocolumn[
-	\begin{@twocolumnfalse}
+\Author[1]{Laurin}{Steidle}
+\Author[2]{Ross}{Vennell}
+
+\affil[1]{Universität Hamburg, Olbersweg 24, 22767 Hamburg, Germany}
+\affil[2]{Cawthron Institute, 98 Halifax Street East,
+Nelson 7010, New Zealand}
+
+%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
+
+%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Smith}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".
+
+%% If authors contributed equally, please mark the respective author names with an asterisk, e.g.
+"\Author[2,*]{Anton}{Smith}" and "\Author[3,*]{Bradley}{Miller}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".
+
+\correspondence{Laurin Steidle (laurin.steidle@uni-hamburg.de)}
+
+\runningtitle{TEXT}
+
+\runningauthor{TEXT}
+
+\received{}
+\pubdiscuss{} %% only important for two-stage journals
+\revised{}
+\accepted{}
+\published{}
+
+%% These dates will be inserted by Copernicus Publications during the typesetting process.
+
+\firstpage{1}
+
 \maketitle
@@
 \begin{abstract}
-			\begin{center}
-				\texttt{All texts in this typewriter font are drafting comments.}
-				\smallskip
-			\end{center}
+    Due to their role as primary producers, phytoplankton is essential for the proliferation of estuarine ecosystems.
+    However, as small drifting organisms, phytoplankton cannot resist currents.
+    Estuaries not only represent challenging habitats due to a strong salinity gradient, but also have net transport into oceans.
+    So far, little is known about how phytoplankton retain themselves in estuaries.
+    We present a new individual-based Lagrangian model of the Elbe estuary examining possible retention mechanisms of phytoplankton.
+In more detail, we investigated how reproduction, sinking and rising, as well as diel vertical migration might allow for population survival.
+    We find that rising favors the retention success, fast sinking cause the opposite.
+    We further provide first estimates on outwashing losses.
+    Our simulations illustrate that shallow stretches of water along the river banks are essential for the long-term survival of phytoplankton populations.
+    These results give important indications for ecosystem-based management of estuaries.
+\end{abstract}
 
-			% \begin{figure}
-			% 	\includegraphics[width=\columnwidth]{picto visual abstract.pdf}
-			% \end{figure}
-			\texttt{placeholder abstract: }Estuaries are a challenging environment for plankton.
-			Gradients are steep and currents are strong.
-			This make retention mechanisms are crucial for their survival.
-			However, very little experimental and theoretical evidence has been gathered on the mechanism of phytoplankton retention.
-			We will conduct a model study.
-			First we will show the importance of retention mechanism for plankton in the Elbe.
-			Secondly we will study different retention mechanism and comment on their importance.
-			\vspace{1cm}
-		\end{abstract}
-  	\end{@twocolumnfalse}
-]
+% \copyrightstatement{TEXT} %% This section is optional and can be used for copyright transfers.
 
-\input{main_text.tex}
+% \introduction  %% \introduction[modified heading if necessary]
+\section{Introduction}
 
+Estuaries represent highly productive ecosystems in relation to their area and contribute to the global carbon cycle along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
+% (Phytoplankton primary production in the world’s estuarine-coastal ecosystems,Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment). 
+While they are heavily influenced by anthropogenic stressors such as diking, dredging, and fishing which can be seen as a stressor on the ecosystem, they are of tremendous importance for anthropogenic usage
+\citep{Jennerjahn2013, Brown2022, Wilson2002}.
+Estuaries present challenging, dynamic  of its residents due to the strong salinity gradient and net transport to the sea.
+Here, we explore how phytoplankton, drifting small primary producers that form the basis of estuarine food webs, can sustain themselves in such dynamic environments.
+% (Pressures, stresses, shocks and trends in estuarine ecosystems: An introduction and synthesis,
+% Anthropogenic-estuarine interactions cause disproportionate greenhouse gas production: A review of the evidence base,
+% Productivity, Fisheries and Aquaculture in Temperate Estuaries)
+
+As for most ecosystems - their dynamics are strongly controlled by primary producers \citep{Chen2023}.
+% (Advances in phytoplankton population ecology in the Pearl river estuary).
+Phytoplankton represent the base of aquatic ecosystems. 
+Apart from biofilm-forming phytoplankton which are attached to their substrate \citep{Cheah2022},
+%A methodological review on the characterization of microalgal biofilm and its  extracellular polymeric substances 
+the vast majority of phytoplankton organisms is drifting passively and cannot withstand currents.
+% The exception being biofilm-forming or benthic phytoplankton which are attached to their substrate (cite).
+With the estuary having a net outwards flow, we would expect phytoplankton to be moving downstream over time and to be washed out from limnic waters, via brackish to marine waters.
+Hence, the question arises how phytoplankton as the drifting base of estuarine food webs is able to maintain their population size without declining due to the transport into the open ocean.
+If we assume that the population is not exclusively maintained by upstream phytoplankton that is washed into the estuary, then there must be some sort of retention mechanism that enables a phytoplankton population to remain in the estuary.
+% While we suspect that such a mechanism is crucial for survival, it is currently not well understood due to the difficulty in either measuring or modeling such processes.
+
+% \subsection{Current state of knowledge}
+
+So far, different theories exist on how estuarine phytoplankton is able to maintain its position.
+Previous observational studies suggested several possible mechanisms that can allow for the retention of phytoplankton populations in estuarine systems - vertical migration in the form of sinking,  rising or diel migration, stickiness.
+
+% \texttt{Comment on sinking and rising?}
+Diel vertical migration is a process where organisms move up and down in the water column in dependence of the sun phase.
+This movement can be favorable for retention as it allows particles to reduce the time in the faster downstream currents at the water surface.
+However, this has only been demonstrated for larger motile estuarine organisms such as zooplankton \citep{Hall2015, Kimmerer2002, Crawford1991,Hall2011}.
+% (Vertical spatio-temporal patterns of phytoplankton due to migration behaviors in two shallow, microtidal estuaries: Influence on phytoplankton function and structure,
+% Persistence of Tidally-oriented Vertical Migration by Zooplankton in a Temperate Estuary,
+% Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate, Flocculation kinetics and mechanisms of microalgae- and clay-containing suspensions in different microalgal growth phases).
+Yet, a study by \citep{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate their outwashing losses through reproduction increasing their abundance.
+However, this also implies that the growing part of the population is somehow retaining their position.
+If the regrowing population is also continuously drifting downstream they will not able to sustain their population in that area and ultimately die out due to unfavorable salinity conditions in marine waters \citep{Admiraal1976, vonAlvensleben2016, Jiang2020}.
+
+Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic due to their complex bathymetry \citep{Michael2016, Fringer2019}.
+% (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
+% The future of coastal and estuarine modeling: Findings from a workshop)
+Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
+In \citep{Simons2006,Kimmerer2014} they use a particle tracking model also known as Lagrangian model to study Zooplankton retention.
+\citep{Simons2006} examined the dispersal and flush time of mussel larvae in the St. Lawrence Estuary while \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
+They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention tactic.
+However, they did so by ignoring key processes like reproduction, mortality, and stranding or sedimentation processes.
+Moreover, both studies were based on low resolution structured grid models, which we suspect to under-represent the complex bathymetry of estuarine systems \citep{Ye2018}.
+
+Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the ground and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}
+% (Species-specific sedimentation and sinking velocities of diatoms).
+Studies also found sticky compounds in phytoplankton aggregates that are suspected to let them stick to suspended particles making them sink to the ground or sticking to their surroundings aiding retention \citep{Kiørboe1993,VanderLee2000}.
+% \texttt{Shall I remove sticking from intro as it is not tested in my model?}
+
+In summary, different retention mechanisms have been observed or examined in modeling studies.
+However, they did so in either in isolation in the case of the observational studies or with major simplifications for the modeling studies.
+There is currently a lack of theoretical studies that allow a more comprehensive overview of possible retention mechanisms.
+
+Here, we are exploring possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
+It is located in the north of Germany and flows into the North Sea.
+Like most alluvial estuaries, it is relatively shallow, with most of it averaging only a few meters in average depth.
+Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centuries.
+Most notably diking to restrain it to a narrow channel and dredging to improve access to the Hamburg harbor.
+Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
+This causes the main channel to experience a sudden jump in bathymetry from approx. 5 \unit{m} at border of the city to approx. 20 \unit{m} in the harbor (see fig. \ref{fig:bathymetry}). 
+This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton biomass causing an increase in oxygen depletion and high ammonium remineralization after the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
+Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
+While important aspects of the along-channel biochemical dynamics have been studied, little is known about their vertical and shore-to-shore dynamics \citep{Goosen1999, Dahnke2008, Sanders2018}.
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/harbor.png}
+    \caption{Bathymetry of the Elbe model around Hamburg.
+Note the bathymetric jumps from ~5 \unit{m} on the right, upstream side to a short 10 \unit{m} step in the upper harbor area to the ~20 \unit{m} in the lower harbor area all the way to the North Sea.
+Also note that there exists no channel that does not pass through the 10 \unit{km} of exclusively 20 \unit{m} deep channel.}
+    \label{fig:bathymetry}
+\end{figure}
+
+For this purpose, we further developed the individual-based Lagrangian model OceanTracker \citep{Vennell2021} and applied it to the Elbe estuary using the hydrodynamics calculated by a recent model SCHISM \citep{Pein2021}.
+While the Lagrangian model simulated the movement of the inanimate organism, we included key phytoplankton features such as reproduction and mortality, sinking and rising, as well as diel vertical migration.
+% \textit{and dormancy?}
+Using this model, we investigate under which conditions phytoplankton retention can be reproduced.
+
+\section{Methods}
+
+\subsection{Model description}
+
+%\subsubsection{Lagrangian model}
+
+In our study we use a Lagrangian approach with the particle tracking model OceanTracker \citep{Vennell2021}.
+While particle tracking on unstructured grids has been relatively computationally expensive until recently \citep{Vennell2021}, it offers several advantages.
+Firstly, it allows us to reuse already computed hydrodynamic models to model tracer-like objects.
+This is overall much faster than recalculating the advection-diffusion-equation in an Eularian model.
+Secondly, because we are simulating individually particles we are able to observe their tracks.
+This makes the interpretation of our results more intuitive and allows us to include individual based properties and processes that can only be parametrized in Eulerian models.
+
+%\subsubsection{Hydrodynamics data}
+We use the hydrodynamics data of the most recent SCHISM model of the Elbe \citep{Pein2021}.
+This model uses a three-dimensional unstructured  grid to represent the full Elbe estuary from the weir at Geesthacht to the North Sea, including several side-channels and the harbor area (see fig.
+\ref{fig:model_domain}).
+% (Look for more details in pein paper that might be interesting)
+The model provides us with a node-based mesh containing a range of information such as water velocity, salinity, water level and dispersion.
+The year represented in that dataset is 2012 with a temporal resolution of 1 hour and a dynamically varying spacial resolution with a median distance between nodes of approx. 60 \unit{m}.
+
+\begin{figure}
+    \includegraphics[width=12cm]{figures/dfg_foto_wettbewerbg_laurin_legendless_steidle_small.png}
+    \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left.
+The gray outline marks the edge of the model domain.
+Blue and green dots show an example snapshot of a phytoplankton distribution in the model.
+Blue represents floating, green settled particles.
+The background map has been provided by OpenStreetMap.}
+    \label{fig:model_domain}
+\end{figure}
+
+%\subsubsection{Particle based processes}
+
+We add a set of biological features on top of the otherwise inanimate organism.
+These features include reproduction and mortality, vertical movement in form of sinking, rising or diel vertical migration, stranding, and sedimentation.
+
+Reproduction is represented as a chance of creating copies of themselves. 
+We perform several simulations for a range of reproduction rates that are constant over the lifetime of the particle and the chance to reproduce is evaluated every minute.
+
+Mortality is induced either by high salinity, when they dry-out while stranded, or due to long term light limitation.
+When particles are exposed to high salinity water above $20$PSU, a mortality probability of $1\%$ per minute is imposed.
+This threshold is chosen based on the salinity tolerances presented in \citep{vonAlvensleben2016}.
+This is only an approximation and many phytoplankton species deviate from this number.
+However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for several hours and are assumed not to return again through this isohaline.
+We consider particles that are stranded outside the water and lie dry for more than $7$ consecutive days to be dead and remove them.
+Particles are also expected to be dead and are removed if they are light-limited for $28$ days \citep{Walter2017}.
+They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
+A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by $1$ minute, while the opposite applies when they are above $1 \unit{m}$.
+Children of light-limited parents inherit the remaining light budget of their parents.
+
+\medskip
+
+We study the effect of different patterns of vertical motion.
+The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton organisms.
+This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the aggregate in which the organism is embedded.
+For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
+The second mode of vertical motion is diel vertical migration.
+Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night using the same velocity range.
+
+% A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
+We include a sedimentation and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
+Particles become stranded when the current grid cell falls dry. 
+They are not allowed to disperse into dry cells.
+A grid cell is considered dry if at least one of the nodes of the cells' triangle is considered dry by the SCHISM wet flag.
+Once this cell is rewettet all stranded particles resuspend.
+Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
+The velocity profile in the bottom layer, or log layer, is calculated by
+% centered equation
+\begin{equation}
+    U(z) = \frac{u_{*}}{\kappa} ln \frac{z}{z_{0}},
+\end{equation}
+where U is the friction velocity representing the drag at height z  above the seabed, $\kappa$ is the van Karman constant, $z_{0}$ is a length
+scale reflecting the bottom roughness, and $u^{*}$ is the critical friction velocity.
+If the friction velocity is above the critical friction velocity the particle is resuspended.
+Particles that are stranded or settled on the bed are allowed to reproduce.
+Particles are not only advected but also dispersed based on eddy diffusivity.
+This allows us implement a dynamic dispersion that is crucial to represent tidal-pumping processes.
+Dispersion was modeled using a random walk using a random number generator with a normal distribution.
+Horizontally the standard distribution of the random walk was set to 0.1 \unit{m s^{-1}}.
+The displacement by vertical dispersion $\partial z$ of particle $i$ is calculated by
+\begin{equation}
+    \partial z_{i} = K_{v}^{'}(z_{i}(n))\partial t + N(0, 2K_{v}(z_{i}))
+\end{equation}
+based on \citep{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
+The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column.
+
+For each particle we log their distance traveled, age, water depth, and status (whether they are drifting or settled on the river bank or bottom).
+This allows us for example to compare successfully retained particles (older than three months) unsuccessfully retained particles (dead after less than three months).
+These observables are measured every 12 hours starting at midnight.
+
+%\subsubsection{Numerics}
+Model simulations and visualizations were performed in Python making heavy use of Numba, a LLVM-based Python JIT compiler \citep{Lam2015} to drastically speed up the simulations \citep{Vennell2021}.
+Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 10 \unit{s}.
+Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates.
+
+\subsection{Experimental configurations}
+
+We perform two sets of experiments to test the influence of different vertical movements on the retention success of phytoplankton in the Elbe estuary.
+
+In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm s^{-1}} in $2$ \unit{mm s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
+Each vertical velocity is examined for a range of different growth rates with effective population doubling times ranging from 772 to 77 days with a logarithmic scaling.
+Hence, a total of 187 different scenarios are tested.
+In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and growth rates.
+
+In both set of experiments, we release the studied phytoplankton population at the beginning of the year homogeneously distributed in a volume covering the full water column at the weir in Geesthacht and examine how the population distributes itself over the estuary and whether it is able to maintain its population size over time.
+Conceptually, we consider a population to be successfully retained if it is showing long term growth. 
+Practically, this is evaluated by comparing the population size at the end of the year to the size after release.
+The first three months of the simulations are considered as initial model spin-up time during which the initial population is dispersed downstream throughout the estuary.
+Population size changes are measured at the end of the year relative to the population size after this initial spin-up time.
+
+\section{Results}
+
+\subsection{Retentions success in different scenarios}
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/retention_success_combined.png}
+    \caption[]{
+        Relative population changes for the monodirectional movement (a) and diel migration (b) scenarios.
+Positive vertical velocities indicate an upwards drift.
+Positive population changes represent a retention success (green) while negative population changes represent a loss of the population (brown).
+        \texttt{I think I'll redo the simulation and shift the pop doubling time slightly to the left}
+        }
+    \label{fig:retention_success}
+\end{figure}
+
+The results of the retention experiments are visualized as heatmap in fig. \ref{fig:retention_success}.
+Fig.
+\ref{fig:retention_success}a shows the results for the monodirectional vertical migration scenarios i.e.
+constant sinking or rising.
+Fig.
+\ref{fig:retention_success}b shows the results for the diel vertical migration scenarios.
+Each pixel in the heatmap represents a simulation with a specific combination of vertical velocity and growth rate.
+The coloring indicates the relative population change after one year.
+White pixels and the boundary between green and brown pixels represent net-zero growth rate simulations.
+In this case, the losses are equal to the growth.
+Therefore, we can use the growth rate as an estimate for the total relative losses due to downstream transport, drying out while being stranded, and light starvation.
+
+Our simulations show that the population is able to successfully retains itself under certain conditions.
+For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and a higher growth rate is more beneficial for retention success than a downward oriented velocity (sinking) and lower growth rates.
+As expected, simulations in which the reproduction is set to zero do not show any retention success.
+While it is easy to understand that high growth rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.
+
+For the case of the diel vertical migration we see a equal or higher retention success compared to no vertical velocity with velocities in the range of 4-10 \unit{mm s^{-1}} vertical velocities but shows to me less successful than the monodirectional movement for all cases.
+A vertical velocity of 2 \unit{mm s^{-1}} is less successful then no migration.
+Most importantly, none of the diel migration scenarios are improving the retention success compared to passively drifting organisms.
+
+\subsection{Spatial analysis}
+
+We are now taking a closer look at spacial factors that allow phytoplankton particles to maintain net growth in the estuary. 
+For this analysis we used data from both sets of experiments.
+Fig.
+\ref{fig:migration-long-vs-short} compares two box plots showing the average water depth at the location of each particle between particles that remained alive for less than three months (short-living) and for more than three months (long-living).
+Depth is measured relative to the current water surface.
+Hence, a value above zero indicate that the particle is stranded on the river bank.
+These analyses demonstrate that long-living particles predominantly live close to the river banks in shallower waters or on tidal flats.
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/retention_boxplot.png}
+    \caption[]{Box and violin plot showing the vertical distribution of particles.
+Short-living are those younger than 3 months and long-living all those older than that.
+Depth is measure relative to the current water surface with positive numbers being above the water surface i.e.
+stranded on shore.}
+    \label{fig:migration-long-vs-short}
+\end{figure}
+
+We moreover analyzed the horizontal spacial distribution of long and short-living particles in fig. \ref{fig:migration-long-vs-short-heatmap}.
+To do this, we divide the model domain into equally sized hexagons.
+The color of each hexagon indicates the average age of the particles within it.
+Hexagons with a yellow color indicate an average age of over three months.
+These yellow areas are mainly found along the river banks in shallow waters or tidal flats.
+
+\begin{figure}
+    \includegraphics[width=8.3cm]{figures/age_hexbin.png}
+    \caption[]{
+        Hex-bin heatmap of the Elbe with Hamburg in the bottom right showing average particle age per bin.
+        Colors indicate the age of the particles, with yellowish colors indicating an average age of over three months.
+        Yellow areas are mainly found along the river banks in shallow waters or tidal flats.
+        The largest areas are the Mühlenberger Loch (a) and the Haseldorfer Binnenelbe (b).
+        Smaller areas can also be found at Wedeler Marsch (c), at the mouth of Wischhafener Süderelbe (d), Ruthenstrom, and Stör (e), and at Nordkedding (g) and Neufelder Marsch (h).
+        \texttt{I am aware that the unit of age (days) is still missing. This figure will be modified slightly before publication.}
+        }
+    \label{fig:migration-long-vs-short-heatmap}
+\end{figure}
+
+To further investigate the reasons for the positive effect of buoyancy and the importance of shallow waters and tidal flats, we repeated the first set of simulations and disabled the reproduction of settled and stranded particles.
+Under these conditions, populations were unable to retain themselves in the estuary, regardless of their vertical velocity and growth rate indicating that tidal flats are essential for the survival of the population.
+
+%\subsubsection{Discussion}
+
+\subsection{Interpretation and contextualization of Results}
+
+In this study, we investigated different strategies to explain how phytoplankton populations are able to maintain their growth in estuaries while constantly being at risk to be transported into the open ocean.
+% We find that reproduction rates in the range of XX-XX are sufficient for phytoplankton organisms to survive in the estuary under certain conditions.
+% In addition, we found that buoyancy aids the retention success, whereas sinking decreases the retention success.
+Finally, especially shallow areas near the river banks and tidal flats represent important habitats for phytoplankton organisms and allow them to survive, whereas deeper areas lead to a faster transport of estuarine phytoplankton into the marine waters.
+
+The limit for population doubling times that we found to be necessary for survival are of the order of around 200 days.
+Doubling times typically realized in natures are of the order of a few days which is two magnitudes small then those that we found necessary in our model \citep{Koch2004,Wirtz2011}.
+The low growth rates required for successful retention demonstrate that our model is also meaningful under more realistic environmental conditions, for example if maximum growth rates cannot be reached due to nutrient or temperature limitation.
+
+Our results suggest that shallow areas are very important for maintaining the estuary phytoplankton population.
+Plankton that lives in this area falls dry on the shore on a regular basis due to the tides. 
+Therefore, they do not move for much of the day.
+We further see that positively buoyant plankton is more successful in retaining themselves. 
+This is probably because they are more likely to be transported high up on the river bank where the water is less likely to reach them.
+This effect is emphasized in flatter regions as the distance between the wash margin and constantly flooded areas is larger increasing the chance of settlement or them stranding again.
+
+Initially, we expected sinking particles to have a higher retention success than buoyant ones. 
+However, we found that faster sinking particles are less successful in retaining themselves.
+Sinking velocities of less then $2 \unit{mm s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
+Sinking particles have a reduced downstream velocity because they sink to the river bed, where they move more slowly, or are stationary if they are temporarily settled or stranded.
+In addition, the deeper water layer of the Elbe typically has a higher upstream velocity than the upper water column due to temperature-induced density stratification \citep{Pein2021}.
+Nevertheless, buoyant particles were more successful in their retention in our simulations.
+The low chance of survival in the estuary for sinking particles might be explained by their light limitation in deeper waters.
+We expected particles to die if they are exposed to dark conditions for more than 2 weeks.
+Thus, sinking particles have a disadvantage to buoyant particles since they are more likely to become light limited and eventually die.
+This suggests that dredging has a negative impact on sinking plankton because it increases both depth and turbidity \citep{DeJonge2014}, which increases the aphotic depth and therefore the volume of dark water relative to the amount of light water.
+
+We suspect that the reason for the increased retention success of diel migrating organisms is similar to the monodirectional case.
+When the upwards diel migration coincides with high tide, particles are more likely to be stranded far out on the shore, reducing their risk of being washed out quickly.
+The higher the upward velocities, the greater the chance of being in the farthest part of the water body during high tide.
+However, because they are sinking for half of the day they also tend to be light limited more frequent than positively buoyant particles.
+It seems like these favorable and unfavorable processes balance each other out, resulting in a similar retention success as for the monodirectional case.
+
+\subsection{Model limitations \& future perspectives}
+
+In this study, we aimed to thoroughly investigate different possible retention mechanisms in a complex Lagrangian model system with a highly resolved bathymetry.  
+Due to this computational and spatial complexity, the complexity of the biological particle properties needed to remain simple to keep computational cost manageable and due to a lack of high resolution validation data.
+
+Our model design does not resolve more complex ecosystem dynamics such as nutrient limitation and grazing by higher trophic levels. 
+The Lagrangian model is performed offline, meaning it is not coupled to the Eulerian model that calculates the hydrodynamics and is performed after the fact.
+Therefore, modeling the advection and dispersal of changes in concentration fields e.g. nutrients due to growth or remineralization was not easily possible.
+Future modeling efforts could couple the Lagrangian model to a Eulerian model that disperses changes in concentrations fields by biotic activity throughout the model domain.
+However, this would have drastically increased both developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
+The key draw back of this is that growth rates could only be modeled as a constant rate in the current model description, similar to "ad libitum" experiments. 
+This can lead to systematic errors in estimating population growth.
+In reality, it can happen that nutrient limitation, which slows down the growth of the population, can occur, especially in the most light-saturated areas near the shore.
+For this reason, we may overestimate the role of shallow areas in our model.
+
+To be consistent with the complexity of the representation of biotic mechanisms, we use a simplistic light limitation.
+Particles are expected to be light limited below a water depth of 1 \unit{m} and not light limited above this threshold. 
+We have not included a more complex light limitation model that takes into account current light availability and attenuation.
+A more realistic formulation of light limitation could particularly favor particles that exhibit diel vertical migration.
+\texttt{I tried to avoid getting into the "light availability" and "attenuation" topic.
+However, I feel writing it all reviewers will not be to happy with the current "bright during night" implementation.
+Do you feel I might get away with it?
+Or should I try to implement it before I submit it because they will ask for it anyway?}
+A process we mostly ignore in our study is dormancy.
+Our organisms can survive for $28$ days in light limited waters.
+However, phytoplankton species have life stages in which they can remain dormant for a longer period of time and germinate again when they find themselves in more favorable waters \citep{ThomasAnderson1998}.
+In the process of choosing the light limitation threshold, we conducted sensitivity studies testing the effect of higher light budgets.
+We found that light budgets over 3 months begin to significantly increase the survivability of sinking organisms, when we crudely assume that they could still reproduce under these conditions.
+Whether dormancy plays a significant role in an environment where the river bed is continuously dredged is unknown.
+
+Another limitation in our modelling efforts is the lack of sub-grid-resolution structure on the shores.
+In our representation we assume perfectly flat surfaces with a median distance between nodes of approximately 60 \unit{m}. 
+This 'polished' model representation can lead to an underestimation of the retention success, since the surface area on which phytoplankton organisms can settle is underestimated.
+In reality, vegetation, rocks or other surface irregularities cause a larger surface area on which the phytoplankton organisms can settle.
+
+While our model does have a sedimentation and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury particles. 
+This masks potential losses due to particles being buried but also decreases resuspension times.
+
+Our results clearly suggest the importance of tidal flats and shallow areas along the river banks for the successful maintenance of primary production in the Elbe estuary. 
+However, their effect can currently not be quantified due to the lack of validation data.
+Chlorophyll data with a sufficient temporal and spacial resolution is only gathered in the center of the river.
+Future monitoring efforts should therefore also include data along the river shores on tidal flats or shore-to-shore to quantify the effect of potential future changes by dredging, diking or restoration attempts.
+
+For several decades, the annual average chlorophyll concentration in the Elbe estuary has been decreasing (data available at www.fgg-elbe.de/elbe-datenportal.html or see \citep{Hardenbicker2014,Schol2014}), while upstream concentrations do not show this effect.
+The reasons for this are not fully understood, but one possible reason is the increase in dredge activity.
+This increases the average turbidity and thus the aphotic depth, reducing the volume of water in which phytoplankton can grow.
+A large fraction of the phytoplankton measured upstream of Hamburg harbor consists of diatoms \citep{Muylaert1999}, which typically have negative buoyancy \citep{Passow1991}, making them particularly susceptible to sinking in light-limited waters.
+Our finding that sinking particles have a harder time surviving in the estuary supports this theory.
+
+Another mechanism that might in part explain the drop in phytoplankton concentration at the bathymetric jump, which is not yet explored in our model, is the phytoplankton stickiness.
+Phytoplankton, especially blooming one, has been shown to be sticky due to exudates \citep{Kiørboe1993,VanderLee2000,Dutz2005}. 
+Some phytoplankton also produce transparent exopolymer particles, which increase their stickiness to other particles \citep{Windler,DeBrouwer2005}.
+We suspect that this in combination with  higher turbidity induced by dredging results in losses due to plankton aggregates sticking to negatively buoyant suspended matter and subsequently sinking to the ground where they are starved of light.
+A future model study could create estimates on chlorophyll concentration losses caused by this effect.
+
+\conclusions  %% \conclusions[modified heading if necessary]
+In this study, we investigated the role of different possible retention strategies for drifting phytoplankton organisms to remain in an estuarine environment.
+As estuarine ecosystems are generally highly productive areas and of great value for anthropogenic use, maintaining their productivity is extremely important.
+However, whether and how drifting phytoplankton populations avoid being transported into unfavorable marine conditions has not been well understood.
+Our model simulations suggest that realistically high growth rates are sufficient for populations to prevent decline of estuarine populations.
+Moreover, buoyancy and strong diel vertical migration support retention within the estuary.
+The reason for this is that phytoplankton organisms with strong buoyancy or the potential to move upward in the water column are more likely to be transported to shallow areas near the river banks or tidal flats, where currents are slower and the chances of settlement under sufficiently light conditions to allow reproduction are higher than in deeper waters.
+These results illustrate the importance of shallow nearshore areas for the productivity of estuarine ecosystems.
+Our results also highlight the need for informed ecosystem-based management to avoid the degradation of estuarine ecosystems by dredging and diking activities.
+
+%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
+%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
+
+% \codeavailability{TEXT} %% use this section when having only software code available
+
+% \dataavailability{TEXT} %% use this section when having only data sets available
+
+\codedataavailability{Input data can be requested from Johannes Pein (johannes.pein@hereon.de).
+OceanTracker's source code is available at\url{github.com/oceantracker/oceantracker}.
+Model configuration and output is available at \url{doi.org/10.25592/uhhfdm.13235}} %% use this section when having data sets and software code available
+
+% \sampleavailability{TEXT} %% use this section when having geoscientific samples available
+
+% \videosupplement{TEXT} %% use this section when having video supplements available
+
+% \appendix
+
+% \section{}    %% Appendix A
+
+% \subsection{}     %% Appendix A1, A2, etc.
+
+% \noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).
+
+%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
+
+%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
+%% They will be correctly named automatically.
+
+%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
+%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:
+
+% \appendixfigures  %% needs to be added in front of appendix figures
+
+\appendixtables   %% needs to be added in front of appendix tables
+
+%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.
+
+\authorcontribution{LS and RV contributed to conception of the study.
+LS designed the studies details and organized the hydrodynamic data.
+RV provided the source code for OceanTracker.
+RS,LS improved on the original physical model and LS developed the biological model.
+LS performed the model simulations, post-processing, and visualization.
+LS wrote the draft of the manuscript.
+All authors contributed to manuscript revision, read, and approved the submitted version.}
+
+\competinginterests{The authors declare that they have no conflict of interest.} %% this section is mandatory even if you declare that no competing interests are present
+
+% \disclaimer{TEXT} %% optional section
+
+\begin{acknowledgements}
+We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data.
+Further, we thank Jana Hinners, Sina Remmers and Philipp Porada for providing helpful comments on the manuscript.
+This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) within the Research Training Group 2530: “Biota-mediated effects on Carbon cycling in Estuaries” (project number 407270017), contribution to Universität Hamburg and Leibniz-Institut für Gewässerökologie und Binnenfischerei (IGB) im Forschungsverbund Berlin e.V.
+\end{acknowledgements}
+
+%% REFERENCES
+
+%% The reference list is compiled as follows:
+
+% \begin{thebibliography}{}
+
+% \end{thebibliography}
+
+%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
+%% authors experienced with BibTeX only have to include the following two lines:
+%%
+\bibliographystyle{copernicus}
 \bibliography{references.bib}
-\bibliographystyle{unsrt}
+%%
+%% URLs and DOIs can be entered in your BibTeX file as:
+%%
+%% URL = {http://www.xyz.org/~jones/idx_g.htm}
+%% DOI = {10.5194/xyz}
+
+%% LITERATURE CITATIONS
+%%
+%% command                        & example result
+%% \citet{jones90}|               & Jones et al. (1990)
+%% \citep{jones90}|               & (Jones et al., 1990)
+%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
+%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
+%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
+%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
+%% \citeauthor{jones90}|          & Jones et al.
+%% \citeyear{jones90}|            & 1990
+
+%% FIGURES
+
+%% When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage
+%% between bibliography and first table and/or figure as well as between each table and/or figure.
+
+% The figure files should be labelled correctly with Arabic numerals (e.g. fig01.jpg, fig02.png).
+
+%% ONE-COLUMN FIGURES
+
+%%f
+%\begin{figure}[t]
+%\includegraphics[width=8.3cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure}
+%
+%%% TWO-COLUMN FIGURES
+%
+%%f
+%\begin{figure*}[t]
+%\includegraphics[width=12cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure*}
+%
+%
+%%% TABLES
+%%%
+%%% The different columns must be seperated with a & command and should
+%%% end with \\ to identify the column brake.
+%
+%%% ONE-COLUMN TABLE
+%
+%%t
+%\begin{table}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table}
+%
+%%% TWO-COLUMN TABLE
+%
+%%t
+%\begin{table*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table*}
+%
+%%% LANDSCAPE TABLE
+%
+%%t
+%\begin{sidewaystable*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{sidewaystable*}
+%
+%
+%%% MATHEMATICAL EXPRESSIONS
+%
+%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
+%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
+%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
+%%%
+%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
+%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
+%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
+%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
+%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
+%%% Vectors are identified in bold italic font using \vec{x}
+%%% Matrices are identified in bold roman font
+%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
+%%% The character * should not be applied as mutliplication sign
+%
+%
+%%% EQUATIONS
+%
+%%% Single-row equation
+%
+%\begin{equation}
+%
+%\end{equation}
+%
+%%% PHYSICAL UNITS
+%%%
+%%% Please use \unit{} and apply the exponential notation
 

### 2023-09-27 ms.tex
draft sent to ross v2

@@
 \begin{abstract}
-    Due to their role as primary producers, phytoplankton is essential for the proliferation of estuarine ecosystems.
-    However, as small drifting organisms, phytoplankton cannot resist currents.
-    Estuaries not only represent challenging habitats due to a strong salinity gradient, but also have net transport into oceans.
-    So far, little is known about how phytoplankton retain themselves in estuaries.
-    We present a new individual-based Lagrangian model of the Elbe estuary examining possible retention mechanisms of phytoplankton.
-In more detail, we investigated how reproduction, sinking and rising, as well as diel vertical migration might allow for population survival.
-    We find that rising favors the retention success, fast sinking cause the opposite.
+    Due to their role as primary producers, phytoplankton are essential to the productivity of estuarine ecosystems.
+    However, it is important to understand how these nearly passive organisms are able to persist within estuaries, when river inflow results in a net outflow to the ocean.
+    Estuaries are also representing challenging habitats due to a strong salinity gradient.
+    So far, little is known about how phytoplankton are able to be retained  within estuaries.
+    We present a new individual-based Lagrangian model of the Elbe estuary which examines possible retention mechanisms for phytoplankton.
+    Specifically, we investigated how reproduction, sinking and rising, as well as diel vertical migration may allow for populations to persist within the estuary. 
+    We find that vertial migration especially rising favors the retention, fast sinking does not.
     We further provide first estimates on outwashing losses.
-    Our simulations illustrate that shallow stretches of water along the river banks are essential for the long-term survival of phytoplankton populations.
-    These results give important indications for ecosystem-based management of estuaries.
+    Our simulations illustrate that riverbanks and tidal flats are essential for the long-term survival of phytoplankton populations, providing refuges from strong downstream currents.
+    These results contribute to the understanding needed to advance ecosystem-based management of estuaries.
+    \texttt{decide how to use "doubling time" and "growth rate" consistently}
+    \texttt{"Just as a superficial reading would say its easy to be retained with normal doubling times with growth rates much smaller than those explored.?"}
 \end{abstract}
@@
 
-Estuaries represent highly productive ecosystems in relation to their area and contribute to the global carbon cycle along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
+Though, Estuaries are highly productive ecosystems.
+Their relative small area, disproportionally contributes to the global carbon cycle, along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
 % (Phytoplankton primary production in the world’s estuarine-coastal ecosystems,Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment). 
-While they are heavily influenced by anthropogenic stressors such as diking, dredging, and fishing which can be seen as a stressor on the ecosystem, they are of tremendous importance for anthropogenic usage
+While they are heavily influenced by anthropogenic stressors such as diking, dredging, and fishing, they are of tremendous importance for anthropogenic usage 
 \citep{Jennerjahn2013, Brown2022, Wilson2002}.
-Estuaries present challenging, dynamic  of its residents due to the strong salinity gradient and net transport to the sea.
-Here, we explore how phytoplankton, drifting small primary producers that form the basis of estuarine food webs, can sustain themselves in such dynamic environments.
+Estuaries present challenging dynamic to their smallest residents, due to strong salinity gradient and a net transport to the ocean.
+Here, we explore how phytoplankton, drifting small primary producers that form the basis of estuarine food webs, can persist within such dynamic environments.
 % (Pressures, stresses, shocks and trends in estuarine ecosystems: An introduction and synthesis,
@@
 
-As for most ecosystems - their dynamics are strongly controlled by primary producers \citep{Chen2023}.
+Like most ecosystems - estuarine ecosystem dynamics are strongly controlled by primary producers, in particular phytoplankton \citep{Chen2023}.
 % (Advances in phytoplankton population ecology in the Pearl river estuary).
-Phytoplankton represent the base of aquatic ecosystems. 
 Apart from biofilm-forming phytoplankton which are attached to their substrate \citep{Cheah2022},
 %A methodological review on the characterization of microalgal biofilm and its  extracellular polymeric substances 
-the vast majority of phytoplankton organisms is drifting passively and cannot withstand currents.
+the vast majority of phytoplankton organisms drifts passively in currents, though they may be able to infulence their vertical movement.
 % The exception being biofilm-forming or benthic phytoplankton which are attached to their substrate (cite).
-With the estuary having a net outwards flow, we would expect phytoplankton to be moving downstream over time and to be washed out from limnic waters, via brackish to marine waters.
-Hence, the question arises how phytoplankton as the drifting base of estuarine food webs is able to maintain their population size without declining due to the transport into the open ocean.
-If we assume that the population is not exclusively maintained by upstream phytoplankton that is washed into the estuary, then there must be some sort of retention mechanism that enables a phytoplankton population to remain in the estuary.
+With the estuary having a net outwards flow, we would expect phytoplankton to be moving downstream over time and to be washed out from limnic waters, via brackish into marine waters.
+Hence, the question arises how phytoplankton, as the drifting base of estuarine food webs, are able to maintain their population size without declining due to the net transport into the open ocean.
+If we assume that the population is not exclusively maintained by  a self maintaining source population upstream, that is washed into the estuary, then there must be some sort of retention mechanism that enables a phytoplankton population to persist within the estuary.
 % While we suspect that such a mechanism is crucial for survival, it is currently not well understood due to the difficulty in either measuring or modeling such processes.
@@
 
-So far, different theories exist on how estuarine phytoplankton is able to maintain its position.
-Previous observational studies suggested several possible mechanisms that can allow for the retention of phytoplankton populations in estuarine systems - vertical migration in the form of sinking,  rising or diel migration, stickiness.
+So far, different theories exist on how estuarine phytoplankton populations are able to maintain their position.
+Previous observational studies suggested several possible mechanisms that could enable retention of phytoplankton populations within estuarine systems - vertical migration in the form of sinking,  rising or diel migration, stickiness.
 
-% \texttt{Comment on sinking and rising?}
-Diel vertical migration is a process where organisms move up and down in the water column in dependence of the sun phase.
-This movement can be favorable for retention as it allows particles to reduce the time in the faster downstream currents at the water surface.
+Diel vertical migration is a process where organisms move up and down in the water column in response to the sun.
+This movement favors retention by allowing plankton to reduce the time in the faster downstream currents at the water surface.
 However, this has only been demonstrated for larger motile estuarine organisms such as zooplankton \citep{Hall2015, Kimmerer2002, Crawford1991,Hall2011}.
@@
 % Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate, Flocculation kinetics and mechanisms of microalgae- and clay-containing suspensions in different microalgal growth phases).
-Yet, a study by \citep{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate their outwashing losses through reproduction increasing their abundance.
+Yet, a study by \citep{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate for outwashing losses through reproduction increasing their abundance.
 However, this also implies that the growing part of the population is somehow retaining their position.
@@
 Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
-In \citep{Simons2006,Kimmerer2014} they use a particle tracking model also known as Lagrangian model to study Zooplankton retention.
-\citep{Simons2006} examined the dispersal and flush time of mussel larvae in the St. Lawrence Estuary while \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
-They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention tactic.
+In \citep{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
+\citep{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
+They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention strategy.
 However, they did so by ignoring key processes like reproduction, mortality, and stranding or sedimentation processes.
@@
 
-Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the ground and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}
+Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the riverbed and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}`'
 % (Species-specific sedimentation and sinking velocities of diatoms).
-Studies also found sticky compounds in phytoplankton aggregates that are suspected to let them stick to suspended particles making them sink to the ground or sticking to their surroundings aiding retention \citep{Kiørboe1993,VanderLee2000}.
-% \texttt{Shall I remove sticking from intro as it is not tested in my model?}
+Studies also found sticky compounds in phytoplankton aggregates that are suspected to allow them stick to suspended particles, enabling them to sink to the riverbed or sticking to their surroundings aiding retention \citep{Kiørboe1993,VanderLee2000}.
 
 In summary, different retention mechanisms have been observed or examined in modeling studies.
-However, they did so in either in isolation in the case of the observational studies or with major simplifications for the modeling studies.
-There is currently a lack of theoretical studies that allow a more comprehensive overview of possible retention mechanisms.
+However, they did so in either in isolation in the case of the observational studies or with major simplifications in the modeling studies.
+% \textit{
+%     There is currently a lack of theoretical studies that allow a more comprehensive overview into the interplay of vertical migration and reproduction in combination with settling and stranding as retention mechanisms.
+% }
 
-Here, we are exploring possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
+Here, we explore possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
 It is located in the north of Germany and flows into the North Sea.
@@
 Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
-This causes the main channel to experience a sudden jump in bathymetry from approx. 5 \unit{m} at border of the city to approx. 20 \unit{m} in the harbor (see fig. \ref{fig:bathymetry}). 
-This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton biomass causing an increase in oxygen depletion and high ammonium remineralization after the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
+To create port access the main channel experience a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the harbor and downstream (see fig.
+\ref{fig:bathymetry}).
+This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton , by causing an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
 Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
@@
 \begin{figure}
-    \includegraphics[width=8.3cm]{figures/harbor.png}
+    \includegraphics[width=8.3cm]{harbor_small.png}
     \caption{Bathymetry of the Elbe model around Hamburg.
@@
 %\subsubsection{Lagrangian model}
-
 In our study we use a Lagrangian approach with the particle tracking model OceanTracker \citep{Vennell2021}.
-While particle tracking on unstructured grids has been relatively computationally expensive until recently \citep{Vennell2021}, it offers several advantages.
-Firstly, it allows us to reuse already computed hydrodynamic models to model tracer-like objects.
+While off-line particle tracking on unstructured grids has been relatively computationally expensive until recently \citep{Vennell2021}, it offers several advantages.
+Firstly, it allows us to reuse computationally expensive hydrodynamic models to model tracer-like objects.
 This is overall much faster than recalculating the advection-diffusion-equation in an Eularian model.
 Secondly, because we are simulating individually particles we are able to observe their tracks.
-This makes the interpretation of our results more intuitive and allows us to include individual based properties and processes that can only be parametrized in Eulerian models.
+This makes the interpretation of our results more intuitive and allows us to include individual based properties and processes that can not or only indirectly be represented in Eulerian models.
 
@@
 The model provides us with a node-based mesh containing a range of information such as water velocity, salinity, water level and dispersion.
-The year represented in that dataset is 2012 with a temporal resolution of 1 hour and a dynamically varying spacial resolution with a median distance between nodes of approx. 60 \unit{m}.
+The year represented in that dataset is 2012 with a temporal resolution of 1 hour and a dynamically varying spacial resolution with distance between nodes ranging from 5 to 1400 \unit{m} with a median distance of approximately 75 \unit{m}.
 
 \begin{figure}
-    \includegraphics[width=12cm]{figures/dfg_foto_wettbewerbg_laurin_legendless_steidle_small.png}
+    \includegraphics[width=12cm]{model_domain.png}
     \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left.
-The gray outline marks the edge of the model domain.
-Blue and green dots show an example snapshot of a phytoplankton distribution in the model.
-Blue represents floating, green settled particles.
+The black outline marks the edge of the model domain.
+Blue and green dots show an example snapshot of a fraction of the phytoplankton in the model.
+The location of the initial release is shown in red.
+Blue represents floating, green particles stranded by the receding tide.
+The red area is the initial release location.
 The background map has been provided by OpenStreetMap.}
@@
 %\subsubsection{Particle based processes}
-
 We add a set of biological features on top of the otherwise inanimate organism.
-These features include reproduction and mortality, vertical movement in form of sinking, rising or diel vertical migration, stranding, and sedimentation.
+These features include reproduction and mortality, vertical movement in form of sinking, rising or diel vertical migration, stranding, and settling on the riverbed.
 
 Reproduction is represented as a chance of creating copies of themselves.
+\textit{
+    This is a novel feature applied in OceanTracker that has not been included in any previous Lagrangian model of this kind.
+    OcenTrackers recent advances in computational efficiency and buffer handling \citep{Vennell2021} make it possible to simulate a large number of particles over a long period of time for the first time.
+}
 We perform several simulations for a range of reproduction rates that are constant over the lifetime of the particle and the chance to reproduce is evaluated every minute.
+\textit{
+    While a fixed reproduction rate is a simplification that does not allow for an accurate simulation of the population dynamics of a particular species, it does allow us to investigate the general mechanisms that enable plankton retention.
+}
 
 Mortality is induced either by high salinity, when they dry-out while stranded, or due to long term light limitation.
-When particles are exposed to high salinity water above $20$PSU, a mortality probability of $1\%$ per minute is imposed.
-This threshold is chosen based on the salinity tolerances presented in \citep{vonAlvensleben2016}.
-This is only an approximation and many phytoplankton species deviate from this number.
-However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for several hours and are assumed not to return again through this isohaline.
-We consider particles that are stranded outside the water and lie dry for more than $7$ consecutive days to be dead and remove them.
-Particles are also expected to be dead and are removed if they are light-limited for $28$ days \citep{Walter2017}.
+When particles are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is imposed.
+This threshold is chosen based on a range of the salinity tolerances of estuarine phytoplankton species presented in \citep{vonAlvensleben2016}.
+This is only an approximation and salinity tolerances many estuarine phytoplankton species deviate from this.
+However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
+\textit{
+    Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
+    Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
+}
+We consider particles that are stranded outside the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
+Particles also die if they are light-limited for $28$ days \citep{Walter2017}.
 They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
@@
 
-We study the effect of different patterns of vertical motion.
-The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton organisms.
-This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the aggregate in which the organism is embedded.
+investigate the effect of different patterns of vertical motion.
+The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton.
+This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which the live.
 For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
 The second mode of vertical motion is diel vertical migration.
-Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night using the same velocity range.
+Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night within the same velocity range.
 
 % A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
-We include a sedimentation and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
-Particles become stranded when the current grid cell falls dry. 
-They are not allowed to disperse into dry cells.
-A grid cell is considered dry if at least one of the nodes of the cells' triangle is considered dry by the SCHISM wet flag.
-Once this cell is rewettet all stranded particles resuspend.
+We include a settling and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
+Particles become stranded when the current grid cell becomes dry. 
+They are not allowed to move from wet cells to dry cells, by the random walk dispersion applied to all particles.
+A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
+Once this cell is rewettet all stranded particles resuspend and are able to move again.
 Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
@@
 This allows us for example to compare successfully retained particles (older than three months) unsuccessfully retained particles (dead after less than three months).
-These observables are measured every 12 hours starting at midnight.
+These observables are recorded every 12 hours starting at midnight.
 
@@
 Model simulations and visualizations were performed in Python making heavy use of Numba, a LLVM-based Python JIT compiler \citep{Lam2015} to drastically speed up the simulations \citep{Vennell2021}.
-Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 10 \unit{s}.
+Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 60 \unit{s}.
 Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates.
@@
 
-In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm s^{-1}} in $2$ \unit{mm s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
-Each vertical velocity is examined for a range of different growth rates with effective population doubling times ranging from 772 to 77 days with a logarithmic scaling.
+In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
+Each vertical velocity is examined for a range of different growth rates with effective population doubling times ranging from 77 to 772 days with a logarithmic scaling.
 Hence, a total of 187 different scenarios are tested.
@@
 
-In both set of experiments, we release the studied phytoplankton population at the beginning of the year homogeneously distributed in a volume covering the full water column at the weir in Geesthacht and examine how the population distributes itself over the estuary and whether it is able to maintain its population size over time.
+\textit{
+    In both sets of experiments, we release 10,000 individuals representing the studied phytoplankton population at the beginning of the year.
+    This results in over 1 billion individual particles simulated for each case with approximately 1 million particles active simultaneously counted over all cases over 500,000 time steps.
+}
+The initial population is homogeneously distributed in a volume covering the full water column at the weir in Geesthacht (see fig.
+\ref{fig:model_domain}) and examine how the population distributes itself over the estuary and whether it is able to maintain its population size over time.
 Conceptually, we consider a population to be successfully retained if it is showing long term growth. 
@@
 Population size changes are measured at the end of the year relative to the population size after this initial spin-up time.
+\texttt{For Ross: I do not consider runs longer then a year due to a lack of hydrodynamic data.
+Do you recommend to comment on it?
+I feel like this is not interesting enough as I do not expect the dynamics to fundamentally change if ie e.g.
+would release them during summer and then waited for a full year.}
+
+\textit{
+    Computations were performed on the supercomputer Mistral at the German Climate Computing Center (DKRZ) in Hamburg, Germany.
+    The simulations were performed on a compute node with two  Intel Xeon E5-2680 v3 12-core processor (Haswell) and 128 GB of RAM with a total run time of approximately 4.5 hours.
+}
 
@@
 Positive population changes represent a retention success (green) while negative population changes represent a loss of the population (brown).
-        \texttt{I think I'll redo the simulation and shift the pop doubling time slightly to the left}
         }
@@
 Our simulations show that the population is able to successfully retains itself under certain conditions.
+\textit{
+    Passively drifting particles are able to sustain themselves in the estuary if they have a growth rate that doubles their population size within approximately 4 months.
+}
 For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and a higher growth rate is more beneficial for retention success than a downward oriented velocity (sinking) and lower growth rates.
@@
 
-For the case of the diel vertical migration we see a equal or higher retention success compared to no vertical velocity with velocities in the range of 4-10 \unit{mm s^{-1}} vertical velocities but shows to me less successful than the monodirectional movement for all cases.
-A vertical velocity of 2 \unit{mm s^{-1}} is less successful then no migration.
-Most importantly, none of the diel migration scenarios are improving the retention success compared to passively drifting organisms.
+For the case of the diel vertical migration in the velocity range of 4 to 10 \unit{mm\;s^{-1}} we see an equal or higher retention success compared to the case with no vertical migration.
+A diel velocity of 2 \unit{mm\;s^{-1}} is less successful than no migration.
+Most importantly, none of the diel migration scenarios improve the retention success, when compared to passively drifting organisms.
 
@@
 \begin{figure}
-    \includegraphics[width=8.3cm]{figures/age_hexbin.png}
+    \includegraphics[width=12cm]{age_hexbin.png}
     \caption[]{
@@
         Yellow areas are mainly found along the river banks in shallow waters or tidal flats.
-        The largest areas are the Mühlenberger Loch (a) and the Haseldorfer Binnenelbe (b).
-        Smaller areas can also be found at Wedeler Marsch (c), at the mouth of Wischhafener Süderelbe (d), Ruthenstrom, and Stör (e), and at Nordkedding (g) and Neufelder Marsch (h).
-        \texttt{I am aware that the unit of age (days) is still missing. This figure will be modified slightly before publication.}
+        The important areas are Mühlenberger Loch (a), Wedeler Marsch (b) Haseldorfer Binnenelbe (c), Asseler- and Schwarztonnensand (d), at the mouth of Wischhafener Süderelbe (f), , and Stör (e), and at Nordkedding (g) and Neufelder Marsch (h).
     }
@@
 
-In this study, we investigated different strategies to explain how phytoplankton populations are able to maintain their growth in estuaries while constantly being at risk to be transported into the open ocean.
+In this study, we investigated different strategies to explain how phytoplankton populations are able to maintain their population growth in estuaries while constantly being at risk to be transported into the open ocean.
 % We find that reproduction rates in the range of XX-XX are sufficient for phytoplankton organisms to survive in the estuary under certain conditions.
@@
 
-The limit for population doubling times that we found to be necessary for survival are of the order of around 200 days.
+The limit for population doubling times that we found to be necessary for survival for passively drifting plankton is of the order of around 4 months.
 Doubling times typically realized in natures are of the order of a few days which is two magnitudes small then those that we found necessary in our model \citep{Koch2004,Wirtz2011}.
@@
 Our results suggest that shallow areas are very important for maintaining the estuary phytoplankton population.
-Plankton that lives in this area falls dry on the shore on a regular basis due to the tides. 
-Therefore, they do not move for much of the day.
-We further see that positively buoyant plankton is more successful in retaining themselves. 
+Plankton that consistently finds itself in this area falls dry on the shore on a regular basis due to the tides.
+Therefore, they do not move during a tidal cycle.
+We further see that positively buoyant plankton are more successful in retaining themselves. 
 This is probably because they are more likely to be transported high up on the river bank where the water is less likely to reach them.
@@
 However, we found that faster sinking particles are less successful in retaining themselves.
-Sinking velocities of less then $2 \unit{mm s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
-Sinking particles have a reduced downstream velocity because they sink to the river bed, where they move more slowly, or are stationary if they are temporarily settled or stranded.
-In addition, the deeper water layer of the Elbe typically has a higher upstream velocity than the upper water column due to temperature-induced density stratification \citep{Pein2021}.
+Sinking velocities of less then $2 \unit{mm\;s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
+Sinking particles have a reduced downstream velocity because find themselves either settled on the riverbed not moving at all or close to the bed where the average downstream velocity is lower.
+In addition, the deeper water layers of the Elbe have on average a lower downstream velocity than the upper water column or moves upstream  due to temperature-induced density stratification \citep{Pein2021}.
 Nevertheless, buoyant particles were more successful in their retention in our simulations.
@@
 When the upwards diel migration coincides with high tide, particles are more likely to be stranded far out on the shore, reducing their risk of being washed out quickly.
-The higher the upward velocities, the greater the chance of being in the farthest part of the water body during high tide.
+The higher the upward velocities, the greater the chance of being at the waterline during high tide.
 However, because they are sinking for half of the day they also tend to be light limited more frequent than positively buoyant particles.
@@
 A more realistic formulation of light limitation could particularly favor particles that exhibit diel vertical migration.
-\texttt{I tried to avoid getting into the "light availability" and "attenuation" topic.
-However, I feel writing it all reviewers will not be to happy with the current "bright during night" implementation.
-Do you feel I might get away with it?
-Or should I try to implement it before I submit it because they will ask for it anyway?}
 A process we mostly ignore in our study is dormancy.
@@
 
-While our model does have a sedimentation and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury particles. 
+While our model does have a settling and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury particles. 
 This masks potential losses due to particles being buried but also decreases resuspension times.
@@
 Future monitoring efforts should therefore also include data along the river shores on tidal flats or shore-to-shore to quantify the effect of potential future changes by dredging, diking or restoration attempts.
+
+\textit{
+    Frequently stranded plankton have been shown to be essential to the survival of populations in our model.
+    However, data on their ability to survive under these conditions are scarce.
+    Our results suggest that these conditions may be as important as their ability to quickly regrow under more favorable conditions, and we suggest further research on plankton survivability when stranded.
+}
 
@@
 In this study, we investigated the role of different possible retention strategies for drifting phytoplankton organisms to remain in an estuarine environment.
-As estuarine ecosystems are generally highly productive areas and of great value for anthropogenic use, maintaining their productivity is extremely important.
+% As estuarine ecosystems are generally highly productive areas and of great value for anthropogenic use, maintaining their productivity is extremely important.
 However, whether and how drifting phytoplankton populations avoid being transported into unfavorable marine conditions has not been well understood.
 Our model simulations suggest that realistically high growth rates are sufficient for populations to prevent decline of estuarine populations.
-Moreover, buoyancy and strong diel vertical migration support retention within the estuary.
+Moreover, buoyancy and strong diel vertical migration enhance retention within the estuary.
 The reason for this is that phytoplankton organisms with strong buoyancy or the potential to move upward in the water column are more likely to be transported to shallow areas near the river banks or tidal flats, where currents are slower and the chances of settlement under sufficiently light conditions to allow reproduction are higher than in deeper waters.
-These results illustrate the importance of shallow nearshore areas for the productivity of estuarine ecosystems.
+These results illustrate the importance of shallow nearshore areas to allowing the productivity of estuarine ecosystems to persist.
 Our results also highlight the need for informed ecosystem-based management to avoid the degradation of estuarine ecosystems by dredging and diking activities.

### 2023-09-27 ms.tex
send to ross v3

@@
     Specifically, we investigated how reproduction, sinking and rising, as well as diel vertical migration may allow for populations to persist within the estuary. 
-    We find that vertial migration especially rising favors the retention, fast sinking does not.
+    We find that vertical migration especially rising favors the retention, fast sinking does not.
     We further provide first estimates on outwashing losses.
@@
     These results contribute to the understanding needed to advance ecosystem-based management of estuaries.
-    \texttt{decide how to use "doubling time" and "growth rate" consistently}
-    \texttt{"Just as a superficial reading would say its easy to be retained with normal doubling times with growth rates much smaller than those explored.?"}
 \end{abstract}
@@
 %A methodological review on the characterization of microalgal biofilm and its  extracellular polymeric substances 
-the vast majority of phytoplankton organisms drifts passively in currents, though they may be able to infulence their vertical movement.
+the vast majority of phytoplankton organisms drifts passively in currents, though they may be able to influence their vertical movement.
 % The exception being biofilm-forming or benthic phytoplankton which are attached to their substrate (cite).
@@
 Reproduction is represented as a chance of creating copies of themselves.
-\textit{
 This is a novel feature applied in OceanTracker that has not been included in any previous Lagrangian model of this kind.
-    OcenTrackers recent advances in computational efficiency and buffer handling \citep{Vennell2021} make it possible to simulate a large number of particles over a long period of time for the first time.
-}
+OceanTrackers recent advances in computational efficiency \citep{Vennell2021} and buffer handling make it possible to simulate a large number of particles over a long period of time for the first time on unstructured grids.
 We perform several simulations for a range of reproduction rates that are constant over the lifetime of the particle and the chance to reproduce is evaluated every minute.
-\textit{
-    While a fixed reproduction rate is a simplification that does not allow for an accurate simulation of the population dynamics of a particular species, it does allow us to investigate the general mechanisms that enable plankton retention.
-}
+While a fixed reproduction rate is a simplification that does not allow for more realistic simulation of the population dynamics of a particular species, it does allow us to investigate the general mechanisms that enable plankton retention.
 
@@
 However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
-\textit{
 Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
 Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
-}
 We consider particles that are stranded outside the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
@@
 based on \citep{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
-The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column.
+The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column from the hydrodynamic model output.
 
@@
 %\subsubsection{Numerics}
-Model simulations and visualizations were performed in Python making heavy use of Numba, a LLVM-based Python JIT compiler \citep{Lam2015} to drastically speed up the simulations \citep{Vennell2021}.
+Model simulations and visualizations were performed in Python making heavy use of Numba, a LLVM-based Python JIT compiler \citep{Lam2015} to significantly speed up the simulations \citep{Vennell2021}.
 Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 60 \unit{s}.
-Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates.
+Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates, with the exception of water velocity in the bottom model cell, where logarithmic vertical interpolation is used.
 
@@
 In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
-Each vertical velocity is examined for a range of different growth rates with effective population doubling times ranging from 77 to 772 days with a logarithmic scaling.
+\textit{
+    Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 77 to 772 days with a logarithmic scaling.
+    In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
+}
 Hence, a total of 187 different scenarios are tested.
-In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and growth rates.
+In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and reproduction rates.
 
-\textit{
 In both sets of experiments, we release 10,000 individuals representing the studied phytoplankton population at the beginning of the year.
 This results in over 1 billion individual particles simulated for each case with approximately 1 million particles active simultaneously counted over all cases over 500,000 time steps.
-}
 The initial population is homogeneously distributed in a volume covering the full water column at the weir in Geesthacht (see fig.
@@
 Population size changes are measured at the end of the year relative to the population size after this initial spin-up time.
-\texttt{For Ross: I do not consider runs longer then a year due to a lack of hydrodynamic data.
-Do you recommend to comment on it?
-I feel like this is not interesting enough as I do not expect the dynamics to fundamentally change if ie e.g.
-would release them during summer and then waited for a full year.}
 
-\textit{
 Computations were performed on the supercomputer Mistral at the German Climate Computing Center (DKRZ) in Hamburg, Germany.
 The simulations were performed on a compute node with two  Intel Xeon E5-2680 v3 12-core processor (Haswell) and 128 GB of RAM with a total run time of approximately 4.5 hours.
-}
 
@@
 \begin{figure}
-    \includegraphics[width=8.3cm]{figures/retention_success_combined.png}
+    \includegraphics[width=8.3cm]{retention_success_combined.png}
     \caption[]{
@@
 Positive population changes represent a retention success (green) while negative population changes represent a loss of the population (brown).
+        \textit{
+            The vertical black lines indicate the boundary between successfully and unsuccessfully retained scenarios.
+        }
+        \texttt{
+            Still playing with the vertical lines. Not happy with the current aesthetic.
+        }
     }
@@
 \ref{fig:retention_success}b shows the results for the diel vertical migration scenarios.
-Each pixel in the heatmap represents a simulation with a specific combination of vertical velocity and growth rate.
+Each pixel in the heatmap represents a simulation with a specific combination of vertical velocity and reproduction rate.
 The coloring indicates the relative population change after one year.
@@
 In this case, the losses are equal to the growth.
-Therefore, we can use the growth rate as an estimate for the total relative losses due to downstream transport, drying out while being stranded, and light starvation.
+Therefore, we can use the reproduction rate as an estimate for the total relative losses due to downstream transport, drying out while being stranded, and light starvation.
 
 Our simulations show that the population is able to successfully retains itself under certain conditions.
+Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 4 months.
 \textit{
-    Passively drifting particles are able to sustain themselves in the estuary if they have a growth rate that doubles their population size within approximately 4 months.
+    Note that the growth rates realized in nature may vary from this value due to e.g.
+nutrient or temperature limitations.
+The reproduction thresholds should be interpreted as an upper bound rather than an accurate estimate of the growth rate.
 }
-For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and a higher growth rate is more beneficial for retention success than a downward oriented velocity (sinking) and lower growth rates.
+For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and a higher reproduction rate is more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
 As expected, simulations in which the reproduction is set to zero do not show any retention success.
-While it is easy to understand that high growth rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.
+While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.
 
@@
 
-\subsection{Spatial analysis}
+\subsection{Spatial factors}
 
@@
 To further investigate the reasons for the positive effect of buoyancy and the importance of shallow waters and tidal flats, we repeated the first set of simulations and disabled the reproduction of settled and stranded particles.
-Under these conditions, populations were unable to retain themselves in the estuary, regardless of their vertical velocity and growth rate indicating that tidal flats are essential for the survival of the population.
+Under these conditions, populations were unable to retain themselves in the estuary, regardless of their vertical velocity and reproduction rate indicating that tidal flats are essential for the survival of the population.
 
@@
 Doubling times typically realized in natures are of the order of a few days which is two magnitudes small then those that we found necessary in our model \citep{Koch2004,Wirtz2011}.
-The low growth rates required for successful retention demonstrate that our model is also meaningful under more realistic environmental conditions, for example if maximum growth rates cannot be reached due to nutrient or temperature limitation.
+The low reproduction rates required for successful retention demonstrate that our model is also meaningful under more realistic environmental conditions, for example if maximum growth rates cannot be reached due to nutrient or temperature limitations.
 
 Our results suggest that shallow areas are very important for maintaining the estuary phytoplankton population.
-Plankton that consistently finds itself in this area falls dry on the shore on a regular basis due to the tides.
-Therefore, they do not move during a tidal cycle.
+\textit{Plankton that consistently finds itself in areas that are dry due to the tides will regularly become stranded and therefore not move for much of the tidal cycle.}
 We further see that positively buoyant plankton are more successful in retaining themselves. 
@@
 However, we found that faster sinking particles are less successful in retaining themselves.
-Sinking velocities of less then $2 \unit{mm\;s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
+Sinking velocities of less than $2 \unit{mm\;s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
 Sinking particles have a reduced downstream velocity because find themselves either settled on the riverbed not moving at all or close to the bed where the average downstream velocity is lower.
-In addition, the deeper water layers of the Elbe have on average a lower downstream velocity than the upper water column or moves upstream  due to temperature-induced density stratification \citep{Pein2021}.
+In addition, the deeper water layers of the Elbe have on average a lower downstream velocity than the upper water column or move upstream  due to temperature-induced density stratification \citep{Pein2021}.
 Nevertheless, buoyant particles were more successful in their retention in our simulations.
@@
 
+\textit{
+    Our hydrodynamics data set was limited to the year 2012.
+    Therefore, we were not able to study different release times with the same methodology.
+    While we do not expect the general dynamics to change, future research could examine the effect of varying discharge throughout the seasons on retention.
+}
+
 While our model does have a settling and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury particles. 
@@
 
-\textit{
 Frequently stranded plankton have been shown to be essential to the survival of populations in our model.
@@
 Our results suggest that these conditions may be as important as their ability to quickly regrow under more favorable conditions, and we suggest further research on plankton survivability when stranded.
-}
 

### 2023-10-04 ms.tex
submitted

@@
 
-Though, Estuaries are highly productive ecosystems.
-Their relative small area, disproportionally contributes to the global carbon cycle, along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
+Estuaries are highly productive ecosystems.
+Their relative small area disproportionally contributes to the global carbon cycle, along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
 % (Phytoplankton primary production in the world’s estuarine-coastal ecosystems,Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment). 
@@
 \citep{Jennerjahn2013, Brown2022, Wilson2002}.
-Estuaries present challenging dynamic to their smallest residents, due to strong salinity gradient and a net transport to the ocean.
+Estuaries present challenging dynamics to their smallest residents, due to strong salinity gradient and a net transport to the ocean.
 Here, we explore how phytoplankton, drifting small primary producers that form the basis of estuarine food webs, can persist within such dynamic environments.
@@
 % Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate, Flocculation kinetics and mechanisms of microalgae- and clay-containing suspensions in different microalgal growth phases).
-Yet, a study by \citep{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate for outwashing losses through reproduction increasing their abundance.
+Yet, a study by \citet{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate for outwashing losses through reproduction increasing their abundance.
 However, this also implies that the growing part of the population is somehow retaining their position.
@@
 
-Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic due to their complex bathymetry \citep{Michael2016, Fringer2019}.
+Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic impacts due to their complex bathymetry \citep{Michael2016, Fringer2019}.
 % (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
@@
 Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
-In \citep{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
-\citep{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
+In \citet{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
+\citet{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while() \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
 They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention strategy.
@@
 
-Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the riverbed and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}`'
+Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the riverbed and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}.
 % (Species-specific sedimentation and sinking velocities of diatoms).
@@
 In summary, different retention mechanisms have been observed or examined in modeling studies.
-However, they did so in either in isolation in the case of the observational studies or with major simplifications in the modeling studies.
-% \textit{
-%     There is currently a lack of theoretical studies that allow a more comprehensive overview into the interplay of vertical migration and reproduction in combination with settling and stranding as retention mechanisms.
-% }
+However, they did so either in isolation in the case of observational studies or with major simplifications in the modeling studies.
+There is currently a lack of theoretical studies that allow for a more comprehensive overview into the interplay of vertical migration and reproduction in combination with settling and stranding as retention mechanisms.
 
@@
 \ref{fig:bathymetry}).
-This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton , by causing an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
+This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton, resulting in an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
 Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
@@
 While the Lagrangian model simulated the movement of the inanimate organism, we included key phytoplankton features such as reproduction and mortality, sinking and rising, as well as diel vertical migration.
-% \textit{and dormancy?}
 Using this model, we investigate under which conditions phytoplankton retention can be reproduced.
@@
 \begin{figure}
-    \includegraphics[width=12cm]{model_domain.png}
+    \includegraphics[width=12cm]{model_domain_wcopyright.png}
     \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left.
@@
 They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
-A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by $1$ minute, while the opposite applies when they are above $1 \unit{m}$.
+A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by one minute, while the opposite applies when they are above $1 \unit{m}$.
 Children of light-limited parents inherit the remaining light budget of their parents.
@@
 
-investigate the effect of different patterns of vertical motion.
+We investigate the effect of different patterns of vertical motion.
 The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton.
-This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which the live.
+This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which they live.
 For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
@@
 A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
-Once this cell is rewettet all stranded particles resuspend and are able to move again.
+Once this cell is rewetted all stranded particles resuspend and are able to move again.
 Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
 The velocity profile in the bottom layer, or log layer, is calculated by
-% centered equation
 \begin{equation}
@@
 \end{equation}
-based on \citep{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
+based on \citet{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
 The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column from the hydrodynamic model output.
@@
 In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
-\textit{
-    Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 77 to 772 days with a logarithmic scaling.
+Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 40 to 404 days with a logarithmic scaling.
 In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
-}
 Hence, a total of 187 different scenarios are tested.
@@
 Positive population changes represent a retention success (green) while negative population changes represent a loss of the population (brown).
-        \textit{
         The vertical black lines indicate the boundary between successfully and unsuccessfully retained scenarios.
-        }
-        \texttt{
-            Still playing with the vertical lines. Not happy with the current aesthetic.
-        }
     }
@@
 The results of the retention experiments are visualized as heatmap in fig. \ref{fig:retention_success}.
-Fig.
-\ref{fig:retention_success}a shows the results for the monodirectional vertical migration scenarios i.e.
-constant sinking or rising.
+Fig. \ref{fig:retention_success}a shows the results for the monodirectional vertical migration scenarios i.e. constant sinking or rising. 
 Fig.
@@
 Our simulations show that the population is able to successfully retains itself under certain conditions.
-Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 4 months.
-\textit{
+Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 3 months.
 Note that the growth rates realized in nature may vary from this value due to e.g.
@@
 The reproduction thresholds should be interpreted as an upper bound rather than an accurate estimate of the growth rate.
-}
-For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and a higher reproduction rate is more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
+
+For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and higher reproduction rates are more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
 As expected, simulations in which the reproduction is set to zero do not show any retention success.
@@
 \begin{figure}
-    \includegraphics[width=8.3cm]{figures/retention_boxplot.png}
-    \caption[]{Box and violin plot showing the vertical distribution of particles.
+    \includegraphics[width=8.3cm]{retention_boxplot.png}
+    \caption[]{Box and violin plot showing the vertical distribution of particles that are passively drifting.
 Short-living are those younger than 3 months and long-living all those older than that.
-Depth is measure relative to the current water surface with positive numbers being above the water surface i.e.
+Depth is measured relative to the current water surface with positive numbers being above the water surface i.e.
 stranded on shore.}
@@
 In this study, we investigated different strategies to explain how phytoplankton populations are able to maintain their population growth in estuaries while constantly being at risk to be transported into the open ocean.
-% We find that reproduction rates in the range of XX-XX are sufficient for phytoplankton organisms to survive in the estuary under certain conditions.
-% In addition, we found that buoyancy aids the retention success, whereas sinking decreases the retention success.
-Finally, especially shallow areas near the river banks and tidal flats represent important habitats for phytoplankton organisms and allow them to survive, whereas deeper areas lead to a faster transport of estuarine phytoplankton into the marine waters.
 
@@
 Our results suggest that shallow areas are very important for maintaining the estuary phytoplankton population.
-\textit{Plankton that consistently finds itself in areas that are dry due to the tides will regularly become stranded and therefore not move for much of the tidal cycle.}
+Plankton that consistently finds itself in areas that are dry due to the tides will regularly become stranded and therefore not move for much of the tidal cycle.
 We further see that positively buoyant plankton are more successful in retaining themselves. 
 This is probably because they are more likely to be transported high up on the river bank where the water is less likely to reach them.
-This effect is emphasized in flatter regions as the distance between the wash margin and constantly flooded areas is larger increasing the chance of settlement or them stranding again.
+This effect is emphasized in flatter regions as the distance between the wash margin and constantly flooded areas is larger, increasing the chance of settlement or them stranding again.
 
@@
 The low chance of survival in the estuary for sinking particles might be explained by their light limitation in deeper waters.
-We expected particles to die if they are exposed to dark conditions for more than 2 weeks.
+We expected particles to die if they are exposed to dark conditions for more than two weeks.
 Thus, sinking particles have a disadvantage to buoyant particles since they are more likely to become light limited and eventually die.
@@
 Future modeling efforts could couple the Lagrangian model to a Eulerian model that disperses changes in concentrations fields by biotic activity throughout the model domain.
-However, this would have drastically increased both developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
+However, this would have drastically increased both, developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
 The key draw back of this is that growth rates could only be modeled as a constant rate in the current model description, similar to "ad libitum" experiments. 
@@
 
-\textit{
 Our hydrodynamics data set was limited to the year 2012.
@@
 While we do not expect the general dynamics to change, future research could examine the effect of varying discharge throughout the seasons on retention.
-}
 
@@
 This masks potential losses due to particles being buried but also decreases resuspension times.
+
+\medskip
 
@@
 \begin{acknowledgements}
-We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data.
-Further, we thank Jana Hinners, Sina Remmers and Philipp Porada for providing helpful comments on the manuscript.
+We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data and Jana Hinners for her guidance through the project.
+Further, we thank, Sina Remmers and Philipp Porada for providing helpful comments on the manuscript.
 This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) within the Research Training Group 2530: “Biota-mediated effects on Carbon cycling in Estuaries” (project number 407270017), contribution to Universität Hamburg and Leibniz-Institut für Gewässerökologie und Binnenfischerei (IGB) im Forschungsverbund Berlin e.V.

### 2023-11-29 ms.tex
first review edit, to proof read for ross

@@
 
-\affil[1]{Universität Hamburg, Olbersweg 24, 22767 Hamburg, Germany}
+\affil[1]{Institute of Marine Ecosystem and Fishery Science, Universität Hamburg, Olbersweg 24, 22767 Hamburg, Germany}
 \affil[2]{Cawthron Institute, 98 Halifax Street East,
@@
 % The future of coastal and estuarine modeling: Findings from a workshop)
-Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
+Nevertheless, there are sophisticated estuarine models that are able to reproduce the complex dynamics of estuaries reasonably well.
+This includes currents and water levels on the physical side, but also chlorophyll concentrations and other biologically driven properties \citep{Pein2021, Schol2014}.
+However, these are Eulerian models.
+This means that they are based on a fixed grid and calculate the concentration of a tracer, such as phytoplankton, at each grid cell.
+This makes it difficult to study concepts such as retention times, as they lack temporal consistency, meaning that the life history and trajectory of a phytoplankton cell cannot be tracked.
+Previous modeling studies have attempted to overcome this problem using a Lagrangian approach.
+A Lagrangian model does not try to track e.g. concentrations at fixed positions, but rather follows the motion of individual particles that can be used to represent e.g. water parcels or organisms.
 In \citet{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
@@
 Most notably diking to restrain it to a narrow channel and dredging to improve access to the Hamburg harbor.
-Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
-To create port access the main channel experience a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the harbor and downstream (see fig.
+Unlike other major European ports, the port of Hamburg is located far inwards roughly 100 \unit{km} behind the coastline.
+To create port access the main channel is dredged and experiences a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the port and downstream (see fig.
 \ref{fig:bathymetry}).
@@
     \caption{Bathymetry of the Elbe model around Hamburg.
-Note the bathymetric jumps from ~5 \unit{m} on the right, upstream side to a short 10 \unit{m} step in the upper harbor area to the ~20 \unit{m} in the lower harbor area all the way to the North Sea.
+Note the bathymetric jumps from ~5 \unit{m} on the right, upstream side to a short 10 \unit{m} step in the upper port area to the ~20 \unit{m} in the lower port area all the way to the North Sea.
 Also note that there exists no channel that does not pass through the 10 \unit{km} of exclusively 20 \unit{m} deep channel.}
@@
 Secondly, because we are simulating individually particles we are able to observe their tracks.
-This makes the interpretation of our results more intuitive and allows us to include individual based properties and processes that can not or only indirectly be represented in Eulerian models.
+In our model, we use these particles to represent phytoplankton cells or aggregates.
+The temporal consistency of a Lagrangian model, the fact that we know the history of each particle, makes the interpretation of our results more intuitive and allows us to include individual-based properties and processes that cannot be represented in Eulerian models like e.g.
+retention times.
 
 %\subsubsection{Hydrodynamics data}
-We use the hydrodynamics data of the most recent SCHISM model of the Elbe \citep{Pein2021}.
-This model uses a three-dimensional unstructured  grid to represent the full Elbe estuary from the weir at Geesthacht to the North Sea, including several side-channels and the harbor area (see fig.
-\ref{fig:model_domain}).
-% (Look for more details in pein paper that might be interesting)
+We use the hydrodynamic data generated by the latest SCHISM model of the Elbe estuary \citep{Pein2021} from the weir at Geesthacht to the North Sea, including several side channels and the port area (see figure \ref{fig:model_domain}).
+SCHISM solves the Reynolds-averaged Navier-Stokes equations on unstructured meshes assuming hydrostatic conditions with a time step of 60 \unit{s}.
+The unstructured mesh is three-dimensional and consists of 32k nodes using terrain-following coordinates based on the LSC2 technique \citep{Zhang2016} for the vertical grid, allowing a maximum number of 20 levels.
+Regions with depths less than 2 \unit{m} are resolved by only one vertical level.
+The bathymetric data were provided by the German Federal Maritime and Hydrographic Agency (Bundesamt fuer Seeschifffahrt und Hydrographie, BSH) and the German Waterways Agency (Wasserstraßen- und Schiffahrtsamt, WSA) with a horizontal resolution of 50 m in the German Bight, 10 m in the Elbe estuary and 5 m in the Hamburg port \cite{Stanev2019}.
+The boundary conditions on the seaward side include sea surface elevation, horizontal currents, salinity and temperature \cite{Stanev2019} and discharge and temperature from the Elbe river on the land-ward side.
+Atmospheric forcing includes wind, air temperature, precipitation, shortwave and longwave radiation \cite{Stanev2019}.
+Model validation is based on tide gauge stations and long-term stationary measurements of salinity, water temperature, and horizontal currents.
+Biochemical variables, including chlorophyll, are based on long-term measurements at the Seemannshöft and Grauerort stations \cite{Pein2021}.
 The model provides us with a node-based mesh containing a range of information such as water velocity, salinity, water level and dispersion.
@@
 The location of the initial release is shown in red.
-Blue represents floating, green particles stranded by the receding tide.
+Blue represents floating, green phytoplankton stranded by the receding tide.
 The red area is the initial release location.
-The background map has been provided by OpenStreetMap.}
+The background map has been provided by © OpenStreetMap contributors 2023.
+Distributed under the Open Data Commons Open Database License (ODbL) v1.0.}
     \label{fig:model_domain}
@@
 
-Reproduction is represented as a chance of creating copies of themselves.
-This is a novel feature applied in OceanTracker that has not been included in any previous Lagrangian model of this kind.
-OceanTrackers recent advances in computational efficiency \citep{Vennell2021} and buffer handling make it possible to simulate a large number of particles over a long period of time for the first time on unstructured grids.
-We perform several simulations for a range of reproduction rates that are constant over the lifetime of the particle and the chance to reproduce is evaluated every minute.
+Reproduction is represented as a fission process, where each phytoplankton cell has a probability to split effectively producing a copy.
+This is a novel feature applied in OceanTracker that has not been included in any previous Lagrangian model of this type.
+OceanTracker's recent advances in computational efficiency \citep{Vennell2021} and buffer handling make it possible for the first time to simulate a large number of particles over a long period of time on unstructured grids.
+We perform multiple simulations for a range of reproduction rates, implemented as a fission probability evaluated every minute, that are constant over the lifetime of the cell.
 While a fixed reproduction rate is a simplification that does not allow for more realistic simulation of the population dynamics of a particular species, it does allow us to investigate the general mechanisms that enable plankton retention.
 
-Mortality is induced either by high salinity, when they dry-out while stranded, or due to long term light limitation.
-When particles are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is imposed.
+Mortality is induced by one of three processes: high salinity, when they dry out while stranded, or due to long term light limitation.
+When particles cells are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is applied removing dead phytoplankton cells from the simulation (see salinity map in fig.
+\ref{fig:salinity} ).
 This threshold is chosen based on a range of the salinity tolerances of estuarine phytoplankton species presented in \citep{vonAlvensleben2016}.
 This is only an approximation and salinity tolerances many estuarine phytoplankton species deviate from this.
-However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
+However, the main motivation for this choice is that most of the phytoplankton cells that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
 Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
 Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
-We consider particles that are stranded outside the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
-Particles also die if they are light-limited for $28$ days \citep{Walter2017}.
-They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
-A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by one minute, while the opposite applies when they are above $1 \unit{m}$.
-Children of light-limited parents inherit the remaining light budget of their parents.
+We consider phytoplankton cells that are stranded out of the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
+Note that these dry cells are not necessarily completely devoid of water, but are considered dry if the majority of its area has a water level is of below 0.1 \unit{m}.
+Additionally, in nature these areas typically contain small sub-resolution structures like tidal ripples or small puddles and often also vegetation.
+There are currently no studies investigating the time range for survival of stranded phytoplankton on tidal-flats or marshes in estuaries.
+Therefore, we performed a sensitivity analysis to determine the effect of this parameter on the retention success of the phytoplankton population (see section \ref{sec:dry-out}).
+Phytoplankton cells will also die if they are light-limited for $14$ days.
+This value is based on measurements presented in \citep{Walter2017} which imply that the majority of phytoplankton is dead after $14$ days of light limitation.
+A sensitivity analysis for this parameter is presented in sec. \ref{sec:light-limitation} suggesting no strong influence on the retention success.
+They are considered light-limited below a depth of $1 \unit{m}$ based on SPM data presented in \citep{Stanev2019}.
+The initial batch of phytoplankton cells starts their life with a full light budget of $14$ days, and each minute below $1 \unit{m}$ reduces this budget by one minute, while the opposite applies if they are above $1 \unit{m}$.
+When a cell splits both inherit the same remaining light budget.
 
@@
 This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which they live.
-For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
+For monodirectional vertical motion, we assign each phytoplankton cell a vertical velocity, which remains constant throughout its lifetime.
 The second mode of vertical motion is diel vertical migration.
-Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night within the same velocity range.
+Here phytoplankton cells change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night.
+This behavior is often assumed to be performed to maximize light capture while avoiding predation - or as we suspect increase retention.
 
 % A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
-We include a settling and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
-Particles become stranded when the current grid cell becomes dry. 
-They are not allowed to move from wet cells to dry cells, by the random walk dispersion applied to all particles.
+We include a settling and resuspension model to represent tidal stranding and phytoplankton cells settling on the bed of the estuary.
+Stranding phytoplankton and microphytobenthos have been shown on several occasions to be a major driver of estuarine primary production \citep{Carlson1984,DeJonge1992,Kromkamp1995,Savelli2019}.
+Phytoplankton cells become stranded when the current grid cell becomes dry and stay in place until they are resuspended or dry-out. 
+They are not allowed to move from wet cells to dry cells, by the random walk diffusion applied to all phytoplankton cells.
 A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
-Once this cell is rewetted all stranded particles resuspend and are able to move again.
-Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
+Once this cell is becomses flooded again all stranded phytoplankton cells resuspend and are able to move again.
+Phytoplankton cells settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}}.
 The velocity profile in the bottom layer, or log layer, is calculated by
@@
 scale reflecting the bottom roughness, and $u^{*}$ is the critical friction velocity.
-If the friction velocity is above the critical friction velocity the particle is resuspended.
-Particles that are stranded or settled on the bed are allowed to reproduce.
-Particles are not only advected but also dispersed based on eddy diffusivity.
-This allows us implement a dynamic dispersion that is crucial to represent tidal-pumping processes.
-Dispersion was modeled using a random walk using a random number generator with a normal distribution.
-Horizontally the standard distribution of the random walk was set to 0.1 \unit{m s^{-1}}.
-The displacement by vertical dispersion $\partial z$ of particle $i$ is calculated by
+If the friction velocity is above the critical friction velocity the phytoplankton cell is resuspended.
+Phytoplankton cells that are stranded or settled on the bed are allowed to reproduce.
+Phytoplankton cells are not only advected but also diffused based on eddy diffusivity which is crucial to represent tidal-pumping processes.
+Diffusion was modeled using a random walk using a random number generator with a normal distribution.
+Horizontally the standard distribution of the random walk is set to 0.1 \unit{m s^{-1}}.
+The vertical displacement of a phytoplankton cell $\partial z$  $i$ is calculated by
 \begin{equation}
@@
 \end{equation}
-based on \citet{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
-The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column from the hydrodynamic model output.
+based on \citet{Yamazaki2014} where $z_{i}$ is the vertical position of the phytoplankton cell, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
+The term based $K_{v}^{'}$ is needed to avoid phytoplankton accumulation on the top and bottom of the water column from the hydrodynamic model output.
 
-For each particle we log their distance traveled, age, water depth, and status (whether they are drifting or settled on the river bank or bottom).
-This allows us for example to compare successfully retained particles (older than three months) unsuccessfully retained particles (dead after less than three months).
+For each phytoplankton cell we log their distance traveled, age, water depth, and status (whether they are drifting or settled on the river bank or bottom).
+This allows us for example to compare successfully retained phytoplankton cell (older than three months) unsuccessfully retained phytoplankton cell (dead after less than three months).
 These observables are recorded every 12 hours starting at midnight.
@@
 Trajectories were calculated using a second order Runge-Kutta scheme with a fixed time step of 60 \unit{s}.
-Flow velocities, like any other hydrodynamic data, were interpolated linearly in time and space using barycentric coordinates, with the exception of water velocity in the bottom model cell, where logarithmic vertical interpolation is used.
+Flow velocities, like any other hydrodynamic data, were interpolated linearly in time, linearly in space on the vertical axis and on the horizontal axis using barycentric coordinates, with the exception of water velocity in the bottom model cell, where logarithmic vertical interpolation is used.
 
@@
 In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
-Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 40 to 404 days with a logarithmic scaling.
+Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in ranging from 40 to 404 days with a logarithmic scaling.
 In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
@@
 This results in over 1 billion individual particles simulated for each case with approximately 1 million particles active simultaneously counted over all cases over 500,000 time steps.
+This corresponds approximately to a one to one ratio of simulated phytoplankton cells to mesh nodes in the hydrodynamic model at each time step.
 The initial population is homogeneously distributed in a volume covering the full water column at the weir in Geesthacht (see fig.
 \ref{fig:model_domain}) and examine how the population distributes itself over the estuary and whether it is able to maintain its population size over time.
-Conceptually, we consider a population to be successfully retained if it is showing long term growth. 
+Conceptually, we consider a population to be successfully retained if it is able to sustain itself long term or even shows growth.
 Practically, this is evaluated by comparing the population size at the end of the year to the size after release.
+The choice of one year is considered reasonable because it covers the full seasonal cycle and is also much longer than the average exit or flushing time of the estuary (see \ref{fig:flushing-heatmap}).
 The first three months of the simulations are considered as initial model spin-up time during which the initial population is dispersed downstream throughout the estuary.
@@
 Our simulations show that the population is able to successfully retains itself under certain conditions.
-Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 3 months.
+Passively drifting phytoplankton is able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 3 months.
 Note that the growth rates realized in nature may vary from this value due to e.g.
@@
 As expected, simulations in which the reproduction is set to zero do not show any retention success.
-While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.
+While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton cells are more successful in maintaining their growth in an estuary than sinking ones.
 
@@
 
-We are now taking a closer look at spacial factors that allow phytoplankton particles to maintain net growth in the estuary. 
+We are now taking a closer look at spacial factors that allow phytoplankton cells to maintain net growth in the estuary. 
 For this analysis we used data from both sets of experiments.
 Fig.
-\ref{fig:migration-long-vs-short} compares two box plots showing the average water depth at the location of each particle between particles that remained alive for less than three months (short-living) and for more than three months (long-living).
+\ref{fig:migration-long-vs-short} compares two box plots showing the average water depth at the location of each phytoplankton cell between those cells that remained alive for less than three months (short-living) and for more than three months (long-living).
 Depth is measured relative to the current water surface.
-Hence, a value above zero indicate that the particle is stranded on the river bank.
-These analyses demonstrate that long-living particles predominantly live close to the river banks in shallower waters or on tidal flats.
+Therefore, a value greater than zero indicates that the phytoplankton cell is stranded on the shore during ebb tide.
+For reference, the water level varies on average by about 5 \unit{m} due to the tides. \citep{Stanev2019, Schol2014}.
+These analyses show that long-living phytoplankton predominantly live close to the river banks in shallower waters or on tidal flats.
 
@@
     \includegraphics[width=8.3cm]{retention_boxplot.png}
-    \caption[]{Box and violin plot showing the vertical distribution of particles that are passively drifting.
+    \caption[]{Box and violin plot showing the vertical distribution of phytoplankton that are passively drifting.
 Short-living are those younger than 3 months and long-living all those older than that.
@@
 
-We moreover analyzed the horizontal spacial distribution of long and short-living particles in fig. \ref{fig:migration-long-vs-short-heatmap}.
+We moreover analyzed the horizontal spacial distribution of long and short-living phytoplankton in fig. \ref{fig:migration-long-vs-short-heatmap}.
 To do this, we divide the model domain into equally sized hexagons.
-The color of each hexagon indicates the average age of the particles within it.
+The color of each hexagon indicates the average age of the phytoplankton cells within it calculated across all cases.
+Note, that the spatial age structure is similar for all cases.
 Hexagons with a yellow color indicate an average age of over three months.
@@
     \caption[]{
-        Hex-bin heatmap of the Elbe with Hamburg in the bottom right showing average particle age per bin.
-        Colors indicate the age of the particles, with yellowish colors indicating an average age of over three months.
+        Hex-bin heatmap of the average age of phytoplankton cells in the Elbe estuary across all cases.
+        The Hamburgs port area is located in the bottom right with the North Sea to the left.
+        Colors indicate the age of the phytoplankton, with yellowish colors indicating an average age of over three months.
         Yellow areas are mainly found along the river banks in shallow waters or tidal flats.
@@
 
-To further investigate the reasons for the positive effect of buoyancy and the importance of shallow waters and tidal flats, we repeated the first set of simulations and disabled the reproduction of settled and stranded particles.
+For comparison, we show the average exit time for each hexagon until the 20 \unit{PSU} isohaline is reached in \ref{fig:flushing-heatmap}.
+This calculation is based on a separate simulation where we released approximately 1.8 million particles homogeneously distributed over the estuary.
+We released one batch during winter in high discharge discharge conditions at the first of January and another batch during the low summer discharge at the first of July.
+We disabled reproduction, light-limitation, stranding and settling on the riverbed for this simulation to isolate the effect of advection and dispersion.
+\begin{figure}
+    \includegraphics[width=12cm]{flush_hexbin.png}
+    \caption[]{
+        Hex-bin heatmap showing average exiting times of the Elbe estuary with Hamburgs port area, as shown in \ref{fig:bathymetry}, in the bottom right without reproduction, light-limitation, stranding and settling on the riverbed.
+        Colors indicate the time of a phytoplankton cell or water parcel to reach the 20 \unit{PSU} isohaline from its origin hexagon.
+    }
+    \label{fig:flushing-heatmap}
+\end{figure}
+
+To further investigate the reasons for the positive effect of buoyancy and the importance of shallow waters and tidal flats, we repeated the first set of simulations and disabled the reproduction of settled and stranded phytoplankton.
 Under these conditions, populations were unable to retain themselves in the estuary, regardless of their vertical velocity and reproduction rate indicating that tidal flats are essential for the survival of the population.
@@
 
-Initially, we expected sinking particles to have a higher retention success than buoyant ones. 
-However, we found that faster sinking particles are less successful in retaining themselves.
+Initially, we expected sinking phytoplankton to have a higher retention success than buoyant ones. 
+However, we found that faster sinking phytoplankton are less successful in retaining themselves.
 Sinking velocities of less than $2 \unit{mm\;s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
-Sinking particles have a reduced downstream velocity because find themselves either settled on the riverbed not moving at all or close to the bed where the average downstream velocity is lower.
+Sinking phytoplankton have a reduced downstream velocity because find themselves either settled on the riverbed not moving at all or close to the bed where the average downstream velocity is lower.
 In addition, the deeper water layers of the Elbe have on average a lower downstream velocity than the upper water column or move upstream  due to temperature-induced density stratification \citep{Pein2021}.
-Nevertheless, buoyant particles were more successful in their retention in our simulations.
-The low chance of survival in the estuary for sinking particles might be explained by their light limitation in deeper waters.
-We expected particles to die if they are exposed to dark conditions for more than two weeks.
-Thus, sinking particles have a disadvantage to buoyant particles since they are more likely to become light limited and eventually die.
+Nevertheless, buoyant phytoplankton were more successful in their retention in our simulations.
+The low chance of survival in the estuary for sinking phytoplankton might be explained by their light limitation in deeper waters.
+We expected phytoplankton to die if they are exposed to dark conditions for more than two weeks.
+Thus, sinking phytoplankton have a disadvantage to buoyant phytoplankton since they are more likely to become light limited and eventually die.
 This suggests that dredging has a negative impact on sinking plankton because it increases both depth and turbidity \citep{DeJonge2014}, which increases the aphotic depth and therefore the volume of dark water relative to the amount of light water.
@@
 We suspect that the reason for the increased retention success of diel migrating organisms is similar to the monodirectional case.
-When the upwards diel migration coincides with high tide, particles are more likely to be stranded far out on the shore, reducing their risk of being washed out quickly.
+When the upwards diel migration coincides with high tide, phytoplankton are more likely to be stranded far out on the shore, reducing their risk of being washed out quickly.
 The higher the upward velocities, the greater the chance of being at the waterline during high tide.
-However, because they are sinking for half of the day they also tend to be light limited more frequent than positively buoyant particles.
+However, because they are sinking for half of the day they also tend to be light limited more frequent than positively buoyant phytoplankton.
 It seems like these favorable and unfavorable processes balance each other out, resulting in a similar retention success as for the monodirectional case.
@@
 To be consistent with the complexity of the representation of biotic mechanisms, we use a simplistic light limitation.
-Particles are expected to be light limited below a water depth of 1 \unit{m} and not light limited above this threshold. 
+Phytoplankton are expected to be light limited below a water depth of 1 \unit{m} and not light limited above this threshold. 
 We have not included a more complex light limitation model that takes into account current light availability and attenuation.
-A more realistic formulation of light limitation could particularly favor particles that exhibit diel vertical migration.
+A more realistic formulation of light limitation could particularly favor phytoplankton that exhibit diel vertical migration.
 A process we mostly ignore in our study is dormancy.
-Our organisms can survive for $28$ days in light limited waters.
+Our organisms can survive for $14$ days in light limited waters.
 However, phytoplankton species have life stages in which they can remain dormant for a longer period of time and germinate again when they find themselves in more favorable waters \citep{ThomasAnderson1998}.
@@
 
-While our model does have a settling and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury particles. 
-This masks potential losses due to particles being buried but also decreases resuspension times.
+While our model does have a settling and resuspension mechanic based on critical sheer velocities we still assume a static bathymetry with sediments not able to move or bury phytoplankton. 
+This masks potential losses due to phytoplankton being buried but also decreases resuspension times.
 
@@
 This increases the average turbidity and thus the aphotic depth, reducing the volume of water in which phytoplankton can grow.
-A large fraction of the phytoplankton measured upstream of Hamburg harbor consists of diatoms \citep{Muylaert1999}, which typically have negative buoyancy \citep{Passow1991}, making them particularly susceptible to sinking in light-limited waters.
-Our finding that sinking particles have a harder time surviving in the estuary supports this theory.
+A large fraction of the phytoplankton measured upstream of Hamburg port consists of diatoms \citep{Muylaert1999}, which typically have negative buoyancy \citep{Passow1991}, making them particularly susceptible to sinking in light-limited waters.
+Our finding that sinking phytoplankton have a harder time surviving in the estuary supports this theory.
 
@@
 \conclusions  %% \conclusions[modified heading if necessary]
-In this study, we investigated the role of different possible retention strategies for drifting phytoplankton organisms to remain in an estuarine environment.
-% As estuarine ecosystems are generally highly productive areas and of great value for anthropogenic use, maintaining their productivity is extremely important.
-However, whether and how drifting phytoplankton populations avoid being transported into unfavorable marine conditions has not been well understood.
-Our model simulations suggest that realistically high growth rates are sufficient for populations to prevent decline of estuarine populations.
+In this study, we investigated the role of different retention strategies for phytoplankton organisms to persist in an estuarine environment.
+We showed that stranding in shallow nearshore areas is essential for phytoplankton retention, and that phytoplankton that do not strand are rapidly washed away.
+Our model simulations suggest that growth rates much lower than those observed in nature may be sufficient for populations to prevent their decline due to outwashing, implying that stranding may be sufficient to maintain the population.
 Moreover, buoyancy and strong diel vertical migration enhance retention within the estuary.
-The reason for this is that phytoplankton organisms with strong buoyancy or the potential to move upward in the water column are more likely to be transported to shallow areas near the river banks or tidal flats, where currents are slower and the chances of settlement under sufficiently light conditions to allow reproduction are higher than in deeper waters.
-These results illustrate the importance of shallow nearshore areas to allowing the productivity of estuarine ecosystems to persist.
-Our results also highlight the need for informed ecosystem-based management to avoid the degradation of estuarine ecosystems by dredging and diking activities.
+These results highlight the importance of shallow nearshore areas in maintaining the productivity of estuarine ecosystems.
+Our results suggest that current state-of-the-art models of estuarine ecosystems may overlook an important process and emphasize the need for informed ecosystem-based management to avoid the degradation of estuarine ecosystems by dredging and diking activities.
 
@@
 
-% \appendix
+\appendix
 
-% \section{}    %% Appendix A
+\section{Sensitivity analysis "dry-out"}    %% Appendix A
+\label{sec:dry-out}
+
+% v26 1 day
+% v27 14 days
+
+In figure \ref{fig:retention_success_short_stranding} and \ref{fig:retention_success_long_stranding} we present the results for a sensitivity analysis of the mortality threshold due to stranding i.e.
+drying out of 1 day and 14 days compared to the 7 days presented in fig.
+\ref{fig:retention_success}
+
+\section{Sensitivity analysis "light-limitation"}
+\label{sec:light-limitation}
+
+% v28 28 days dry out
+% v29 7 days
+
+In figure \ref{fig:retention_success_low_light} and \ref{fig:retention_success_high_light} we present the results for a sensitivity analysis of the mortality threshold due to light limitation of 7 day and 28 days compared to the 14 days presented in fig.
+\ref{fig:retention_success}.
+
+\section{Salinity}
+\label{sec:salinity}
+
+In figure \ref{fig:salinity} shows a map of average salinity of the Elbe estuary. Salinity is averaged in depth and over the whole year.
 
@@
 
-% \noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).
+\noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).
 
@@
 
-% \appendixfigures  %% needs to be added in front of appendix figures
+\appendixfigures  %% needs to be added in front of appendix figures
 
-\appendixtables   %% needs to be added in front of appendix tables
+\begin{figure}
+    \includegraphics[width=8.3cm]{retention_success_sa_26.png}
+    \caption[]{
+        Sensitivity analysis for \textit{mortality due to stranding} i.e.
+drying out showing the retention success similar to fig.
+\ref{fig:retention_success} with a \textit{threshold of 1 days} without resuspension compared to the 7 days in fig \ref{fig:retention_success} before phytoplankton are culled
+    }
+    \label{fig:retention_success_short_stranding}
+\end{figure}
+\clearpage
+\begin{figure}
+    \includegraphics[width=8.3cm]{retention_success_sa_27.png}
+    \caption[]{
+        Sensitivity analysis for \textit{mortality due to stranding} i.e.
+drying out showing the retention success similar to fig.
+\ref{fig:retention_success} with a \textit{threshold of 14 days} without resuspension compared to the 7 days in fig \ref{fig:retention_success} before phytoplankton are culled
+    }
+    \label{fig:retention_success_long_stranding}
+\end{figure}
+\clearpage
+
+\appendixfigures 
+\begin{figure}
+    \includegraphics[width=8.3cm]{retention_success_sa_29.png}
+    \caption[]{
+        Sensitivity analysis for \textit{mortality due to light limitation} showing the retention success similar to fig.
+\ref{fig:retention_success} with a light deficit \textit{threshold of 7 days} compared to the 14 days in fig \ref{fig:retention_success} before phytoplankton are culled
+    }
+    \label{fig:retention_success_low_light}
+\end{figure}
+\clearpage
+\begin{figure}
+    \includegraphics[width=8.3cm]{retention_success_sa_28.png}
+    \caption[]{
+        Sensitivity analysis for \textit{mortality due to light limitation} showing the retention success similar to fig.
+\ref{fig:retention_success} with a light deficit \textit{threshold of 28 days} compared to the 14 days in fig \ref{fig:retention_success} before phytoplankton are culled
+    }
+    \label{fig:retention_success_high_light}
+\end{figure}
+\clearpage
+
+\appendixfigures 
+\begin{figure}
+    \includegraphics[width=12cm]{salinity.png}
+    \caption[]{
+        Salinity map of the Elbe estuary with Hamburgs port area, as shown in \ref{fig:bathymetry}, in the bottom right.
+        Salinity is averaged in depth and over the whole year.
+        The 20 PSU isohaline is marked with a black line.
+        Note that this plotted area has been extended downstream compared to fig. \ref{fig:migration-long-vs-short-heatmap}.
+        Further note that the color map has been capped at 25 \unit{PSU} for better visibility in low salinity areas.
+    }
+    \label{fig:salinity}
+\end{figure}
+\clearpage
+
+% \appendixtables   %% needs to be added in front of appendix tables
 
@@
 RV provided the source code for OceanTracker.
-RS,LS improved on the original physical model and LS developed the biological model.
+RS, LS improved on the original physical model and LS developed the biological model.
 LS performed the model simulations, post-processing, and visualization.
@@
 \begin{acknowledgements}
-We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data and Jana Hinners for her guidance through the project.
+We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data, Jana Hinners for her guidance through the project, and Hans Burchard for his input on dispersion.
 Further, we thank, Sina Remmers and Philipp Porada for providing helpful comments on the manuscript.
